### 2020-09-21 Manuscript/Manuscript_intro_methods_results.Rmd
Manuscript added

---
title: 'Identifying optimized decision criteria and experimental designs by simulating preclinical experiments in silico'
csl: biomed-central.csl
indent: yes
output: bookdown::pdf_document2
toc: false
linestretch: 1.5
fontsize: 12pt
header-includes:
       - \usepackage{subfig}
bibliography: refs_repro.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(bookdown)
library(knitr)
library(tidyverse)
```

```{r data, include = FALSE, cache = FALSE}
load("all_outcomes_init_samp_size_10.RData")
```

# Theoretical Background

\ \ \ \ Preclinical research stands at the beginning of the drug development process. Its purpose is to detect promising candidate drugs out of hundreds of potential interventions on the one hand, and to generate robust evidence to support translation to humans on the other hand. It has been argued that it is crucial to acknowledge the two operating modes of preclinical research in order for it to be conducted rigorously and interpreted appropriately [@kimmelman2014distinguishing]. Early-stage preclinical experiments are exploratory. Their aim is to generate hypotheses that can at a later stage be tested under more strict conditions [@landis2012call]. At early stages in drug devlopment, preclinical research also plays the role of a gatekeeper -- straining out promising interventions and denying access to further investigation to non-effective ones. It is vital, however, that decision criteria are lenient enough to allow a range of effect sizes to pass, as by that time it is unclear which effect sizes to expect. If the filters applied at this stage are too strict, chances are that potentially effective candidates are falsely eliminated.

If animal models of diseases provide the necessary criteria for proceeding drug development to a Phase I clinical trial, one would expect that they predict efficacy and safety in humans, at least to a certain extent. However, decades of failed clinical trials prove otherwise. A report of clinical development success rates from 2006 through to 2015 documents low success rates for Phase II trials, with only 30.7 % of candidate drugs progressing to Phase III [@thomas2016clinical]. A look at specific medical conditions shows an even bleaker picture. Neurological diseases like amyotrophic lateral sclerosis (ALS), Alzheimer's disease (AD), and stroke, as well as cancer have kept researchers busy for decades with few to no advances with regard to effective treatments [@dirnagl2016thomas; @moreno2013can; @mullane2019preclinical; @perrin2014preclinical; @petrov2017clinical; @van2019revised].

Preclinical research has increasingly come under scrutiny with many reports pointing out low study quality, lack of scientific rigor and high levels of irreproducibility [@freedman2014increasing]. Among the issues raised are selective reporting of results [@sena2010publication], insufficient reporting on measures to reduce bias [@vogt2016authorization], low sample sizes [@macleod2008evidence; @howells2014bringing] and concomitant increased false positive and false negative rates [@button2013power]. These shortcomings threaten the replicability of preclinical results and hamper translation of promising findings to humans [@begley2012drug; @prinz2011believe]. Additionally, effect size estimates from low power studies carry a lot of uncertainty and are prone to type M errors. That is, effect magnitudes are larger than the true unknown effect [@colquhoun2014investigation; @gelman2014beyond].

Given the importance of preclinical research on the one hand, and its shortcomings on the other hand, it is necessary to devise strategies to design preclinical experiments that yield robust evidence to guide decision making towards translation. We consider how to do this by simulating a preclinical research trajectory. We understand such a trajectory as a series of experiments, successive in time, generating evidence to support the decision to carry an intervention forward to clinical testing. While these experiments may include different types of studies, we focus on studies examining the efficacy of a given intervention. The trajectory, for the purpose of this study, comprises exploratory studies as the first stage and within-lab replications as the second stage. When moving closer towards a decision regarding translation to humans, more stages are needed (for example, between-lab replications, multicenter studies, inclusion of positive and negative controls), however, in this study, we consider the state and quality of evidence generated throughout the first two stages only. Specifically, we examine the effect of different decision criteria and design choices on transition rates from the exploratory to the replication stage, as well as success rates after within-lab replication.

For the purpose of this project, we understand replications as studies whose outcome is taken as "diagnostic evidence about a claim from prior research" [@nosek2020replication]. In the framework of preclinical research trajectories, replications are considered part of a confirmation process. Given that a series of experiments is needed to confirm a hypothesis about a directional relationship (compound X causes reaction Y), replications incrementally solidify evidence supporting (or refuting) an initial claim.

When planning a replication study, the sample size has to be determined *a priori*. This is usually done using an effect size estimate from prior research. This can be taken from the literature or a pilot study. In both cases, the effect size estimate used is most likely inflated due to reasons outlined above. Employing an overestimated effect size for sample size calculation given a desired power level (for example, 0.8) and significance threshold (for example, .05), will yield a sample size that is too low to reliably detect effects. This is especially true if effects are small. When statistical power is low, non-replications will often denote a false negative under the assumption that an underlying effect exists. An insufficiently powered replication is not suited to support or refute a previously made claim, as success and failure might be accounted for simply by inference errors.

Above and beyond the concerns mentioned, preclinical animal research raises challenging ethical concerns. Therefore, the practice oriented 3R principles [@russell1959principles] seek to reduce harm and constrain the use of animals for scientific purposes to a minimum. Current animal welfare regulations undermine the scientific value of animal research, as they fail to integrate principles that ensure high quality of conducted studies [@strech20193rs]. From an ethical point of view, this is problematic in two ways: animals are sacrificed for inconclusive research and translation to patients is impeded. Preclinical animal research needs to balance animal welfare considerations and scientific benefit. Efforts to promote and implement the 3R’s must not come at the cost of scientific value.

Ethical, time, and budget constraints prevent preclinical researchers from increasing their sample size and thus complicate the detection and precise estimation of effects. This study investigates how to design and conduct preclinical animal experiments that balance the number of animals tested against the likelihood of false negative and false positive outcomes. Our approach strives to provide a systemic perspective that does not evaluate outcomes of single experiments but evidence that is generated throughout a preclinical research trajectory. Further, we explore conditions under which replication in preclinical settings is most sensible and feasible given prior evidence and practical constraints.

<!-- Throughout the trajectory decisions have to be made based on pre-specified criteria. We simulated different scenarios with varying decision criteria and experimental design choices and evaluate outcomes as indicators of studies' conclusiveness. -->

# Methods

<!-- \ \ \ \ This simulation-based investigation explores different approaches to design and conduct experiments in preclinical settings. We explore preclinical research trajectories and determine efficiency and feasibility qua positive predictive value (PPV) and number of animals tested. These criteria ensure that while the number of animals is kept at a minimum, low sample sizes do not compromise the conclusiveness of the study. -->

## Simulation

\ \ \ \ We explored different approaches to perform preclinical animal experiments via simulations. To this end, we modeled a simplified preclinical research trajectory from the exploratory stage to the results of a within-lab replication study (Figure \@ref(fig:trajectory)). Along the trajectory, there are different ways to increase the probability of not missing potentially meaningful effects. After an initial exploratory study, a first decision identifies experiments for replication. In our simulation, we employed two different decision criteria that indicate when one should move from the exploratory stage to the replication stage. If a decision has been made to replicate an initial study, we applied two approaches to determine the sample size for a replication study (smallest effect size of interest (SESOI) and standard power analysis), as outlined in detail below.

```{r, echo = FALSE, results = FALSE}
library(png)
library(knitr)
img1_path <- "trajectory_proposal.png"
img1 <- readPNG(img1_path, native = TRUE, info = TRUE)
attr(img1, "info")
```

```{r trajectory, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "A preclinical research trajectory from the exploratory stage to a within-lab replication. The four panels along the arrow display four possible combinations of decision criteria and approaches to calculate the sample size for replication that can be employed throughout the trajectory.", fig.align = 'center'}
include_graphics(img1_path)
```

### Empirical effect size distributions

```{r ES data, include = FALSE, cache = FALSE}
load("ES_data_Szucs.RData")
load("ES_data_Carneiro.RData")
```

```{r ES distribution, echo = FALSE, results = "hide"}
min(ES_data_Szucs$D)
max(ES_data_Szucs$D)

sum(ES_data_Szucs$D > 20)

median(ES_data_Szucs$D)
mean(ES_data_Szucs$D)

names(ES_data_Carneiro)[1] <- "D"

ES_data_Carneiro$D <- ifelse(ES_data_Carneiro$D < 0, -ES_data_Carneiro$D, -ES_data_Carneiro$D)

min(ES_data_Carneiro$D)
max(ES_data_Carneiro$D)
median(ES_data_Carneiro$D)
mean(ES_data_Carneiro$D)

ES_data_Carneiro <-
  ES_data_Carneiro %>% 
  mutate(distribution = "Pessimistic")

ES_Szucs <-
  ES_data_Szucs %>% 
  select(D) %>% 
  filter(D <= 20) %>% 
  mutate(distribution = "Optimistic")

ES_data <- bind_rows(ES_Szucs, ES_data_Carneiro)
```

\ \ \ \ Simulations were based on empirical effect size distributions from the recently published literature  [@szucs2017empirical; @carneiro2018effect]. This enabled us to determine the prior probability (pre-study odds) of a certain alternative hypothesis ($H_{1}$) which we defined as an effect of a given size (e.g. a Cohen's *d* of 0.5). The two distributions reflect different research fields and cover vastly different amounts of data.

The distribution of effect sizes extracted from Szucs & Ioannidis (2017) [@szucs2017empirical] contains 26841 effect sizes from the cognitive neuroscience and psychology literature published between January 2011 and August 2014. All effect sizes are calculated as the standardized difference in means (Cohen's *d*). Effect size estimates range from `r round(min(ES_data_Szucs$D), 2)` to `r round(max(ES_data_Szucs$D), 2)`, and have a median of `r round(median(ES_data_Szucs$D), 2)` (Figure \@ref(fig:histograms)a). As the pre-study odds of a medium effect of 0.5 are rather large (`r round(outcomes_all_distributions_init_samp_size_10$prev_pop[6], 2)`), we will refer to this distribution as "optimistic". We acknowledge that the effect sizes were mainly extracted from human studies. However, in large parts the distribution is in agreement with effect sizes reported to be typical of (some areas of) preclinical research (find good ref here). Moreover, this collection of empirical effect sizes is unique in its size and we reasoned that it constitutes a good refrence (# or something similar to motivate this choice; or maybe mention this only later in discussion as limitation).

As our study is concerned specifically with preclinical research, we chose a second distribution of empirical effect sizes to represent one field of the preclinical realm. Carneiro et al.'s (2018) [@carneiro2018effect] study sytematically examined effect sizes in the rodent fear conditioning literature. Effect sizes were extracted from 410 experiments published in PubMed in 2013. The publication included a data file containing all extracted effect sizes. After removing NA's, the data set consisted of 336 effect sizes, again, calculated as Cohen's *d*. The effect sizes range from `r round(min(ES_data_Carneiro$D), 2)` to `r round(max(ES_data_Carneiro$D), 2)`, and have a median of `r round(median(ES_data_Carneiro$D), 2)` (Figure \@ref(fig:histograms)b). The prior probability of observing an effect of 0.5 is `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[2], 2)`. We will therefore refer to this distribution as "pessimistic".

Drawing from empirical effect size distributions, enables the direct comparison between pre-study odds and the positive predictive value across the trajectory. If, as outlined in the introduction, throughout the preclinical research trajectory evidence for an initial claim is strengthened, we would observe an increased positive predictive value compared to pre-study odds.

```{r histograms, echo = FALSE, fig.cap = 'Empirical effect size distributions. Note that in (a) 13 values $>$ 20 were removed in order to display the distribution.', fig.subcap = c('Optimistic distribution', 'Pessimistic distribution'), out.width = "50%"}
ES_histrogram_Szucs <-
  ggplot(data = ES_Szucs, aes(x = D)) +
  geom_histogram(binwidth = .3, color = "black", fill = "white", size = 0.3) +
  theme_bw() +
  labs(x = expression(paste("Empirical effect sizes (Cohen's ", italic("d"), ") ")), 
       y = "Frequency") + 
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black"))

ES_histrogram_Szucs

ES_histrogram_Carneiro <-
  ggplot(data = ES_data_Carneiro, aes(x = D)) +
  geom_histogram(binwidth = .3, color = "black", fill = "white", size = 0.3) +
  theme_bw() +
  labs(x = expression(paste("Empirical effect sizes (Cohen's ", italic("d"), ") ")), 
       y = "Frequency") + 
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black"))

ES_histrogram_Carneiro
```

### Exploratory stage

\ \ \ \ From each of the two distributions, we drew 10000 samples of effect sizes from which we created 10000 study data sets using R's `rnorm` function (ref or better list of all packages used for simulation with credits). Each data set comprised data of two groups consisting of ten experimental units each. An experimental unit (EU) is the entity that is randomly and independently assigned to one of the experimental groups [@lazic2018exactly]. We chose a number of ten EUs based on reported sample sizes in preclinical studies [@howells2014bringing]. Our simulated design mimics a comparison between two groups where one receives an intervention and the other functions as a control group. The study data sets are compared using a two-sided two-sample *t*-test. This procedure constituted the exploratory study in our preclinical research trajectory. From the exploratory study results, we extracted the *p*-values and 95 percent confidence intervals (CI). We then employed two different criteria based on the *p*-value or 95 percent CI, respectively, to decide whether to continue to a replication.

### Decision criteria to proceed to replication
\ \ \ \ The first decision criterion employs the conventional significance threshold  ($\alpha$ = .05) to decide whether to replicate an exploratory study. If a *p*-value extracted from a two-sided two-sample *t*-test is $\leq$ .05, this study will proceed to the replication stage. If not, the trajectory is terminated after the exploratory study. We chose this decision criterion as our reference, as this is what we consider to be current practice.

As an alternative to this approach, we propose to set a smallest effect size of interest (SESOI) and examine whether the 95 percent CI around the exploratory effect size estimate covers this SESOI. A SESOI is the effect size that the researcher based their knowledge of the literature in their respective field (domain knowledge) and given practical constraints considers biologically and clinically meaningful [@lakens2018equivalence]. In our simulation, we used 0.5 and 1.0 as SESOI. This approach emphasizes the importance of effect sizes rather than statistical significance to evaluate an intervention's effect. Further, we expected this approach to be more lenient than statistical significance (at least if the significance threshold was set at $\alpha$ = .05) and to allow a broader range of effect sizes to pass on to be further investigated.

### Approaches to determine sample size for replication

\ \ \ \ Once the decision to continue to replication has been made, we employed two different approaches to determine the sample size for the replication study. After having conducted an exploratory study, we have an estimate of the direction of the effect. Only effect sizes that showed an effect that favors the treatment over the control group were considered for further investigation. Thus, for the replication study, a one-sided two-sample *t*-test was performed. The desired power level for replication was set to .80, $\alpha$ was set to .05. In order to calculate the sample size for replication given power and $\alpha$, an effect size estimate is required. In one approach, we used the exploratory effect size estimate to compute the replication sample size. In an alternative approach, we employed the same SESOI used as decision criterion earlier. In statistical terms, our SESOI was set such that the replication study would have a power of .50 to detect an effect of this size (# explain in more detail and find good ref to motivate this: Lakens). Consequently, in the first approach, the replication sample size was dependent on the outcome of the exploratory study, whereas using a SESOI always yielded the same sample size regardless of the exploratory effect size (e.g. 23 EUs in each group for a SESOI of 0.5).

## Trajectories and outcome variables

\ \ \ \ We tested a set of four possible combinations of decision criteria and approaches to calculate the sample size for a replication study that can be employed along the trajectory (Figure \@ref(fig:trajectory)).

In the results section, we compare only the two trajectories that are the extremes of the spectrum of trajectories we have modeled: First, conventional significance threshold -- standard sample size calculation using initial
effect size estimate, second, SESOI within CI around initial effect size estimate -- standard sample size calculation using SESOI. In the following, we refer to the first trajectory as T1 and to the second as T2. We have stored data, results, and figures of all four trajectories in an online repository (insert URL here).

T1 constitutes our reference, as we consider it to be closest to current practice. T1 and T2 are compared regarding the number of experiments proceeding to the replication stage, number of animals needed in the replication, and positive predictive value across the trajectory. Secondary outcomes are the false positive rate, false negative rate, and effect size precision.

### Percentage of experiments proceeding to replication stage

\ \ \ \ Given the pre-study odds of finding an effect of a certain size (in our case 0.5 and 1.0 as SESOI), this outcome is an indication of how strict or lenient the filters are we use to select effects sizes that we deem relevant for further examination (i.e. to confirm them in a within-lab replication study). We employed two different decision criteria to determine when to continue to the replication stage.

### Number of animals needed in the replication

### Positive predictive value across trajectory

\ \ \ \ The positive predictive value (PPV) of a study is the post-study probability that a positive finding which is based on statistical significance reflects a true effect [@ioannidis2005most]. To calculate the positive predictive value, one (ideally) needs the pre-study odds (prevalence), as well as the sensitivity and specificity of the test (measurement).

### False positive rate across trajectory

\ \ \ \ The false positive rate (FPR) is the proportion of tests that detected a positive result when there was no true effect present. In our simulation, a false positive correponds to a significant finding given an underlying true effect $<$ 0.5 or 1.0, respectively.

### False negative rate across trajectory

\ \ \ \ The false negative rate (FNR) is the rpoportion of tests that missed a true positive relationship. In our simulation, this correponds to non-significant results given that the underlying true effect was $\geq$ 0.5 or 1.0, respectively.

### Effect size precision

\ \ \ \ text

## Robustness checks

\ \ \ \ As a robustness check, we ran additional simulations using seven and 15 EUs in each group of the exploratory study, repectively. We also used additional effect sizes as SESOI (0.3 and 0.7). To compare the SESOI decision criterion at the first stage to a more lenient criterion based on statistical significance, we also simulated trajectories using $\alpha$ = .1 as significance threshold. We have stored data, results, and figures of the robustness checks in an online repository (insert URL here).

# Results

## Percentage of experiments proceeding to replication stage

### Conventional significance threshold

\ \ \ \ For this decision criterion, we extracted the *p*-value from the two-sided two-sample *t*-test we conducted at the exploratory stage, and compared it to the significance level we chose as a cut off ($\alpha$ = .05). The decision to proceed to replication was solely based on the *p*-value. Effect size estimates were not considered in this step. In case of our "optimistic" scenario based on the empirical effect size distribution reported by Szucs & Ioannidis (2017) [@szucs2017empirical], `r outcomes_all_distributions_init_samp_size_10$rep_attempts[8] / 100` percent of experiments met the criterion *p* $\leq$ .05. In our "pessimistic" scenario based on the empirical effect size distribution reported by Carneiro et al. (2018)[ @carneiro2018effect], `r outcomes_all_distributions_init_samp_size_10$rep_attempts[4] / 100` percent of experiments had a *p*-value $\leq$ .05. We removed effect sizes that were smaller than zero, as we reasoned that only experiments showing effects in the direction treatment $>$ control would be further investigated. Thus the `r outcomes_all_distributions_init_samp_size_10$rep_attempts[8] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[4] / 100` percent, respectively, do include only effect sizes larger than zero. A closer look at the effect sizes of the experiments that proceeded to replication reveals that the conventional significance threshold is a conservative filter. There aren't many effect sizes around zero, but only larger effect sizes (in both directions before removal of negative effect sizes). #figure? This has consequences for the sample size calculation as well as the PPV.

### SESOI within the 95 percent CI of the exploratory effect size estimate

\ \ \ \ In another approach, after conducting a two-sided two-sample *t*-test at the exploratory stage, we estimated the effect size and 95 percent CI around that estimate. We examined whether the CI covered our SESOI (0.5 and 1.0, respectively). If this was the case for an experiment, it advanced to the replication stage. Importantly, we did not consider the *p*-value additionally. Applying this decision criterion resulted in `r outcomes_all_distributions_init_samp_size_10$rep_attempts[6] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[5] / 100` percent of experiments moving to replication in case of the "optimistic" distribution and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[2] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[1] / 100` percent for thes "pessimistic" distribution based on an SESOI of 0.5 and 1.0, respectively. Compared to the conventional significance threshold, the range of effect size estimates that proceeded to replication shifts and inlcudes less extreme values, and a lot more values closer to zero. #figure?

## Number of animals needed in the replication

\ \ \ \ In trajectory T1, sample sizes were calculated using the standard approach. We extracted the effect size estimate from the exploratory study and used it to calculate the sample size for the replication study. This resulted in a mean number of `r round(outcomes_all_distributions_init_samp_size_10$mean_N[7], 2)` (SD = `r round(outcomes_all_distributions_init_samp_size_10$mean_N[7] - outcomes_all_distributions_init_samp_size_10$mean_N_min[7], 2)`) animals in the "optimistic", and `r round(outcomes_all_distributions_init_samp_size_10$mean_N[3], 2) ` (SD = `r round(outcomes_all_distributions_init_samp_size_10$mean_N[3] - outcomes_all_distributions_init_samp_size_10$mean_N_min[3], 2)`) in the "pessimistic" scenario. 
In trajectory T2, the number of animals varied with the SESOI that was chosen. For an SESOI of 1.0, `r round(outcomes_all_distributions_init_samp_size_10$mean_N[1])` animals were needed in the replication in both the "optimistic" and "pessimistic" scenario. If the SESOI was 0.5, animal numbers increased to `r round(outcomes_all_distributions_init_samp_size_10$mean_N[2])` (Figure \@ref(fig:animalnumbers)). Note that the sample sizes reported are the number of animals needed in each group (control and intervention).

```{r, echo = FALSE}
plot_data <-
  outcomes_all_distributions_init_samp_size_10 %>% 
  filter(init_sample_size == 10) %>% 
  group_by(distribution, SESOI)
```

```{r animalnumbers, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "Number of animals needed in replication. Error bars represent standard deviations. Note that in case of trajectories using a SESOI to calculate sample size for replication the number of animals is fixed.", fig.align = 'center'}
plot_mean_N <-
  ggplot(data = plot_data,
         aes(x = factor(trajectory), y = mean_N, fill = factor(SESOI))) +
  geom_bar(stat = "identity", position = "dodge",
           size = .3,
           color = "black") +
  geom_errorbar(aes(ymax = mean_N_max, ymin = mean_N_min), width = 0.1,
                position = position_dodge(width = 0.9)) +
  facet_grid(~ distribution) +
  # ggtitle("Mean number of animals for each trajectory") +
  labs(x = "Trajectory", y = "Mean # of animals in replication",
       fill = "SESOI") +
  scale_x_discrete(labels = c("T1", "T2")) +
  scale_fill_manual(labels = c("0.5", "1.0"), values = c("grey27", "grey47")) + 
  theme_bw() +
  theme(axis.title.x = element_text(size = 14)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.text.x = element_text(size = 12, colour = "black")) +
  theme(axis.text.y = element_text(size = 12, colour = "black")) +
  theme(strip.text.x = element_text(size = 15, colour = "black", face = "bold")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.title = element_text(size = 15, face = "bold")) +
  theme(legend.text = element_text(size = 14))
  # theme(title = element_text(size = 15))

plot_mean_N
```

## Positive predictive value across trajectory

\ \ \ \ In our study, the pre-study odds were determined by the empirical effect size distributions and varied with the SESOI. In the "optimistic" scenario, the pre-study odds were `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[5], 2)` for SESOI of 0.5 and 1.0, respectively. In the "pessimistic" scenario pre-study odds were `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[1], 2)`, respectively. 
Across trajectory T1, the PPV drops below pre-study odds in both scenarios (Figure \@ref(fig:PPV)). After the within-lab replication study, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[8], 2)` and  `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[7], 2)` in the "optimitsic" scenario for SESOI of 0.5 and 1.0, respectively. In the "pessimistic" scenario, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[4], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[3], 2)`.
In trajectory T2, employing a SESOI at both stages along the decision-making process elevates the PPV above pre-study odds. Given our "optimistic" emoirical distribution, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[5], 2)` for SESOI of 0.5 and 1.0, respectively. In the "pessimistic" scenario, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[1], 2)`

```{r PPV, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "Positive predictive value across trajectory. Dashed lines indicate pre-study odds based on empirical effect size distributions.", fig.align = 'center'}
facet_names <- c(
  "0.5" = "SESOI = 0.5",
  "1" = "SESOI = 1.0")

plot_PPV <-
  ggplot(data = plot_data, 
         aes(x = factor(trajectory), y = PPV_pop_prev)) +
  geom_point(size = 3, alpha = .8) +
  facet_grid(SESOI ~ distribution, labeller = labeller(.rows = facet_names)) +
  # ggtitle("Positive predictive value for each trajectory") +
  labs(x = "Trajectory", y = "Positive predictive value") +
  scale_x_discrete(labels = c("T1", "T2")) +
  # scale_x_discrete(labels = c("Equivalence \nSESOI", "Significance \nSESOI",
  #                             "Equivalence \nStandard", "Significance \nStandard")) +
  # scale_fill_manual(breaks = c("0.5", "1"), 
  #                   values = c("steelblue", "deeppink3")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 14)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.text.x = element_text(size = 12, colour = "black")) +
  theme(axis.text.y = element_text(size = 12, colour = "black")) +
  theme(strip.text.x = element_text(size = 15, colour = "black", face = "bold")) +
  theme(strip.text.y = element_text(size = 15, colour = "black", face = "bold")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.title = element_text(size = 15, face = "bold")) +
  theme(legend.text = element_text(size = 14))
  # theme(title = element_text(size = 15))

hlines <- data.frame(pre_study_odds = c(plot_data$Prevalence[1], plot_data$Prevalence[2],
                                        plot_data$Prevalence[5], plot_data$Prevalence[6]), 
                     distribution   = c(rep(plot_data$distribution[1], 2), 
                                        rep(plot_data$distribution[5], 2)),
                     SESOI          = rep(c("1", "0.5")))

plot_PPV <- 
  plot_PPV + 
  geom_hline(data = hlines, 
             aes(yintercept = pre_study_odds),
             color = "black", lty = 2, size = .5)

plot_PPV
```

## False positive rate across trajectory

\ \ \ \ Across trajectory T1, given the "optimistic" scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[8], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[7], 2)` for SESOI of 0.5 and 1.0, respectively. In the "pessimistic" scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[4], 3)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[3], 2)` for SESOI of 0.5 and 1.0, respectively.
Across trajectory T2, the FPR increased to `r round(outcomes_all_distributions_init_samp_size_10$FPR[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[5], 2)` in the "optimistic" scenario for SESOI of 0.5 and 1.0. Given the "pessimistic" scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[1], 2)` for SESOI set to 0.5 and 1.0.

```{r FPR, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "False positive rate across trajectory.", fig.align = 'center'}
facet_names <- c(
  "0.5" = "SESOI = 0.5",
  "1" = "SESOI = 1.0")

plot_FPR <-
  ggplot(data = plot_data, 
         aes(x = factor(trajectory), y = FPR)) +
  geom_point(size = 3, alpha = .8) +
  facet_grid(SESOI ~ distribution, labeller = labeller(.rows = facet_names)) +
  # ggtitle("False positive rate for each trajectory") +
  labs(x = "Trajectory", y = "False positive rate") +
  scale_x_discrete(labels = c("T1", "T2")) +
  # scale_x_discrete(labels = c("Equivalence \nSESOI", "Significance \nSESOI",
  #                             "Equivalence \nStandard", "Significance \nStandard")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 14)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.text.x = element_text(size = 12, colour = "black")) +
  theme(axis.text.y = element_text(size = 12, colour = "black")) +
  theme(strip.text.x = element_text(size = 15, colour = "black", face = "bold")) +
  theme(strip.text.y = element_text(size = 15, colour = "black", face = "bold")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.title = element_text(size = 15, face = "bold")) +
  theme(legend.text = element_text(size = 14))
  # theme(title = element_text(size = 15))

plot_FPR
```

## False negative rate across trajectory

\ \ \ \ Across trajectory T1, given the "optimistic" scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[8], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[7], 2)` for SESOI set to 0.5 and 1.0, respectively.  In the "pessimistic" scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[4], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[3], 2)` for SESOI of 0.5 and 1.0, respectively.
Across trajectory T2, the FNR decreased to `r round(outcomes_all_distributions_init_samp_size_10$FNR[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[5], 2)` in the "optimistic" scenario for SESOI set to 0.5 and 1.0. Given the "pessimistic" scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[1], 2)` for SESOI of 0.5 and 1.0

```{r FNR, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "False negative rate across trajectory.", fig.align = 'center'}
facet_names <- c(
  "0.5" = "SESOI = 0.5",
  "1" = "SESOI = 1.0")

plot_FNR <-
  ggplot(data = plot_data, 
         aes(x = factor(trajectory), y = FNR)) +
  geom_point(size = 3, alpha = .8) +
  facet_grid(SESOI ~ distribution, labeller = labeller(.rows = facet_names)) +
  # ggtitle("False positive rate for each trajectory") +
  labs(x = "Trajectory", y = "False negative rate") +
  scale_x_discrete(labels = c("T1", "T2")) +
  # scale_x_discrete(labels = c("Equivalence \nSESOI", "Significance \nSESOI",
  #                             "Equivalence \nStandard", "Significance \nStandard")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 14)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.text.x = element_text(size = 12, colour = "black")) +
  theme(axis.text.y = element_text(size = 12, colour = "black")) +
  theme(strip.text.x = element_text(size = 15, colour = "black", face = "bold")) +
  theme(strip.text.y = element_text(size = 15, colour = "black", face = "bold")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.title = element_text(size = 15, face = "bold")) +
  theme(legend.text = element_text(size = 14))
  # theme(title = element_text(size = 15))

plot_FNR
```

## Effect size precision

\ \ \ \ text

# Discussion

\ \ \ \ text

\newpage

# References
\ \ \ \

### 2020-09-21 correct_ES_distributions_safeguard.R

@@ -1,4 +1,0 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
-
-source("load_packages.R")
-

### 2020-09-21 plots/plots_manuscript.R

@@ -1,0 +1,115 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+
+
+load("./data/all_outcomes.RData")
+
+plot_data <-
+  outcomes_all_distributions %>% 
+  filter(init_sample_size == 10) %>% 
+  group_by(distribution, SESOI)
+
+
+plot_mean_N <-
+  ggplot(data = plot_data,
+         aes(x = factor(trajectory), y = mean_N, fill = factor(SESOI))) +
+  geom_bar(stat = "identity", position = "dodge",
+           size = .3,
+           color = "black") +
+  geom_errorbar(aes(ymax = mean_N_max, ymin = mean_N_min), width = 0.1,
+                position = position_dodge(width = 0.9)) +
+  facet_grid(~ distribution) +
+  ggtitle("Mean number of animals for each trajectory") +
+  labs(x = "Trajectory", y = "Mean # of animals in replication",
+       fill = "SESOI") +
+  scale_x_discrete(labels = c("SESOI within CI \nSESOI", "Significance \nStandard")) +
+  scale_fill_manual(labels = c("0.5", "1.0"), values = c("grey17", "grey47")) + 
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 20)) +
+  theme(axis.title.y = element_text(size = 20)) +
+  theme(axis.text.x = element_text(size = 18, colour = "black")) +
+  theme(axis.text.y = element_text(size = 18, colour = "black")) +
+  theme(strip.text.x = element_text(size = 20, colour = "black", face = "bold")) +
+  theme(strip.background = element_rect(fill = "white", color = "black")) +
+  theme(legend.title = element_text(size = 20, face = "bold")) +
+  theme(legend.text = element_text(size = 18)) +
+  theme(title = element_text(size = 15))
+
+plot_mean_N
+
+# ggsave("./Manuscript/plot_mean_N.png")
+
+
+
+facet_names <- c(
+  "0.5" = "SESOI = 0.5",
+  "1" = "SESOI = 1.0")
+
+plot_PPV <-
+  ggplot(data = plot_data, 
+         aes(x = factor(trajectory), y = PPV_pop_prev)) +
+  geom_point(size = 4, alpha = .8) +
+  facet_grid(SESOI ~ distribution, labeller = labeller(.rows = facet_names)) +
+  ggtitle("Positive predictive value for each trajectory") +
+  labs(x = "Trajectory", y = "Positive predictive value") +
[…]

### 2020-09-21 scripts/analysis/analysis_effect_size_precision.R

@@ -1,0 +1,125 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation/")
+
+rm(list = ls())
+
+library(tidyverse)
+library(compute.es)
+
+
+### read in data sets from the different trajectories 
+### using fixed-N design in replication study
+
+# final <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_1.0")
+
+final_sig_Szucs <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1")
+final_sig_Szucs$distribution <- "Optimistic"
+final_sig_Szucs$trajectory <- "T1"
+
+final_equiv_Szucs <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_equiv_method2_0.5")
+final_equiv_Szucs$distribution <- "Optimistic"
+final_equiv_Szucs$trajectory <- "T2 SESOI = 0.5"
+
+final_equiv_Szucs_1 <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_equiv_method2_1.0")
+final_equiv_Szucs_1$distribution <- "Optimistic"
+final_equiv_Szucs_1$trajectory <- "T2 SESOI = 1.0"
+
+final_sig_Carneiro <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_sig_method1")
+final_sig_Carneiro$distribution <- "Pessimistic"
+final_sig_Carneiro$trajectory <- "T1"
+
+final_equiv_Carneiro <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_0.5")
+final_equiv_Carneiro$distribution <- "Pessimistic"
+final_equiv_Carneiro$trajectory <- "T2 SESOI = 0.5"
+
+final_equiv_Carneiro_1 <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_1.0")
+final_equiv_Carneiro_1$distribution <- "Pessimistic"
+final_equiv_Carneiro_1$trajectory <- "T2 SESOI = 1.0"
+
+final <- bind_rows(final_sig_Szucs,
+                   final_equiv_Szucs,
+                   final_equiv_Szucs_1,
[…]

### 2020-09-21 scripts/analysis/analysis_simulation_results_across_trajectory.R

@@ -1,11 +1,18 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation/data")
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation/")
 
 rm(list = ls())
 library(tidyverse)
 
+# source("./scripts/analysis/prior_odds_for_analysis_Szucs.R")
+
+source("./scripts/analysis/prior_odds_for_analysis_Carneiro.R")
+
 ### read in data sets from the different trajectories 
-### using fixed-N design in confirmatory study
-final <- read.csv(file = "Carneiro_distribution_sig_method1")
+### using fixed-N design in replication study
+
+final <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_1.0")
+
+# final <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1")
 
 
 hist(final$effect, breaks = 100)
@@ -14,10 +21,10 @@ hist(final$ES_true, breaks = 100)
 
 
 ### add column that codes decision criterion from exploratory stage to confirmatory stage
-final$decision_crit <- "significance"
+final$decision_crit <- "equivalence"
 
 ### add column that codes approach to determine sample size for confirmatory study
-final$sampsize_approach <- "standard"
+final$sampsize_approach <- "SESOI"
 
 ### add column for outcome significant / not significant to match outcome column
 ### of sequential design data set
@@ -39,15 +46,18 @@ dat <-
          rep_sample_size, ES_true, effect, H0,
          prev_pop, all_positives, all_negatives, rep_attempts)
 
-# write.csv(dat, file = "./Carneiro_replication_equiv_method2_0.5")
+# write.csv(dat, file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_replication_sig_method1")
 
-data_false_positives <-
-  dat %>%
-  filter(H0 == 2 & ES_true < .5)
+# write.csv(dat, file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_replication_equiv_method2_0.5")
 
-hist(data_false_positives$effect)
 
-hist(data_false_positives$ES_true)
+# data_false_positives <-
+#   dat %>%
+#   filter(H0 == 2 & ES_true < .5)
+# 
+# hist(data_false_positives$effect)
+# 
[…]

### 2020-09-21 scripts/analysis/analysis_simulation_results_across_trajectory_BF_analysis.R

@@ -1,37 +1,54 @@
 setwd("~/Documents/QUEST/PhD/R/SimulateTranslation/")
 
+rm(list = ls())
 
-BF_analysis_data <- 
-  read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/Carneiro_distribution_sig_p0.1_method1")
+
+source("./scripts/analysis/calculate_prestudy_odds_Carneiro.R")
+
+BF_analysis_data_sig <- 
+  read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/Carneiro_distribution_sig_p0.05_method1")
+
+# BF_analysis_data_equiv <- 
+#   read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/Carneiro_distribution_equiv_method1_0.5")
 
 ### add column that codes decision criterion from exploratory stage to confirmatory stage
-BF_analysis_data$decision_crit <- "significance"
+BF_analysis_data_sig$decision_crit <- "significance"
+
+# BF_analysis_data_equiv$decision_crit <- "equivalence"
 
 ### add column that codes approach to determine sample size for confirmatory study
-BF_analysis_data$sampsize_approach <- "standard"
+BF_analysis_data_sig$sampsize_approach <- "standard"
 
-BF_analysis_data$H0 <- ifelse(BF_analysis_data$BF >= 3, 2, 1)
+# BF_analysis_data_equiv$sampsize_approach <- "standard"
+
+# BF_data <- bind_rows(BF_analysis_data_sig, BF_analysis_data_equiv)
+
+BF_data <- BF_analysis_data_sig
+
+BF_data$trajectory <- interaction(BF_data$decision_crit, BF_data$sampsize_approach)
+
+
+BF_data$H0 <- ifelse(BF_data$BF >= 10, 2, 1)
 
 ### add columns for prevalence and positives and negatives
 
-BF_analysis_data$prev_pop      <- mat[1, 1]
-BF_analysis_data$all_positives <- mat[2, 1]
-BF_analysis_data$all_negatives <- mat[3, 1]
+BF_data$prev_pop      <- mat[1, 2]
+BF_data$all_positives <- mat[2, 2]
+BF_data$all_negatives <- mat[3, 2]
 
 mat
 
 ### select columns relevant for plotting
 dat <-
-  BF_analysis_data %>% 
-  select(init_sample_size, rep_no, decision_crit, sampsize_approach, 
+  BF_data %>%
[…]

### 2020-09-21 scripts/analysis/calculate_prestudy_odds_Carneiro.R

@@ -1,0 +1,85 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+
+# Here we simulate a preclinical research trajectory
+# in the first stage an exploratory study is conducted with a limited number of animals
+# in the second stage this experiment is replicated based on some information from the first experiment
+# in the third stage a multi center study is conducted
+# outcome for each experiment is 
+# 1. the measured effect size after each run (with associated CI) and p-value
+# 2. the number of animals 
+# 3. the true effect size
+
+
+# source additional functions
+source("./scripts/simulation/load_packages.R")
+source("./scripts/simulation/load_data_Carneiro.R")
+
+library(foreach)
+library(doMC)
+registerDoMC(cores = 4)
+library(doParallel)
+registerDoParallel()
+getDoParWorkers()
+
+n_exp <- 10000
+ES_true <- ES_data_Carneiro$ES_d
+hist(ES_true, breaks = 200)
+
+sum(ES_true < 0)
+sum(ES_true > 0)
+sum(ES_true == 0)
+
+ES_true <- ifelse(ES_true < 0, -ES_true, -ES_true)
+sum(ES_true > 0)
+sum(ES_true < 0)
+sum(ES_true == 0)
+
+hist(ES_true, breaks = 200)
+median(ES_true)
+mean(ES_true)
+
+set.seed(4321)
+current_ES <- sample(ES_true, n_exp, replace = TRUE)
+
+hist(current_ES, breaks = 200)
+median(current_ES)
+mean(current_ES)
+sum(current_ES > 0)
+sum(current_ES < 0)
+sum(current_ES == 0)
+
+#how many hypothesis over SESOI threshold
+SESOI <- c(.5, 1)
+
+mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
+              dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
+                              c(.5, 1)))
+
+prev_pop      <- vector()
+all_positives <- vector()
+all_negatives <- vector()
+
+counter = 0
+
+for (ES in SESOI) {
+  
+  counter = counter + 1
+  
+  prev <- round(sum(ES_true > ES)/length(ES_true), 3)
+  all_pos <- sum(current_ES > ES)
+  all_neg <- n_exp - all_pos
+  
+  print(ES)
+  
+  prev_pop[counter] <- prev
+  all_positives[counter] <- all_pos
[…]

### 2020-09-21 scripts/analysis/combine_outcomes_all_distributions.R

@@ -1,0 +1,80 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+
+
+### read in Carneiro outcomes data
+
+res_Carneiro_equiv_SESOI_1.0 <- 
+  read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_equiv_method2_1.0")
+
+res_Carneiro_equiv_SESOI_1.0$distribution <- "Pessimistic"
+res_Carneiro_equiv_SESOI_1.0$SESOI        <- 1
+
+res_Carneiro_equiv_SESOI_0.5 <- 
+  read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_equiv_method2_0.5")
+
+res_Carneiro_equiv_SESOI_0.5$distribution <- "Pessimistic"
+res_Carneiro_equiv_SESOI_0.5$SESOI        <- 0.5
+
+res_Carneiro_sig_standard_1.0 <- 
+  read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_sig_method1_1.0")
+
+res_Carneiro_sig_standard_1.0$distribution <- "Pessimistic"
+res_Carneiro_sig_standard_1.0$SESOI        <- 1
+
+res_Carneiro_sig_standard_0.5 <- 
+  read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_sig_method1_0.5")
+
+res_Carneiro_sig_standard_0.5$distribution <- "Pessimistic"
+res_Carneiro_sig_standard_0.5$SESOI        <- 0.5
+
+### read in Szucs outcomes data
+
+res_Szucs_equiv_SESOI_1.0 <- 
+  read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_outcomes_equiv_method2_1.0")
+
+res_Szucs_equiv_SESOI_1.0$distribution <- "Optimistic"
+res_Szucs_equiv_SESOI_1.0$SESOI        <- 1
+
+res_Szucs_equiv_SESOI_0.5 <- 
+  read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_outcomes_equiv_method2_0.5")
+
+res_Szucs_equiv_SESOI_0.5$distribution <- "Optimistic"
[…]

### 2020-09-21 scripts/analysis/inspection_BF_analysis_results.R

@@ -1,0 +1,49 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation/")
+
+library(tidyverse)
+
+BF_analysis_data_sig <- 
+  read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/Carneiro_distribution_sig_p0.05_method1")
+
+BF_analysis_data_equiv <- 
+  read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/Carneiro_distribution_equiv_method1_0.5")
+### add column that codes decision criterion from exploratory stage to confirmatory stage
+
+
+BF_analysis_data_sig$decision_crit <- "significance"
+
+BF_analysis_data_equiv$decision_crit <- "equivalence"
+
+### add column that codes approach to determine sample size for confirmatory study
+BF_analysis_data_sig$sampsize_approach <- "standard"
+
+BF_analysis_data_equiv$sampsize_approach <- "SESOI"
+
+BF_data <- bind_rows(BF_analysis_data_sig, BF_analysis_data_equiv)
+
+BF_data$trajectory <- interaction(BF_data$decision_crit, BF_data$sampsize_approach)
+
+BF_summary <- 
+  BF_data %>% 
+  group_by(trajectory, init_sample_size) %>% 
+  summarize(max_BF = max(BF),
+            mean_BF = mean(BF),
+            sd_BF = sd(BF),
+            median_BF = median(BF))
+
+
+BF_data <-
+  BF_data %>% 
+  filter(init_sample_size == 10) %>% 
+  filter(BF < 1e+2)
+
+hist(BF_data$BF, breaks = 200)
+
+ggplot(data = BF_data) +
+  geom_histogram(aes(x = BF), bins = 100) +
+  facet_wrap(~ trajectory)
+
+
+
+ggplot(data = BF_data) +
+  geom_boxplot(aes(x = trajectory, y = BF))

### 2020-09-21 scripts/analysis/prior_odds_for_analysis_Carneiro.R

@@ -1,0 +1,69 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+
+
+# source additional functions
+source("./scripts/simulation/load_packages.R")
+source("./scripts/simulation/load_data_Carneiro.R")
+
+n_exp <- 10000
+ES_true <- ES_data_Carneiro$ES_d
+hist(ES_true, breaks = 200)
+
+sum(ES_true < 0)
+sum(ES_true > 0)
+sum(ES_true == 0)
+
+ES_true <- ifelse(ES_true < 0, -ES_true, -ES_true)
+sum(ES_true > 0)
+sum(ES_true < 0)
+sum(ES_true == 0)
+
+hist(ES_true, breaks = 200)
+median(ES_true)
+mean(ES_true)
+
+set.seed(4321)
+current_ES <- sample(ES_true, n_exp, replace = TRUE)
+
+hist(current_ES, breaks = 200)
+median(current_ES)
+mean(current_ES)
+sum(current_ES > 0)
+sum(current_ES < 0)
+sum(current_ES == 0)
+
+#how many hypothesis over SESOI threshold
+SESOI       <- c(.5, 1)
+
+mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
+              dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
+                              c(.5, 1)))
+
+prev_pop      <- vector()
+all_positives <- vector()
+all_negatives <- vector()
+
+counter = 0
+
+for (ES in SESOI) {
+  
+  counter = counter + 1
+  
+  prev <- round(sum(ES_true > ES)/length(ES_true), 3)
+  all_pos <- sum(current_ES > ES)
+  all_neg <- n_exp - all_pos
+  
+  print(ES)
+  
+  prev_pop[counter] <- prev
+  all_positives[counter] <- all_pos
+  all_negatives[counter] <- all_neg
+  
+}
+
+
+mat[1, ] <- prev_pop
+mat[2, ] <- all_positives
+mat[3, ] <- all_negatives
+
+mat

### 2020-09-21 scripts/analysis/prior_odds_for_analysis_Szucs.R

@@ -1,0 +1,50 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+
+
+# source additional functions
+source("./scripts/simulation/load_packages.R")
+source("./scripts/simulation/load_data_Szucs.R")
+
+n_exp <- 10000
+ES_true <- ES_data_Szucs$D
+
+set.seed(4321)
+current_ES <- sample(ES_true, n_exp)
+hist(ES_true, breaks = 200)
+hist(current_ES, breaks = 200)
+
+#how many hypothesis over SESOI threshold
+SESOI       <- c(.5, 1)
+
+mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
+              dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
+                              c(.5, 1)))
+
+prev_pop      <- vector()
+all_positives <- vector()
+all_negatives <- vector()
+
+counter = 0
+
+for (ES in SESOI) {
+  
+  counter = counter + 1
+  
+  prev <- round(sum(ES_true > ES)/length(ES_true), 3)
+  all_pos <- sum(current_ES > ES)
+  all_neg <- n_exp - all_pos
+  
+  print(ES)
+  
+  prev_pop[counter] <- prev
+  all_positives[counter] <- all_pos
+  all_negatives[counter] <- all_neg
+  
+}
+
+
+mat[1, ] <- prev_pop
+mat[2, ] <- all_positives
+mat[3, ] <- all_negatives
+
+mat

### 2020-09-21 scripts/plotting/plot_outcomes_Carneiro_distribution.R

@@ -8,52 +8,46 @@ library(tidyverse)
 
 #source("./scripts/calculate_outcomes_onesided_across_trajectory.R")
 
-out1 <- read.csv(file = "./data/Carneiro_outcomes_equiv_method1_1.0")
-out2 <- read.csv(file = "./data/Carneiro_outcomes_equiv_method2_1.0")
-out3 <- read.csv(file = "./data/Carneiro_outcomes_sig_method1_1.0")
-out4 <- read.csv(file = "./data/Carneiro_outcomes_sig_method2_1.0")
+out1 <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_equiv_method1_1.0")
+out2 <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_equiv_method2_1.0")
+out3 <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_sig_method1_1.0")
+out4 <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_sig_method2_1.0")
 
 
 outcomes1 <- bind_rows(out1, out2, out3, out4)
 outcomes1$SESOI <- 1.0
 
-out5 <- read.csv(file = "./data/Carneiro_outcomes_equiv_method1_0.7")
-out6 <- read.csv(file = "./data/Carneiro_outcomes_equiv_method2_0.7")
-out7 <- read.csv(file = "./data/Carneiro_outcomes_sig_method1_0.7")
-out8 <- read.csv(file = "./data/Carneiro_outcomes_sig_method2_0.7")
+out5 <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_equiv_method1_0.7")
+out6 <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_equiv_method2_0.7")
+out7 <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_sig_method1_0.7")
+out8 <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_outcomes_sig_method2_0.7")
 
 
 outcomes2 <- bind_rows(out5, out6, out7, out8)
 outcomes2$SESOI <- 0.7
 
[…]

### 2020-09-21 scripts/plotting/plot_outcomes_Carneiro_distribution_BF.R

@@ -1,0 +1,307 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+
+rm(list = ls())
+
+library(svglite)
+library(tidyverse)
+
+# setwd("~/Documents/QUEST/PhD/R/SimulateTranslation/data/Carneiro_distribution/Bayes_factor_analysis")
+# 
+# list_filenames <- list.files(pattern ="*_1.0_BF10")
+# list_filenames
+# 
+# list_data <- list()
+# 
+# for (i in 1:length(list_filenames)) {
+#   list_data[[i]]<-read.csv(list_filenames[i])
+# }
+# 
+# names(list_data) <- list_filenames
+# 
+# length(list_data)
+
+#source("./scripts/calculate_outcomes_onesided_across_trajectory.R")
+
+out1 <- read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/outcomes_equiv_method1_1.0_BF10")
+out2 <- read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/outcomes_equiv_method2_1.0_BF10")
+out3 <- read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/outcomes_sig_p0.05_method1_1.0_BF10")
+out4 <- read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/outcomes_sig_p0.05_method2_1.0_BF10")
+out5 <- read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/outcomes_sig_p0.1_method1_1.0_BF10")
+out6 <- read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/outcomes_sig_p0.1_method2_1.0_BF10")
+
+levels(out5$decision_crit) <- c(levels(out5$decision_crit), "significance_0.1")
+out5$decision_crit[out5$decision_crit == 'significance'] <- 'significance_0.1'
+
+levels(out6$decision_crit) <- c(levels(out6$decision_crit), "significance_0.1")
+out6$decision_crit[out6$decision_crit == 'significance'] <- 'significance_0.1'
+
+outcomes1 <- bind_rows(out1, out2, out3, out4, out5, out6)
+outcomes1$SESOI <- 1.0
+
+out7 <- read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/outcomes_equiv_method1_0.5_BF10")
[…]

### 2020-09-21 scripts/plotting/plot_outcomes_Carneiro_distribution_all_BF.R

@@ -1,0 +1,129 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+
+
+outcomes_BF3  <- read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/outcomes_all_SESOI_BF3")
+outcomes_BF10 <- read.csv(file = "./data/Carneiro_distribution/Bayes_factor_analysis/outcomes_all_SESOI_BF10")
+
+
+outcomes_BF3$BF_crit <- 3
+outcomes_BF10$BF_crit <- 10
+
+outcomes <- bind_rows(outcomes_BF3, outcomes_BF10)
+
+outcomes <-
+  outcomes %>% 
+  filter(init_sample_size == 10)
+
+outcomes$trajectory <- interaction(outcomes$decision_crit, outcomes$sampsize_approach)
+
+outcomes <-
+  outcomes %>% 
+  filter(trajectory != "significance.SESOI") %>% 
+  filter(trajectory != "equivalence.standard") %>% 
+  filter(trajectory != "significance_0.1.SESOI")
+
+plot_data <-
+  outcomes %>% 
+  select(SESOI, init_sample_size, trajectory, 
+         Prevalence, PPV_pop_prev, mean_N, FPR, BF_crit) 
+
+plot_mean_N <-
+  ggplot(data = plot_data,
+         aes(x = factor(trajectory), y = mean_N, fill = factor(SESOI))) +
+  geom_bar(stat = "identity", position = "dodge", 
+           alpha = 0.7, color = "black", size = .3) +
+  facet_wrap(~ BF_crit) +
+  # ggtitle("Mean number of animals for each trajectory") +
+  labs(x = "Trajectory", y = "Mean # of animals", 
+       fill = "SESOI") +
+  scale_x_discrete(labels = c("Equivalence \nSESOI", "Significance \nStandard", "Significance at 0.1 \nStandard")) +
+  scale_fill_manual(breaks = c("0.5", "1"), 
+                    values = c("steelblue", "deeppink3")) +
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 20)) +
+  theme(axis.title.y = element_text(size = 20)) +
+  theme(axis.text.x = element_text(size = 18, colour = "black")) +
+  theme(axis.text.y = element_text(size = 18, colour = "black")) +
+  # theme(strip.text.x = element_text(size = 20, colour = "black", face = "bold")) +
+  theme(strip.background = element_rect(fill = "white", color = "black")) +
[…]

### 2020-09-21 scripts/simulation/load_data_Carneiro.R

@@ -11,6 +11,8 @@ ES_data_Carneiro <-
   select(ES_d) %>% 
   drop_na()
 
+# save(ES_data_Carneiro, file = "ES_data_Carneiro.RData")
+
 # zeros <- which(ES_data_Carneiro$ES_d == 0)
 # sum(ES_data_Carneiro$ES_d == 0)
 # mean(ES_data_Carneiro$ES_d)

### 2020-09-21 scripts/simulation/load_data_Szucs.R

@@ -19,7 +19,7 @@ ES_data <- as.data.frame(cbind(t_value, df))
 
 ES_data <-
   ES_data %>% 
-  mutate(Dt1= t_value / (sqrt(df + 1)),
+  mutate(Dt1 = t_value / (sqrt(df + 1)),
          Dt2 = 2 * t_value / (sqrt(df + 2)))
 
 ES_data <-
@@ -45,20 +45,28 @@ names(ES_large_sample)[3] <- "D"
 ES_data_Szucs <-
   bind_rows(ES_small_sample, ES_large_sample)
 
+# save(ES_data_Szucs, file = "ES_data_Szucs.RData")
+
+min(ES_data_Szucs$D)
 max(ES_data_Szucs$D)
 
-sum(ES_data_Szucs$D > 16)
+sum(ES_data_Szucs$D > 20)
 
 median(ES_data_Szucs$D)
 mean(ES_data_Szucs$D)
+
+ES_small <-
+  ES_data_Szucs %>%
+  filter(D <= 20)
+
 
 
 # ggplot(data = ES_data_Szucs, aes(x = df, y = D)) +
 #   geom_jitter(alpha = .5) +
 #   theme_bw()
 # 
-# ggplot(data = ES_data_Szucs, aes(x = D)) +
-#   geom_histogram(binwidth = .09, color = "black", fill = "white") +
+# ggplot(data = ES_small, aes(x = D)) +
+#   geom_histogram(binwidth = .3, color = "black", fill = "white", size = 0.3) +
 #   theme_bw()
 # 
 # ggplot(data = ES_data_Szucs, aes(y = D)) +

### 2020-09-21 scripts/simulation/sim_fixN_Carneiro_decision_equivalence_with_SESOI.R

@@ -134,7 +134,7 @@ selection_equiv <- list()
 for (i in 1:length(samp_size_vector)) {
   
   selection_equiv[[i]] <- future_map(exploratory_data_summary[[i]], get_decision_equiv,
-                                     SESOI = 0.5)
+                                     SESOI = 1.0)
   
 }
 

### 2020-09-21 scripts/simulation/sim_fixN_Carneiro_replication.R

@@ -5,9 +5,9 @@ rm(list = ls())
 
 # source additional scripts
 
-# source("./scripts/sim_fixN_Szucs_decision_equivalence_with_SESOI.R")
+source("./scripts/simulation/sim_fixN_Carneiro_decision_equivalence_with_SESOI.R")
 
-source("./scripts/simulation/sim_fixN_Carneiro_decision_significance.R")
+# source("./scripts/simulation/sim_fixN_Carneiro_decision_significance.R")
 
 dat
 
@@ -18,7 +18,7 @@ sum(dat$ES_true < 0)
 data <-
   dat %>% 
   group_by(init_sample_size, study_id) %>% 
-  filter(selection_sig == 1)
+  filter(selection_equiv == 1)
 
 sum(data$effect < 0)
 sum(data$effect > 0)
@@ -27,7 +27,7 @@ sum(data$effect == 0)
 selected <-
   data %>% 
   group_by(init_sample_size) %>% 
-  summarize(selected = sum(selection_sig == 1))
+  summarize(selected = sum(selection_equiv == 1))
 
 # now estimate sample size for replication study
 # A. Initial effect size (probably the worst solution but good as a floor benchmark, 1 in function)
@@ -39,7 +39,7 @@ for (i in 1:nrow(data)) {
 
   rep_sample_size_std[i] <-
     ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-                             method = 2, SESOI = .7))
+                             method = 2, SESOI = 1.0, power = 0.5))
 }
 
 data$rep_samp_size_std <- rep_sample_size_std
@@ -55,6 +55,7 @@ hist(data$ES_true, breaks = 200)
 
 sum(data$effect > 0)
 sum(data$effect < 0)
+sum(data$effect == 0)
 
 negative_ES <-
   data %>% 
@@ -79,15 +80,13 @@ replication_data <- list()
 
 rep_exp_no <- 0
 
-select_experiments <- which(data$selection_sig == 1)
+# select_experiments <- which(data$selection_sig == 1)
 
-# select_experiments <- which(data$selection_equiv == 1)
+select_experiments <- which(data$selection_equiv == 1)
 
 select_experiments <- select_experiments[data$effect[select_experiments] >= 0 ]
 
 current_ES_rep <- data$ES_true
-
-#rep_sample_size_std <- rep(25, length(select_experiments))
 
 for(i in select_experiments) {
   
[…]

### 2020-09-21 scripts/simulation/sim_fixN_Szucs_decision_equivalence_with_SESOI.R

@@ -34,11 +34,11 @@ hist(ES_true, breaks = 200)
 hist(current_ES, breaks = 200)
 
 #how many hypothesis over SESOI threshold
-SESOI       <- c(.3, .5, .7, 1)
+SESOI       <- c(.5, 1)
 
 mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
               dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
-                              c(.3, .5, .7, 1)))
+                              c(.5, 1)))
 
 prev_pop      <- vector()
 all_positives <- vector()

### 2020-09-21 scripts/simulation/sim_fixN_Szucs_decision_significance.R

@@ -25,7 +25,7 @@ library(doParallel)
 registerDoParallel()
 getDoParWorkers()
 
-n_exp <- 10000
+n_exp <- 10
 ES_true <- ES_data_Szucs$D
 
 set.seed(4321)
@@ -34,11 +34,11 @@ hist(ES_true, breaks = 200)
 hist(current_ES, breaks = 200)
 
 #how many hypothesis over SESOI threshold
-SESOI       <- c(.3, .5, .7, 1)
+SESOI       <- c(.5, 1)
 
 mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
               dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
-                              c(.3, .5, .7, 1)))
+                              c(.5, 1)))
 
 prev_pop      <- vector()
 all_positives <- vector()
@@ -111,7 +111,7 @@ selection_sig <- list()
 for (i in 1:length(samp_size_vector)) {
   
   selection_sig[[i]] <- 
-    future_map(exploratory_data_summary[[i]], get_decision_sig)
+    future_map(exploratory_data_summary[[i]], get_decision_sig, pval_threshold = 0.05)
   
 }
 
@@ -138,6 +138,6 @@ dat <- bind_cols(df, df_sig)
 
 dat$ES_true <- current_ES
 
-write.csv(dat, file = "./data/Szucs_distribution_only_exploratory_stage")
+# write.csv(dat, file = "./data/Szucs_distribution_only_exploratory_stage")
 
 

### 2020-09-21 scripts/simulation/sim_fixN_Szucs_replication.R

@@ -5,42 +5,71 @@ rm(list = ls())
 
 # source additional scripts
 
-# source("./scripts/sim_fixN_Szucs_decision_equivalence_with_SESOI.R")
+source("./scripts/simulation/sim_fixN_Szucs_decision_equivalence_with_SESOI.R")
 
-source("./scripts/simulation/sim_fixN_Szucs_decision_significance.R")
+# source("./scripts/simulation/sim_fixN_Szucs_decision_significance.R")
 
 dat
+
+sum(dat$effect < 0) # empirical effect sizes are negative because t.test function takes control - treat
+sum(dat$effect > 0) 
+sum(dat$ES_true < 0)
+
+# data <-
+#   dat %>% 
+#   group_by(init_sample_size, study_id) %>% 
+#   filter(selection_sig == 1)
 
 data <-
   dat %>% 
   group_by(init_sample_size, study_id) %>% 
-  filter(selection_sig == 1)
+  filter(selection_equiv == 1)
+
+sum(data$effect < 0)
+sum(data$effect > 0)
+sum(data$effect == 0)
+
+
+# selected <-
+#   data %>% 
+#   group_by(init_sample_size) %>% 
+#   summarize(selected = sum(selection_sig == 1))
 
 selected <-
   data %>% 
   group_by(init_sample_size) %>% 
-  summarize(selected = sum(selection_sig == 1))
+  summarize(selected = sum(selection_equiv == 1))
 
 
 # now estimate sample size for replication study
 # A. Initial effect size (probably the worst solution but good as a floor benchmark, 1 in function)
 # B. Set a SESOI. With this all experiments will have the same number of EU (2 as function parameter)
 
-# rep_sample_size_std <- NULL
-# 
-# for (i in 1:nrow(data)) {
-# 
-#   rep_sample_size_std[i] <-
-#     ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-#                              method = 1))
-# }
-# 
-# 
-# data$rep_samp_size_std <- rep_sample_size_std
-# 
-# max(data$rep_samp_size_std)
+rep_sample_size_std <- NULL
+
+for (i in 1:nrow(data)) {
+  
+  rep_sample_size_std[i] <-
+    ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
+                             method = 2, SESOI = 1.0, power = 0.5))
+}
+
+data$rep_samp_size_std <- rep_sample_size_std
+
[…]

### 2020-09-21 simulation_two_stages.R

@@ -1,71 +1,0 @@
-# Here we simulate a preclinical study
-# in the first stage an exploratory study is cinducted with a limited number of animals
-# in the second stage this experiment is replicated based on some information from the first experiment
-# in the third stage a multi center study is conducted
-# outcome for each experiment is 
-# 1. the measured effect size after each run (with associated CI) and p-value
-# 2. the number of animals 
-# 3. the true effect size
-#source additional functions
-source("functions_for_sim.R")
-
-# Step  1 Generate an effect size distribution
-ES_true<-c(rbeta(100000,6,5),rbeta(100000,1,5))
-ES_true<-c(rbeta(100000,1,5))
-hist(ES_true)
-#how many hypothesis over .3 threshold
-sum(ES_true>0.3)
-n_exp<-1000
-current_ES<-sample(ES_true,n_exp)
-hist(current_ES)
-#conduct intial study
-exploratory_data<-list()
-for(i in 1:n_exp)
-  exploratory_data[[i]]<-generate_study(current_ES[i])
-#the confidence interval generated here is used in the safeguard power analysis
-exp_data_summary<-list()
-for(i in 1:n_exp)
-  exp_data_summary[[i]]<-get_summary_study(study_data = exploratory_data[[i]], confidence = .8)
-#now estimate sample size
-#A. Safeguard (1 in function)
-#B. Initial effect size (probably the worst solution but good as a floor benchmark, 2 in function)
-#C. Set a minimum detectable effect size. With this all experiments will have the same number of EU (3 as function parameter)
-rep_sample_size<-NULL
-for(i in 1:n_exp)
-  rep_sample_size[i]<-ceiling(calc_sample_size(study_summary=exp_data_summary[[i]],study_data=exploratory_data[[i]],method=3))
-#Decision to go on
-#this decision depends on an equivalence test with a bound of .3
-aa<-(unlist(map(exp_data_summary,"CI")))
-select_experiments<-which(((apply(cbind(aa[seq(1,length(aa)-1,2)],aa[seq(2,length(aa),2)]),1,function(x){min((x))})< -.3)))
-#sum(current_ES[bb]>.3)/length(bb)
-length(select_experiments)
-sum(current_ES[-select_experiments]>.3)/length(current_ES[-select_experiments])
-hist(current_ES[-select_experiments])
-hist(rep_sample_size[select_experiments])
[…]

### 2020-09-23 Manuscript/Manuscript_intro_methods_results.Rmd
manuscript updated

@@
 Employing an overestimated effect size for sample size calculation given a desired power level (for example, 0.8) and significance threshold (for example, .05), will yield a sample size that is too low to reliably detect effects.
-This is especially true if effects are small.
+This is especially true if true underlying effects are small.
 When statistical power is low, non-replications will often denote a false negative under the assumption that an underlying effect exists.
@@
 # Methods
-
-<!-- \ \ \ \ This simulation-based investigation explores different approaches to design and conduct experiments in preclinical settings.
-We explore preclinical research trajectories and determine efficiency and feasibility qua positive predictive value (PPV) and number of animals tested.
-These criteria ensure that while the number of animals is kept at a minimum, low sample sizes do not compromise the conclusiveness of the study.
--->
 
@@
 
-Drawing from empirical effect size distributions, enables the direct comparison between pre-study odds and the positive predictive value across the trajectory.
-If, as outlined in the introduction, throughout the preclinical research trajectory evidence for an initial claim is strengthened, we would observe an increased positive predictive value compared to pre-study odds.
-
 ```{r histograms, echo = FALSE, fig.cap = 'Empirical effect size distributions.
@@
 
-## Trajectories and outcome variables
+### Replication stage
+
+\ \ \ \ For each of the studies that met the decision criterion after the exploratory study (either *p* $\leq$ .05 or SESOI within the 95 percent CI of the exploratory effect size estimate) a replication study was performed.
+The number of replication studies conducted varied with the decision criterion used and, in case of the criterion employing a SESOI, also with the SESOI (0.5 and 1.0).
+A replication study was performed as a one-sided two-sample *t*-test, where the number of animals in each group was determined by the approach to calculate the sample size.
+For a replication be considered "successful", the *p*-value had to be below the conventional significance threshold ($\alpha$ = .05)
+
+## Trajectories
 
@@
 
-In the results section, we compare only the two trajectories that are the extremes of the spectrum of trajectories we have modeled: First, conventional significance threshold -- standard sample size calculation using initial
-effect size estimate, second, SESOI within CI around initial effect size estimate -- standard sample size calculation using SESOI.
+In the results section, we compare only the two trajectories that are the extremes of the spectrum of trajectories we have modeled: The first trajectory uses the conventional significance threshold as decision criterion and employs a standard sample size calculation using initial effect size estimate.
+The second trajectory applies a SESOI as decision criterion to move from the exploratory to the replication stage, and to calculate the sample size for the replication.
 In the following, we refer to the first trajectory as T1 and to the second as T2.
@@
 Secondary outcomes are the false positive rate, false negative rate, and effect size precision.
+Outcome variables are outlined in more detail in the following section.
+
+## Outcome variables
 
@@
 
-\ \ \ \ Given the pre-study odds of finding an effect of a certain size (in our case 0.5 and 1.0 as SESOI), this outcome is an indication of how strict or lenient the filters are we use to select effects sizes that we deem relevant for further examination (i.e.
+\ \ \ \ This outcome serves as an indication of how strict or lenient the filters are that we used to select effect sizes that we deem relevant for further examination (i.e.
 to confirm them in a within-lab replication study).
-We employed two different decision criteria to determine when to continue to the replication stage.
 
@@
 To calculate the positive predictive value, one (ideally) needs the pre-study odds (prevalence), as well as the sensitivity and specificity of the test (measurement).
+Given that we drew effect sizes from empirical effect size distributions, we know the pre-study odds of an effect of 0.5 and 1.0.
+Drawing from empirical effect size distributions, enables the direct comparison between pre-study odds and the positive predictive value across the trajectory.
+If, as outlined in the introduction, throughout the preclinical research trajectory evidence for an initial claim is strengthened, we would observe an increased positive predictive value compared to pre-study odds.
 
@@
 \ \ \ \ The false positive rate (FPR) is the proportion of tests that detected a positive result when there was no true effect present.
-In our simulation, a false positive correponds to a significant finding given an underlying true effect $<$ 0.5 or 1.0, respectively.
+In our simulation, a false positive correponds to a significant finding given an underlying true effect smaller than 0.5 or 1.0, respectively.
 
@@
 
-\ \ \ \ The false negative rate (FNR) is the rpoportion of tests that missed a true positive relationship.
-In our simulation, this correponds to non-significant results given that the underlying true effect was $\geq$ 0.5 or 1.0, respectively.
+\ \ \ \ The false negative rate (FNR) is the proportion of tests that missed a true positive relationship.
+In our simulation, this correponds to non-significant results given that the underlying true effect was greater or equal to 0.5 or 1.0, respectively.
 

### 2020-09-23 scripts/simulation/sim_fixN_Szucs_decision_significance.R

@@ -25,7 +25,7 @@ library(doParallel)
 registerDoParallel()
 getDoParWorkers()
 
-n_exp <- 10
+n_exp <- 3
 ES_true <- ES_data_Szucs$D
 
 set.seed(4321)

### 2020-09-23 scripts/simulation/sim_fixN_Szucs_replication.R

@@ -15,30 +15,30 @@ sum(dat$effect < 0) # empirical effect sizes are negative because t.test functio
 sum(dat$effect > 0) 
 sum(dat$ES_true < 0)
 
+data <-
+  dat %>%
+  group_by(init_sample_size, study_id) %>%
+  filter(selection_sig == 1)
+
 # data <-
 #   dat %>% 
 #   group_by(init_sample_size, study_id) %>% 
-#   filter(selection_sig == 1)
-
-data <-
-  dat %>% 
-  group_by(init_sample_size, study_id) %>% 
-  filter(selection_equiv == 1)
+#   filter(selection_equiv == 1)
 
 sum(data$effect < 0)
 sum(data$effect > 0)
 sum(data$effect == 0)
 
 
+selected <-
+  data %>%
+  group_by(init_sample_size) %>%
+  summarize(selected = sum(selection_sig == 1))
+
 # selected <-
 #   data %>% 
 #   group_by(init_sample_size) %>% 
-#   summarize(selected = sum(selection_sig == 1))
-
-selected <-
-  data %>% 
-  group_by(init_sample_size) %>% 
-  summarize(selected = sum(selection_equiv == 1))
+#   summarize(selected = sum(selection_equiv == 1))
 
 
 # now estimate sample size for replication study
@@ -51,7 +51,11 @@ for (i in 1:nrow(data)) {
   
   rep_sample_size_std[i] <-
     ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-                             method = 2, SESOI = 1.0, power = 0.5))
+                             method = 1))
+  
+  # rep_sample_size_std[i] <-
+  #   ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
+  #                            method = 2, SESOI = 0.5, power = .5))
 }
 
 data$rep_samp_size_std <- rep_sample_size_std
@@ -94,9 +98,9 @@ rep_attempts <-
 replication_data <- list()
 rep_exp_no <- 0
 
-# select_experiments <- which(data$selection_sig == 1)
+select_experiments <- which(data$selection_sig == 1)
 
-select_experiments <- which(data$selection_equiv == 1)
+# select_experiments <- which(data$selection_equiv == 1)
 
 select_experiments <- select_experiments[data$effect[select_experiments] >= 0 ]
 

### 2020-09-23 scripts/troubleshooting.R

@@ -20,7 +20,9 @@ round(power.t.test(n = 61, delta = NULL	, sd = 1, sig.level = .05, power = .8,
 
 abs(aa$mean_effect[2] - aa$mean_effect[1]) > 16
 
-
+ceiling(power.t.test(n = NULL, delta = 1.4190	, sd = 1, sig.level = .05, power = .8,
+                     type = "two.sample",
+                     alternative = "one.sided")$n)
 
 
 ES_true=1

### 2020-10-19 Manuscript/Manuscript_intro_methods_results.Rmd

@@
 
-\ \ \ \ Preclinical research stands at the beginning of the drug development process.
-Its purpose is to detect promising candidate drugs out of hundreds of potential interventions on the one hand, and to generate robust evidence to support translation to humans on the other hand.
-It has been argued that it is crucial to acknowledge the two operating modes of preclinical research in order for it to be conducted rigorously and interpreted appropriately [@kimmelman2014distinguishing].
-Early-stage preclinical experiments are exploratory.
-Their aim is to generate hypotheses that can at a later stage be tested under more strict conditions [@landis2012call].
-At early stages in drug devlopment, preclinical research also plays the role of a gatekeeper -- straining out promising interventions and denying access to further investigation to non-effective ones.
-It is vital, however, that decision criteria are lenient enough to allow a range of effect sizes to pass, as by that time it is unclear which effect sizes to expect.
+\ \ \ \ Preclinical research is essential for identifying promising interventions and to generate robust evidence to support translation to humans.
+The experiments in a preclinical setting are conducted  in  two different operating modes [@kimmelman2014distinguishing].
+Early-stage preclinical experiments are *exploratory* with the aim is to generate hypotheses.
+These are tested at a later stage under more strict conditions in *confirmatory* mode [@landis2012call].
+At early stages of treatment discovery, preclinical research also plays the role of a gatekeeper -- straining out promising interventions and stopping further investigation for non-effective ones.
+It is vital, however, that decision criteria are lenient enough to allow also uncertain effects **[this is too early or abrupt to already introduce effect sizes]**to pass.
 If the filters applied at this stage are too strict, chances are that potentially effective candidates are falsely eliminated.
 
-If animal models of diseases provide the necessary criteria for proceeding drug development to a Phase I clinical trial, one would expect that they predict efficacy and safety in humans, at least to a certain extent.
-However, decades of failed clinical trials prove otherwise.
+If animal models of disease provide the necessary criteria for proceeding drug development to engage in clinical trials, one would expect that they predict efficacy and safety in humans, at least to a certain extent [It is not entirely clear what you mean with previous sentence particularly criteria.
+Shouldn't this be something like: enable an informed decision to engage in ct].
+However, decades of failed clinical trials suggest otherwise.
 A report of clinical development success rates from 2006 through to 2015 documents low success rates for Phase II trials, with only 30.7 % of candidate drugs progressing to Phase III [@thomas2016clinical].
-A look at specific medical conditions shows an even bleaker picture.
-Neurological diseases like amyotrophic lateral sclerosis (ALS), Alzheimer's disease (AD), and stroke, as well as cancer have kept researchers busy for decades with few to no advances with regard to effective treatments [@dirnagl2016thomas; @moreno2013can; @mullane2019preclinical; @perrin2014preclinical; @petrov2017clinical; @van2019revised].
+For specific medical conditions like amyotrophic lateral sclerosis (ALS), Alzheimer's disease (AD), and stroke, as well as cancer [I would keep cancer out of this for now]  decades of research have yielded surprisingly few advances with regard to effective treatments [@dirnagl2016thomas; @moreno2013can; @mullane2019preclinical; @perrin2014preclinical; @petrov2017clinical; @van2019revised].
+[I toned this down as it is not your main focus to criticise everything up until now, but motivate the need for a change]
 
@@
 That is, effect magnitudes are larger than the true unknown effect [@colquhoun2014investigation; @gelman2014beyond].
+[I dont know yet whether we need the other shortcomings.
+Sample size may be enough.]
 
-Given the importance of preclinical research on the one hand, and its shortcomings on the other hand, it is necessary to devise strategies to design preclinical experiments that yield robust evidence to guide decision making towards translation.
-We consider how to do this by simulating a preclinical research trajectory.
+[Here you need to focus more.
+You need to start with the actual problem.
+The question you want to answer is: How to switch from exploratory to confirmatory mode.
+No need to go into the shortcomings, etc.
+In the paragraph above I would also take up the exploratory confirmatory narrative from the first paragraph and find literature that describes how to do this.
+Ulis paper is pretty explicit that Type I and Type II error plays a major role here.
+Build on this and also look whether other literature makes recommendations for this.
+Describe these, but make clear that there is no satisfying solution for this important problem.
+We will provide suggestion here how to solve this problem.
+I would definitively remove many of the papers describing the other shortcomings, but focus, focus, focus...
+We can talk about this soon.
+] The stark contrast between the importance of preclinical research and its shortcomings demands a change to design preclinical experiments that yield robust evidence to guide decision making towards translation.
+Here, we consider how to do this by simulating a preclinical research trajectory.
 We understand such a trajectory as a series of experiments, successive in time, generating evidence to support the decision to carry an intervention forward to clinical testing.
-While these experiments may include different types of studies, we focus on studies examining the efficacy of a given intervention.
+Whereas these experiments may include different types of studies, we focus on studies examining the efficacy of a given intervention.
 The trajectory, for the purpose of this study, comprises exploratory studies as the first stage and within-lab replications as the second stage.
-When moving closer towards a decision regarding translation to humans, more stages are needed (for example, between-lab replications, multicenter studies, inclusion of positive and negative controls), however, in this study, we consider the state and quality of evidence generated throughout the first two stages only.
+When moving closer towards a decision regarding translation to humans, more stages are needed (for example, between-lab replications, multicenter studies, inclusion of positive and negative controls), however, in this study, we consider the state and quality of evidence generated throughout the first two stages only [If you consider only this, describe also only this.
+Do not describe every little puzzle piece but only the one you are interested in in detail.].
 Specifically, we examine the effect of different decision criteria and design choices on transition rates from the exploratory to the replication stage, as well as success rates after within-lab replication.
 
-For the purpose of this project, we understand replications as studies whose outcome is taken as "diagnostic evidence about a claim from prior research" [@nosek2020replication].
+For the purpose of this project, we understand replications as studies whose outcome is taken as "diagnostic evidence about a claim from prior research" [@nosek2020replication] [As you know there was some backlash in the elife paper on this so we need to be more specific here.
+You also introduce replication here.
+Wouldn't it in this context be enough to refer to the switch between exploratory and confirmatory mode.
+You are also looking in your simulation at a direct replication without changes to design despite the sample size.
+You are in our framework thus not concerned with validity but reliability of a finding].
 In the framework of preclinical research trajectories, replications are considered part of a confirmation process.
-Given that a series of experiments is needed to confirm a hypothesis about a directional relationship (compound X causes reaction Y), replications incrementally solidify evidence supporting (or refuting) an initial claim.
+Given that a series of experiments is needed to confirm a hypothesis about a directional relationship (compound X causes reaction Y), replications incrementally solidify evidence supporting (or refuting) an initial claim [I think we need replication as a term, but perhaps not in such a detailed manner].
 
@@
 This can be taken from the literature or a pilot study.
-In both cases, the effect size estimate used is most likely inflated due to reasons outlined above.
+In both cases, the effect size estimate used is most likely inflated due to reasons outlined above [readers will have forgotten about these reasons already, please also cite literature here that shows an inflated effect size].
 Employing an overestimated effect size for sample size calculation given a desired power level (for example, 0.8) and significance threshold (for example, .05), will yield a sample size that is too low to reliably detect effects.
 This is especially true if true underlying effects are small.
-When statistical power is low, non-replications will often denote a false negative under the assumption that an underlying effect exists.
+When statistical power is low, non-replications will often denote a false negative under the assumption that an underlying effect exists [Under the assumption that an effect exists this will always be a false negative.
+rewrite to remove ambiguity].
 An insufficiently powered replication is not suited to support or refute a previously made claim, as success and failure might be accounted for simply by inference errors.
@@
 Efforts to promote and implement the 3R’s must not come at the cost of scientific value.
+[This comes pretty late, also the beginning of the next paragraph could be higher up.
+If researchers balance exploratory and confirmatory research they need to consider ethical, time, and budget constraints.]
 
@@
 After an initial exploratory study, a first decision identifies experiments for replication.
-In our simulation, we employed two different decision criteria that indicate when one should move from the exploratory stage to the replication stage.
+In our simulation, we employed two different decision criteria that indicate when one should move from the exploratory stage to the replication [if you use this lingo, it should match the one in the intro] stage.
 If a decision has been made to replicate an initial study, we applied two approaches to determine the sample size for a replication study (smallest effect size of interest (SESOI) and standard power analysis), as outlined in detail below.
@@
 a Cohen's *d* of 0.5).
-The two distributions reflect different research fields and cover vastly different amounts of data.
+The two distributions reflect different research fields and cover vastly different amounts of data [the last is not clear, what does amount of  data mean].
 
@@
 We acknowledge that the effect sizes were mainly extracted from human studies.
-However, in large parts the distribution is in agreement with effect sizes reported to be typical of (some areas of) preclinical research (find good ref here).
-Moreover, this collection of empirical effect sizes is unique in its size and we reasoned that it constitutes a good refrence (# or something similar to motivate this choice; or maybe mention this only later in discussion as limitation).
+However, in large parts the distribution is in agreement with effect sizes reported to be typical of (some areas of) preclinical research (find good ref here)[This could be infectious diseases, usually antibiotics are all or nothing effects].
+~~Moreover, this collection of empirical effect sizes is unique in its size and we reasoned that it constitutes a good refrence (# or something similar to motivate this choice; or maybe mention this only later in discussion as limitation).
+~~
 
 As our study is concerned specifically with preclinical research, we chose a second distribution of empirical effect sizes to represent one field of the preclinical realm.
-Carneiro et al.'s (2018) [@carneiro2018effect] study sytematically examined effect sizes in the rodent fear conditioning literature.
+Carneiro et al.'s (2018) [@carneiro2018effect] study systematically examined effect sizes in the rodent fear conditioning literature.
 Effect sizes were extracted from 410 experiments published in PubMed in 2013.
 The publication included a data file containing all extracted effect sizes.
-After removing NA's, the data set consisted of 336 effect sizes, again, calculated as Cohen's *d*.
+After removing missing values, the data set consisted of 336 effect sizes, again, calculated as Cohen's *d*.
 The effect sizes range from `r round(min(ES_data_Carneiro$D), 2)` to `r round(max(ES_data_Carneiro$D), 2)`, and have a median of `r round(median(ES_data_Carneiro$D), 2)` (Figure \@ref(fig:histograms)b).
@@
 We will therefore refer to this distribution as "pessimistic".
+[in the limitations section you could cite the Bioarxiv paper and compare the two distributions to support your choice and that it may not be overly pessimistic.
+Particular in fields where there was no progress like neurodegenerative diseases]
 
@@
 
-\ \ \ \ From each of the two distributions, we drew 10000 samples of effect sizes from which we created 10000 study data sets using R's `rnorm` function (ref or better list of all packages used for simulation with credits).
-Each data set comprised data of two groups consisting of ten experimental units each.
-An experimental unit (EU) is the entity that is randomly and independently assigned to one of the experimental groups [@lazic2018exactly].
-We chose a number of ten EUs based on reported sample sizes in preclinical studies [@howells2014bringing].
+\ \ \ \ From each of the two distributions, we drew 10000 samples of effect sizes from which we created 10000 study data sets using ~~R's `rnorm` function (ref or better list of all packages used for simulation with credits)~~.
+Each data set comprised data of two groups consisting of ten experimental units each drawn from a normal distribution.
+~~An experimental unit (EU) is the entity that is randomly and independently assigned to one of the experimental groups [@lazic2018exactly].~~ We chose a number of ten EUs based on reported sample sizes in preclinical studies [@howells2014bringing].
+[For the limitation section we need to argue that different initial sample sizes like 7 or 15 will not change results dramatically].
 Our simulated design mimics a comparison between two groups where one receives an intervention and the other functions as a control group.
 The study data sets are compared using a two-sided two-sample *t*-test.
-This procedure constituted the exploratory study in our preclinical research trajectory.
-From the exploratory study results, we extracted the *p*-values and 95 percent confidence intervals (CI).
+From these exploratory study results, we extracted the *p*-values and 95 percent confidence intervals (CI).
 We then employed two different criteria based on the *p*-value or 95 percent CI, respectively, to decide whether to continue to a replication.
@@
 This approach emphasizes the importance of effect sizes rather than statistical significance to evaluate an intervention's effect.
-Further, we expected this approach to be more lenient than statistical significance (at least if the significance threshold was set at $\alpha$ = .05) and to allow a broader range of effect sizes to pass on to be further investigated.
+Further, we expected this approach to be more lenient than statistical significance (at least if the significance threshold was set at $\alpha$ = .05) and to allow a broader range of effect sizes to pass on to be further investigated [Perhaps we need to enrich Figure 1 with more detail, we should brainstorm on a whiteboard for this] .
 
@@
 In statistical terms, our SESOI was set such that the replication study would have a power of .50 to detect an effect of this size (# explain in more detail and find good ref to motivate this: Lakens).
-Consequently, in the first approach, the replication sample size was dependent on the outcome of the exploratory study, whereas using a SESOI always yielded the same sample size regardless of the exploratory effect size (e.g.
+[This is mainly to ensure that the likelihood of a type I error below this threshold is negligible.
+The goal is to reduce Type I error in this second phase.] Consequently, in the first approach, the replication sample size was dependent on the outcome of the exploratory study, whereas using a SESOI always yielded the same sample size regardless of the exploratory effect size (e.g.
 23 EUs in each group for a SESOI of 0.5).
@@
 
-\ \ \ \ For each of the studies that met the decision criterion after the exploratory study (either *p* $\leq$ .05 or SESOI within the 95 percent CI of the exploratory effect size estimate) a replication study was performed.
-The number of replication studies conducted varied with the decision criterion used and, in case of the criterion employing a SESOI, also with the SESOI (0.5 and 1.0).
+\ \ \ \ For each of the studies that met the decision criterion after the exploratory study (either *p* $\leq$ .05 or SESOI within the 95 percent CI of the exploratory effect size estimate), a replication study was performed.
+The number of replication studies conducted varied with the decision criterion and, in case of the criterion employing a SESOI, also with the SESOI (0.5 and 1.0).
 A replication study was performed as a one-sided two-sample *t*-test, where the number of animals in each group was determined by the approach to calculate the sample size.
@@
 The second trajectory applies a SESOI as decision criterion to move from the exploratory to the replication stage, and to calculate the sample size for the replication.
-In the following, we refer to the first trajectory as T1 and to the second as T2.
+In the following, we refer to the first trajectory as T1 [I think we need a good name for the trajectories as this will help readers keep track] and to the second as T2.
 We have stored data, results, and figures of all four trajectories in an online repository (insert URL here).
@@
 ### Number of animals needed in the replication
-
+[something missing here?]
 ### Positive predictive value across trajectory
@@
 If, as outlined in the introduction, throughout the preclinical research trajectory evidence for an initial claim is strengthened, we would observe an increased positive predictive value compared to pre-study odds.
+[it will be important to have supplementary material for low SESOI as well.]
 
@@
 \ \ \ \ The false positive rate (FPR) is the proportion of tests that detected a positive result when there was no true effect present.
-In our simulation, a false positive correponds to a significant finding given an underlying true effect smaller than 0.5 or 1.0, respectively.
+In our simulation, a false positive corresponds to a significant finding given an underlying true effect smaller than 0.5 or 1.0, respectively.
+[We need to motivate this.
+In a standard approach there is no SESOI.
+So every effect that is significant will be considered meaningful.
+The truth is, however, that studies powered for a certain effect in the standard paradigm, so every effect below this threshold should be considered a false positive.
+This is a point to discuss!]
 
@@
 \ \ \ \ The false negative rate (FNR) is the proportion of tests that missed a true positive relationship.
-In our simulation, this correponds to non-significant results given that the underlying true effect was greater or equal to 0.5 or 1.0, respectively.
+In our simulation, this correponds to non-significant results given that the underlying true effect was greater or equal to 0.5 or 1.0, respectively [or again the effect the study powered for].
 
@@
 #figure?
-This has consequences for the sample size calculation as well as the PPV.
+This has consequences for the sample size calculation as well as the PPV [This sounds already a bit like disussion].
 
@@
 Applying this decision criterion resulted in `r outcomes_all_distributions_init_samp_size_10$rep_attempts[6] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[5] / 100` percent of experiments moving to replication in case of the "optimistic" distribution and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[2] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[1] / 100` percent for thes "pessimistic" distribution based on an SESOI of 0.5 and 1.0, respectively.
-Compared to the conventional significance threshold, the range of effect size estimates that proceeded to replication shifts and inlcudes less extreme values, and a lot more values closer to zero.
+Compared to the conventional significance threshold, the range of effect size estimates that proceeded to replication shifted and included less extreme values, and more values closer to zero.
 #figure?
@@
 
-```{r animalnumbers, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "Number of animals needed in replication.
+```{r animalnumbers, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "Number of animals in replication [give some description of what we see in the Figure, speaking captions!].
 Error bars represent standard deviations.

### 2020-11-10 Manuscript/Manuscript_intro_methods_results.Rmd
first internal revision and supplementary plots

--- 2020-09-23 Manuscript/Manuscript_intro_methods_results.Rmd
@@
 
-\ \ \ \ Preclinical research stands at the beginning of the drug development process.
-Its purpose is to detect promising candidate drugs out of hundreds of potential interventions on the one hand, and to generate robust evidence to support translation to humans on the other hand.
-It has been argued that it is crucial to acknowledge the two operating modes of preclinical research in order for it to be conducted rigorously and interpreted appropriately [@kimmelman2014distinguishing].
-Early-stage preclinical experiments are exploratory.
-Their aim is to generate hypotheses that can at a later stage be tested under more strict conditions [@landis2012call].
-At early stages in drug devlopment, preclinical research also plays the role of a gatekeeper -- straining out promising interventions and denying access to further investigation to non-effective ones.
-It is vital, however, that decision criteria are lenient enough to allow a range of effect sizes to pass, as by that time it is unclear which effect sizes to expect.
+\ \ \ \ Preclinical research is essential for identifying promising interventions and to generate robust evidence to support translation to humans.
+The experiments in a preclinical setting are conducted  in  two different operating modes [@kimmelman2014distinguishing].
+Early-stage preclinical experiments are *exploratory* with the aim to generate hypotheses.
+These are tested at a later stage under more strict conditions in *confirmatory* mode [@landis2012call].
+At early stages of treatment discovery, preclinical research also plays the role of a gatekeeper -- straining out promising interventions and denying access to further investigation to non-effective ones.
+It is vital, however, that decision criteria are lenient enough to allow also uncertain effects to pass.
 If the filters applied at this stage are too strict, chances are that potentially effective candidates are falsely eliminated.
 
-If animal models of diseases provide the necessary criteria for proceeding drug development to a Phase I clinical trial, one would expect that they predict efficacy and safety in humans, at least to a certain extent.
+Animal models are designed to enable an informed decision to engage in a clinical trial, so 
+one would expect that they predict efficacy and safety in humans, at least to a certain extent.
 However, decades of failed clinical trials prove otherwise.
 A report of clinical development success rates from 2006 through to 2015 documents low success rates for Phase II trials, with only 30.7 % of candidate drugs progressing to Phase III [@thomas2016clinical].
-A look at specific medical conditions shows an even bleaker picture.
-Neurological diseases like amyotrophic lateral sclerosis (ALS), Alzheimer's disease (AD), and stroke, as well as cancer have kept researchers busy for decades with few to no advances with regard to effective treatments [@dirnagl2016thomas; @moreno2013can; @mullane2019preclinical; @perrin2014preclinical; @petrov2017clinical; @van2019revised].
+For specific medical conditions like amyotrophic lateral sclerosis (ALS), Alzheimer's disease (AD), and stroke, decades of research have yielded surprisingly few advances with regard to effective treatments [@dirnagl2016thomas; @mullane2019preclinical; @perrin2014preclinical; @petrov2017clinical; @van2019revised].
 
 Preclinical research has increasingly come under scrutiny with many reports pointing out low study quality, lack of scientific rigor and high levels of irreproducibility [@freedman2014increasing].
-Among the issues raised are selective reporting of results [@sena2010publication], insufficient reporting on measures to reduce bias [@vogt2016authorization], low sample sizes [@macleod2008evidence; @howells2014bringing] and concomitant increased false positive and false negative rates [@button2013power].
+Among the issues raised are (selective reporting of results [@sena2010publication], insufficient reporting on measures to reduce bias [@vogt2016authorization],) low sample sizes [@macleod2008evidence; @howells2014bringing] and concomitant increased false positive and false negative rates [@button2013power].
 These shortcomings threaten the replicability of preclinical results and hamper translation of promising findings to humans [@begley2012drug; @prinz2011believe].
@@
 
-Given the importance of preclinical research on the one hand, and its shortcomings on the other hand, it is necessary to devise strategies to design preclinical experiments that yield robust evidence to guide decision making towards translation.
-We consider how to do this by simulating a preclinical research trajectory.
+\textcolor{red}{[Here you need to focus more.
+You need to start with the actual problem.
+The question you want to answer is: How to switch from exploratory to confirmatory mode.
+No need to go into the shortcomings, etc.
+In the paragraph above I would also take up the exploratory confirmatory narrative from the first paragraph and find literature that describes how to do this.
+Ulis paper is pretty explicit that Type I and Type II error plays a major role here.
+Build on this and also look whether other literature makes recommendations for this.
+Describe these, but make clear that there is no satisfying solution for this important problem.
+We will provide suggestion here how to solve this problem.
+I would definitively remove many of the papers describing the other shortcomings, but focus, focus, focus...
+We can talk about this soon.
+]}
+The stark contrast between the importance of preclinical research and its shortcomings demands a change to design preclinical experiments that yield robust evidence to guide decision making towards translation.
+Here, we consider how to do this by simulating a preclinical research trajectory.
 We understand such a trajectory as a series of experiments, successive in time, generating evidence to support the decision to carry an intervention forward to clinical testing.
-While these experiments may include different types of studies, we focus on studies examining the efficacy of a given intervention.
-The trajectory, for the purpose of this study, comprises exploratory studies as the first stage and within-lab replications as the second stage.
-When moving closer towards a decision regarding translation to humans, more stages are needed (for example, between-lab replications, multicenter studies, inclusion of positive and negative controls), however, in this study, we consider the state and quality of evidence generated throughout the first two stages only.
+Whereas these experiments may include different types of studies, we focus on studies examining the efficacy of a given intervention.
+The trajectory, for the purpose of this study, comprises exploratory studies as the first stage and within-lab replications as the second stage and we consider the state and quality of evidence generated throughout the first two stages only.
 Specifically, we examine the effect of different decision criteria and design choices on transition rates from the exploratory to the replication stage, as well as success rates after within-lab replication.
+**maybe** remove (When moving closer towards a decision regarding translation to humans, more stages are needed (for example, between-lab replications, multicenter studies, inclusion of positive and negative controls))
 
 For the purpose of this project, we understand replications as studies whose outcome is taken as "diagnostic evidence about a claim from prior research" [@nosek2020replication].
-In the framework of preclinical research trajectories, replications are considered part of a confirmation process.
+\textcolor{red}{[As you know there was some backlash in the elife paper on this so we need to be more specific here.
+You also introduce replication here.
+Wouldn't it in this context be enough to refer to the switch between exploratory and confirmatory mode.
+You are also looking in your simulation at a direct replication without changes to design despite the sample size.
+You are in our framework thus not concerned with validity but reliability of a finding]} In the framework of preclinical research trajectories, replications are considered part of a confirmation process.
 Given that a series of experiments is needed to confirm a hypothesis about a directional relationship (compound X causes reaction Y), replications incrementally solidify evidence supporting (or refuting) an initial claim.
+\textcolor{red}{[I think we need replication as a term, but perhaps not in such a detailed manner]}
 
@@
 In both cases, the effect size estimate used is most likely inflated due to reasons outlined above.
-Employing an overestimated effect size for sample size calculation given a desired power level (for example, 0.8) and significance threshold (for example, .05), will yield a sample size that is too low to reliably detect effects.
+\textcolor{red}{[readers will have forgotten about these reasons already, please also cite literature here that shows an inflated effect size]} Employing an overestimated effect size for sample size calculation given a desired power level (for example, 0.8) and significance threshold (for example, .05), will yield a sample size that is too low to reliably detect effects.
 This is especially true if true underlying effects are small.
 When statistical power is low, non-replications will often denote a false negative under the assumption that an underlying effect exists.
-An insufficiently powered replication is not suited to support or refute a previously made claim, as success and failure might be accounted for simply by inference errors.
+\textcolor{red}{[Under the assumption that an effect exists this will always be a false negative.
+rewrite to remove ambiguity]} An insufficiently powered replication is not suited to support or refute a previously made claim, as success and failure might be accounted for simply by inference errors.
 
@@
 Efforts to promote and implement the 3R’s must not come at the cost of scientific value.
+\textcolor{red}{ [This comes pretty late, also the beginning of the next paragraph could be higher up.
+If researchers balance exploratory and confirmatory research they need to consider ethical, time, and budget constraints.]}
 
@@
 After an initial exploratory study, a first decision identifies experiments for replication.
-In our simulation, we employed two different decision criteria that indicate when one should move from the exploratory stage to the replication stage.
+In our simulation, we employed two different decision criteria that indicate when one should move from the exploratory stage to the replication stage \textcolor{red}{[if you use this lingo, it should match the one in the intro]}.
 If a decision has been made to replicate an initial study, we applied two approaches to determine the sample size for a replication study (smallest effect size of interest (SESOI) and standard power analysis), as outlined in detail below.
@@
 a Cohen's *d* of 0.5).
-The two distributions reflect different research fields and cover vastly different amounts of data.
+The two distributions reflect different research fields and cover vastly different amounts of data \textcolor{red}{[the last is not clear, what does amount of  data mean]}.
 
@@
 We acknowledge that the effect sizes were mainly extracted from human studies.
-However, in large parts the distribution is in agreement with effect sizes reported to be typical of (some areas of) preclinical research (find good ref here).
+However, in large parts the distribution is in agreement with effect sizes reported to be typical of (some areas of) preclinical research (find good ref here \textcolor{red}{[This could be infectious diseases, usually antibiotics are all or nothing effects]}).
 Moreover, this collection of empirical effect sizes is unique in its size and we reasoned that it constitutes a good refrence (# or something similar to motivate this choice; or maybe mention this only later in discussion as limitation).
+
+~~Moreover, this collection of empirical effect sizes is unique in its size and we reasoned that it constitutes a good refrence (# or something similar to motivate this choice; or maybe mention this only later in discussion as limitation).
+~~
 
@@
 Further, we expected this approach to be more lenient than statistical significance (at least if the significance threshold was set at $\alpha$ = .05) and to allow a broader range of effect sizes to pass on to be further investigated.
+\textcolor{red}{[Perhaps we need to enrich Figure 1 with more detail, we should brainstorm on a whiteboard for this]}
 
@@
 In an alternative approach, we employed the same SESOI used as decision criterion earlier.
-In statistical terms, our SESOI was set such that the replication study would have a power of .50 to detect an effect of this size (# explain in more detail and find good ref to motivate this: Lakens).
+In statistical terms, our SESOI was set such that the replication study would have a power of .50 to detect an effect of this size (# explain in more detail and find good ref to motivate this: Lakens?
+\textcolor{red}{[This is mainly to ensure that the likelihood of a type I error below this threshold is negligible.
+The goal is to reduce Type I error in this second phase.]}).
 Consequently, in the first approach, the replication sample size was dependent on the outcome of the exploratory study, whereas using a SESOI always yielded the same sample size regardless of the exploratory effect size (e.g.
@@
 
-\ \ \ \ For each of the studies that met the decision criterion after the exploratory study (either *p* $\leq$ .05 or SESOI within the 95 percent CI of the exploratory effect size estimate) a replication study was performed.
+\ \ \ \ For each of the studies that met the decision criterion after the exploratory study (either *p* $\leq$ .05 or SESOI within the 95 percent CI of the exploratory effect size estimate), a replication study was performed.
 The number of replication studies conducted varied with the decision criterion used and, in case of the criterion employing a SESOI, also with the SESOI (0.5 and 1.0).
@@
 The second trajectory applies a SESOI as decision criterion to move from the exploratory to the replication stage, and to calculate the sample size for the replication.
-In the following, we refer to the first trajectory as T1 and to the second as T2.
+In the following, we refer to the first trajectory as T1 and to the second as T2 \textcolor{red}{[I think we need a good name for the trajectories as this will help readers keep track]}.
 We have stored data, results, and figures of all four trajectories in an online repository (insert URL here).
@@
 
+\textcolor{red}{[something missing here?]}
+
 ### Positive predictive value across trajectory
@@
 If, as outlined in the introduction, throughout the preclinical research trajectory evidence for an initial claim is strengthened, we would observe an increased positive predictive value compared to pre-study odds.
+\textcolor{red}{[it will be important to have supplementary material for low SESOI as well.]}
 
@@
 \ \ \ \ The false positive rate (FPR) is the proportion of tests that detected a positive result when there was no true effect present.
-In our simulation, a false positive correponds to a significant finding given an underlying true effect smaller than 0.5 or 1.0, respectively.
+In our simulation, a false positive corresponds to a significant finding given an underlying true effect smaller than 0.5 or 1.0, respectively.
+\textcolor{red}{[We need to motivate this.
+In a standard approach there is no SESOI.
+So every effect that is significant will be considered meaningful.
+The truth is, however, that studies powered for a certain effect in the standard paradigm, so every effect below this threshold should be considered a false positive.
+This is a point to discuss!]}
 
@@
 In our simulation, this correponds to non-significant results given that the underlying true effect was greater or equal to 0.5 or 1.0, respectively.
+\textcolor{red}{ [or again the effect the study powered for]}
 
@@
 This has consequences for the sample size calculation as well as the PPV.
+\textcolor{red}{[This sounds already a bit like disussion]}
 
@@
 
+\textcolor{red}{[give some description of what we see in the Figure, speaking captions!]}
 ```{r animalnumbers, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "Number of animals needed in replication.
@@
 Dashed lines indicate pre-study odds based on empirical effect size distributions.", fig.align = 'center'}
-facet_names <- c(
-  "0.5" = "SESOI = 0.5",
+facet_names <- 
+  c("0.5" = "SESOI = 0.5",
     "1" = "SESOI = 1.0")
@@
 
+```{r data_effect_sizes_after_replication, echo = FALSE, message = FALSE}
+load("effect_sizes_with_CI_after_replication.RData")
+
+plot_data <-
+  final_init_samp_size_10 %>% 
+  group_by(distribution, trajectory) %>% 
+  summarize(mean_effect = mean(effect),
+            median_effect = median(effect),
+            mean_CI_lower = mean(CI_lower),
+            mean_CI_upper = mean(CI_upper),
+            median_ES_true = median(ES_true),
+            mean_ES_true = mean(ES_true))
+```
+
+```{r, plot_effect_size_precision, echo = FALSE, message = FALSE, out.width = "80%", fig.align = 'center'}
+colors <- c("Median of true effect sizes" = "blue", 
+            "Mean of empirical effect sizes \nwith 95% CI" = "red")
+
+plot_ES_precision <- 
+  ggplot(data = plot_data,
+         aes(x = trajectory)) +
+  facet_wrap(~ distribution) +
+  geom_errorbar(aes(ymin = mean_effect - mean_CI_lower,
+                    ymax = mean_effect + mean_CI_upper),
+                width = 0.09,
+                color = "red",
+                position = "dodge") +
+  geom_point(aes(y = mean_effect, color = "Mean of empirical effect sizes \nwith 95% CI"), size = 2.5) +
+  geom_point(aes(y = median_ES_true, color = "Median of true effect sizes"), size = 2.5) +
+  labs(x = "Trajectory",
+       y = "Mean effect size (with 95% CI)",
+       color = " ") +
+  scale_x_discrete(labels = c("T1", "T2 (0.5)", "T2 (1.0)")) +
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 14)) +
+  theme(axis.title.y = element_text(size = 14)) +
+  theme(axis.text.x = element_text(size = 12, colour = "black")) +
+  theme(axis.text.y = element_text(size = 12, colour = "black")) +
+  theme(strip.text.x = element_text(size = 15, colour = "black", face = "bold")) +
+  theme(strip.background = element_rect(fill = "white", color = "black")) +
+  scale_color_manual(values = colors) + 
+  theme(legend.position =  "top") +
+  theme(legend.text = element_text(size = 12))
+
+plot_ES_precision
+```
+
 \ \ \ \ text

### 2020-11-10 Manuscript/internal_revision/analysis_effects_standard_trajectory.Rmd
first internal revision and supplementary plots

---
title: "Analysis of effect sizes in standard trajectory"
#csl: biomed-central.csl
indent: yes
output: bookdown::pdf_document2
toc: false
linestretch: 1.5
fontsize: 12pt
header-includes:
       - \usepackage{subfig}
#bibliography: refs_repro.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(bookdown)
library(knitr)
library(pwr)
library(tidyverse)
```

```{r data_effect_sizes_after_replication, echo = FALSE, message = FALSE, results = 'hide'}
load("effect_sizes_with_CI_after_replication.RData")

# filter all effects > 5 in T1 (standard trajectory)
T1_data <-
  final_init_samp_size_10 %>% 
  filter(trajectory == "T1") %>% 
  filter(effect < 5)

mean_N <- 
  T1_data %>% 
  group_by(distribution) %>% 
  summarize(mean_N = mean(rep_sample_size))

ceiling(mean(T1_data$rep_sample_size))

powered_d_50 <-
  pwr.t.test(n = ceiling(mean(T1_data$rep_sample_size)),
             d = NULL, sig.level = .05, power = .5,
             type = "two.sample",
             alternative = "greater")$d

powered_d_80 <-
  pwr.t.test(n = ceiling(mean(T1_data$rep_sample_size)),
             d = NULL, sig.level = .05, power = .8,
             type = "two.sample",
             alternative = "greater")$d
```

```{r plot_effects_powered_at_50, echo = FALSE, message = FALSE, out.width = "80%", fig.align = 'center'}
ggplot(data = T1_data,
       aes(x = distribution, y = effect)) +
  # facet_wrap(~ distribution) +
  geom_violin(position = position_dodge(width = 0.9)) +
  geom_rect(fill = "lightgrey", alpha = .01, 
            aes(xmin = -Inf, xmax = Inf, 
                ymin = powered_d_50, ymax = Inf)) +
  # geom_rect(fill = "lightgrey", alpha = .01, 
  #           aes(xmin = -Inf, xmax = Inf, 
  #               ymin = powered_d_50, ymax = powered_d_80)) +
  labs(x = "Distribution",
       y = expression(paste("Effect sizes (Cohen's ", italic("d"), ") "))) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 14)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.text.x = element_text(size = 12, colour = "black")) +
  theme(axis.text.y = element_text(size = 12, colour = "black"))
  # geom_quasirandom(dodge.width = 0.9,
  #                  varwidth = FALSE,
  #                  size = .5, alpha = 0.3)
```

```{r plot_effects_powered_at_80, echo = FALSE, message = FALSE, out.width = "80%", fig.align = 'center'}
ggplot(data = T1_data,
       aes(x = distribution, y = effect)) +
  # facet_wrap(~ distribution) +
  geom_violin(position = position_dodge(width = 0.9)) +
  # geom_rect(fill = "lightgrey", alpha = .01, 
  #           aes(xmin = -Inf, xmax = Inf, 
  #               ymin = powered_d_50, ymax = Inf)) +
  geom_rect(fill = "lightgrey", alpha = .01,
            aes(xmin = -Inf, xmax = Inf,
                ymin = powered_d_80, ymax = Inf)) +
  labs(x = "Distribution",
       y = expression(paste("Effect sizes (Cohen's ", italic("d"), ") "))) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 14)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.text.x = element_text(size = 12, colour = "black")) +
  theme(axis.text.y = element_text(size = 12, colour = "black"))
  # geom_quasirandom(dodge.width = 0.9,
  #                  varwidth = FALSE,
  #                  size = .5, alpha = 0.3)
```

### 2020-11-10 Manuscript/internal_revision/supplemetary_plots.Rmd
first internal revision and supplementary plots

---
title: 'Supplementary plots'
#csl: biomed-central.csl
indent: yes
output: bookdown::pdf_document2
toc: false
linestretch: 1.5
fontsize: 12pt
header-includes:
       - \usepackage{subfig}
#bibliography: refs_repro.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(bookdown)
library(knitr)
library(pwr)
library(tidyverse)
```

# PPV for all trajectories and SESOI  
Pessimistic scenario (Carneiro distribution)

```{r data_Carneiro, include = FALSE, cache = FALSE}
outcomes <- load("outcomes_Carneiro_all_trajectories.RData")
```

```{r, PPVplot_Carneiro, echo = FALSE, fig.width = 45, fig.height = 10, out.width = "100%", fig.align = 'center'}
facet_names <- 
  c("0.3" = "SESOI = 0.3",
    "0.5" = "SESOI = 0.5",
    "0.7" = "SESOI = 0.7",
    "1" = "SESOI = 1.0")

plot_PPV <- 
  ggplot(data = outcomes_10EU,
         aes(x = trajectory, y = PPV_pop_prev)) + 
  geom_point(size = 3) + 
  facet_wrap(~ SESOI, nrow = 1, ncol = 4,  labeller = labeller(.rows = facet_names)) + 
  labs(x = "Trajectory", y = "Positive predictive value") +
  scale_x_discrete(labels = c("SESOI within CI \nSESOI",
                              "Significance \nSESOI",
                              "SESOI within CI \nStandard",
                              "Significance \nStandard")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 35)) +
  theme(axis.title.y = element_text(size = 35)) +
  theme(axis.text.x = element_text(size = 28, colour = "black")) +
  theme(axis.text.y = element_text(size = 28, colour = "black")) +
  theme(strip.text.x = element_text(size = 25, colour = "black", face = "bold")) +
  theme(strip.text.y = element_text(size = 25, colour = "black", face = "bold")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.title = element_text(size = 15, face = "bold")) +
  theme(legend.text = element_text(size = 14))

hlines <- data.frame(pre_study_odds = c(outcomes_10EU$Prevalence[1], outcomes_10EU$Prevalence[5],
                                        outcomes_10EU$Prevalence[9], outcomes_10EU$Prevalence[13]), 
                     # distribution   = c(rep(plot_data$distribution[1], 2), 
                     #                    rep(plot_data$distribution[5], 2)),
                     SESOI          = rep(c("1", "0.7", "0.5", "0.3")))

plot_PPV <- 
  plot_PPV + 
  geom_hline(data = hlines, 
             aes(yintercept = pre_study_odds),
             color = "red", lty = 2, size = .5)

plot_PPV
```

Optimistic scenario (Szucs distribution)

```{r data_Szucs, include = FALSE, cache = FALSE}
outcomes <- load("outcomes_Szucs_all_trajectories.RData")
```

```{r, PPVplot_Szucs, echo = FALSE, fig.width = 45, fig.height = 10, out.width = "100%", fig.align = 'center'}
facet_names <- 
  c("0.3" = "SESOI = 0.3",
    "0.5" = "SESOI = 0.5",
    "0.7" = "SESOI = 0.7",
    "1" = "SESOI = 1.0")

plot_PPV <- 
  ggplot(data = outcomes_10EU,
         aes(x = trajectory, y = PPV_pop_prev)) + 
  geom_point(size = 3) + 
  facet_wrap(~ SESOI, nrow = 1, ncol = 4,  labeller = labeller(.rows = facet_names)) + 
  labs(x = "Trajectory", y = "Positive predictive value") +
  scale_x_discrete(labels = c("SESOI within CI \nSESOI",
                              "Significance \nSESOI",
                              "SESOI within CI \nStandard",
                              "Significance \nStandard")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 35)) +
  theme(axis.title.y = element_text(size = 35)) +
  theme(axis.text.x = element_text(size = 28, colour = "black")) +
  theme(axis.text.y = element_text(size = 28, colour = "black")) +
  theme(strip.text.x = element_text(size = 25, colour = "black", face = "bold")) +
  theme(strip.text.y = element_text(size = 25, colour = "black", face = "bold")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.title = element_text(size = 15, face = "bold")) +
  theme(legend.text = element_text(size = 14))

hlines <- data.frame(pre_study_odds = c(outcomes_10EU$Prevalence[1], outcomes_10EU$Prevalence[5],
                                        outcomes_10EU$Prevalence[9], outcomes_10EU$Prevalence[13]), 
                     # distribution   = c(rep(plot_data$distribution[1], 2), 
                     #                    rep(plot_data$distribution[5], 2)),
                     SESOI          = rep(c("1", "0.7", "0.5", "0.3")))

plot_PPV <- 
  plot_PPV + 
  geom_hline(data = hlines, 
             aes(yintercept = pre_study_odds),
             color = "red", lty = 2, size = .5)

plot_PPV
```

### 2020-11-10 Manuscript/internal_revision/analysis_all_trajectories_Carneiro.R

@@ -1,0 +1,64 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+
+
+rm(list = ls())
+
+
+library(pwr)
+library(tidyverse)
+
+outcomes <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_outcomes_all_SESOI_combined")
+
+outcomes$trajectory <- interaction(outcomes$decision_crit, outcomes$sampsize_approach)
+
+outcomes_10EU <- 
+  outcomes %>% 
+  filter(init_sample_size == 10)
+
+# save(outcomes_10EU,
+#           file = "./Manuscript/internal_revision/outcomes_Carneiro_all_trajectories.RData")
+
+
+facet_names <- 
+  c("0.3" = "SESOI = 0.3",
+    "0.5" = "SESOI = 0.5",
+    "0.7" = "SESOI = 0.7",
+    "1" = "SESOI = 1.0")
+
+
+plot_PPV <- 
+  ggplot(data = outcomes_10EU,
+         aes(x = trajectory, y = PPV_pop_prev)) + 
+  geom_point() + 
+  facet_wrap(~ SESOI, nrow = 1, ncol = 4,  labeller = labeller(.rows = facet_names)) + 
+  labs(x = "Trajectory", y = "Positive predictive value") +
+  scale_x_discrete(labels = c("SESOI within CI \nSESOI",
+                              "Significance \nSESOI",
+                              "SESOI within CI \nStandard",
+                              "Significance \nStandard")) +
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 14)) +
+  theme(axis.title.y = element_text(size = 14)) +
+  theme(axis.text.x = element_text(size = 12, colour = "black")) +
+  theme(axis.text.y = element_text(size = 12, colour = "black")) +
+  theme(strip.text.x = element_text(size = 15, colour = "black", face = "bold")) +
+  theme(strip.text.y = element_text(size = 15, colour = "black", face = "bold")) +
+  theme(strip.background = element_rect(fill = "white", color = "black")) +
+  theme(legend.title = element_text(size = 15, face = "bold")) +
+  theme(legend.text = element_text(size = 14))
+
+hlines <- data.frame(pre_study_odds = c(outcomes_10EU$Prevalence[1], outcomes_10EU$Prevalence[5],
+                                        outcomes_10EU$Prevalence[9], outcomes_10EU$Prevalence[13]), 
[…]

### 2020-11-10 Manuscript/internal_revision/analysis_all_trajectories_Szucs.R

@@ -1,0 +1,53 @@
+setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+
+rm(list = ls())
+
+library(pwr)
+library(tidyverse)
+
+outcomes <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_outcomes_all_SESOI_combined")
+
+outcomes$trajectory <- interaction(outcomes$decision_crit, outcomes$sampsize_approach)
+
+
+outcomes_10EU <- 
+  outcomes %>% 
+  filter(init_sample_size == 10) %>% 
+  filter(mean_N != 25)
+
+# save(outcomes_10EU,
+#      file = "./Manuscript/internal_revision/outcomes_Szucs_all_trajectories.RData")
+
+plot_PPV <- 
+  ggplot(data = outcomes_10EU,
+         aes(x = trajectory, y = PPV_pop_prev)) + 
+  geom_point(size = 2) + 
+  facet_wrap(~ SESOI) + 
+  labs(x = "Trajectory", y = "Positive predictive value") +
+  scale_x_discrete(labels = c("SESOI within CI \nSESOI",
+                              "Significance \nSESOI",
+                              "SESOI within CI \nStandard",
+                              "Significance \nStandard")) +
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 14)) +
+  theme(axis.title.y = element_text(size = 14)) +
+  theme(axis.text.x = element_text(size = 12, colour = "black")) +
+  theme(axis.text.y = element_text(size = 12, colour = "black")) +
+  theme(strip.text.x = element_text(size = 15, colour = "black", face = "bold")) +
+  theme(strip.text.y = element_text(size = 15, colour = "black", face = "bold")) +
+  theme(strip.background = element_rect(fill = "white", color = "black")) +
+  theme(legend.title = element_text(size = 15, face = "bold")) +
+  theme(legend.text = element_text(size = 14))
+
+hlines <- data.frame(pre_study_odds = c(outcomes_10EU$Prevalence[1], outcomes_10EU$Prevalence[5],
+                                        outcomes_10EU$Prevalence[9], outcomes_10EU$Prevalence[13]), 
+                     # distribution   = c(rep(plot_data$distribution[1], 2), 
+                     #                    rep(plot_data$distribution[5], 2)),
+                     SESOI          = rep(c("1", "0.7", "0.5", "0.3"))) 
+
+
[…]

### 2020-11-10 scripts/analysis/analysis_effect_size_precision.R

@@ -4,12 +4,12 @@ rm(list = ls())
 
 library(tidyverse)
 library(compute.es)
+library(pwr)
+library(ggbeeswarm)
 
 
 ### read in data sets from the different trajectories 
 ### using fixed-N design in replication study
-
-# final <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_1.0")
 
 final_sig_Szucs <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1")
 final_sig_Szucs$distribution <- "Optimistic"
@@ -47,31 +47,30 @@ final_init_samp_size_10 <-
   final %>% 
   filter(init_sample_size == 10)
 
-### add column for outcome significant / not significant to match outcome column
-### of sequential design data set
+### add column for outcome significant / not significant
 final_init_samp_size_10$H0 <- ifelse(final_init_samp_size_10$p_value <= .05, 2, 1)
 
-
+### filter only significant outcomes 
 final_init_samp_size_10 <-
   final_init_samp_size_10 %>% 
   group_by(distribution, trajectory) %>% 
   filter(H0 == 2)
 
-# hist(final_init_samp_size_10$ES_true, breaks = 100)
-# max(final_init_samp_size_10$ES_true)
-# median(final_init_samp_size_10$ES_true)
-# mean(final_init_samp_size_10$ES_true)
-# 
-# hist(final_init_samp_size_10$effect, breaks = 100)
-# max(final_init_samp_size_10$effect)
-# median(final_init_samp_size_10$effect)
-# mean(final_init_samp_size_10$effect)
-# 
+hist(final_init_samp_size_10$ES_true, breaks = 100)
+max(final_init_samp_size_10$ES_true)
+median(final_init_samp_size_10$ES_true)
+mean(final_init_samp_size_10$ES_true)
+
+
+hist(final_init_samp_size_10$effect, breaks = 100)
+max(final_init_samp_size_10$effect)
+median(final_init_samp_size_10$effect)
+mean(final_init_samp_size_10$effect)
+
 # plot(final_init_samp_size_10$effect)
 # plot(final_init_samp_size_10$ES_true)
 
 ### compute CI around effect size
-
 final_init_samp_size_10$t_value <- ifelse(final_init_samp_size_10$t_value < 0,
                                           - final_init_samp_size_10$t_value, 
[…]

### 2020-11-10 scripts/analysis/limited_resources.R

@@ -7,7 +7,7 @@ library(viridis)
 
 #source("./scripts/calculate_outcomes_onesided_across_trajectory.R")
 
-outcomes <- read.csv(file = "./data/Carneiro_distribution_outcomes_all_SESOI_combined")
+outcomes <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_outcomes_all_SESOI_combined")
 
 n_animals    <- 1000000
 n_exp        <- 10000

### 2020-11-10 scripts/simulation/sim_fixN_Carneiro_decision_equivalence_with_SESOI.R

@@ -134,7 +134,7 @@ selection_equiv <- list()
 for (i in 1:length(samp_size_vector)) {
   
   selection_equiv[[i]] <- future_map(exploratory_data_summary[[i]], get_decision_equiv,
-                                     SESOI = 1.0)
+                                     SESOI = 0.5)
   
 }
 
@@ -164,11 +164,25 @@ dat$ES_true <- current_ES
 hist(dat$ES_true, breaks = 100)
 
 # dat <-
-#   dat %>% 
+#   dat %>%
 #   filter(init_sample_size == 10)
 # 
-# ggplot(data = dat, aes(x = dat$ES_true, fill = factor(selection_equiv))) +
+# # dat$effect <- ifelse(dat$effect < 0, -dat$effect, -dat$effect)
+# 
+# dat_large_ES <-
+#   dat %>%
+#   filter(ES_true > 0)
+# 
+# dat_selected <-
+#   dat_large_ES %>% 
+#   filter(selection_equiv == 1)
+# 
+# 6618/6844*100
+# 
+# ggplot(data = dat_large_ES, aes(x = dat_large_ES$ES_true, fill = factor(selection_equiv))) +
 #   geom_histogram(bins = 50, color = "black", size = 0.3, alpha = 0.8) +
+#   # geom_histogram(data = dat_large_ES, aes(x = ),
+#                  # bins = 50, color = "black", fill = "white")
 #   labs(x = expression(paste("Sampled effect sizes (Cohen's ", italic("d"), ") ")),
 #        y = "Frequency",
 #        fill = "Selected for \nreplication") +

### 2020-11-10 scripts/simulation/sim_fixN_Carneiro_decision_significance.R

@@ -160,13 +160,53 @@ dat <- bind_cols(df, df_sig)
 
 dat$ES_true <- current_ES
 
+dat <-
+  dat %>%
+  filter(init_sample_size == 10)
+
+# dat$effect <- ifelse(dat$effect < 0, -dat$effect, -dat$effect)
+
+dat_large_ES <-
+  dat %>% 
+  filter(ES_true > 0) 
+
+dat_selected <-
+  dat_large_ES %>% 
+  filter(selection_sig == 1)
+
+3169/6844*100
+
+ggplot(data = dat_large_ES, aes(x = dat_large_ES$ES_true, fill = factor(selection_sig))) +
+  geom_histogram(bins = 50, color = "black",
+                 size = 0.3, alpha = 0.8) +
+  labs(x = expression(paste("Sampled effect sizes (Cohen's ", italic("d"), ") ")),
+       y = "Frequency",
+       fill = "Selected for \nreplication") +
+  scale_fill_manual(breaks = c("0", "1"),
+                    labels = c("no",
+                               "yes"),
+                    values = c("lightgrey", "deeppink3")) +
+  theme_bw() +
+  theme(axis.title.x = element_blank()) +
+  theme(axis.title.y = element_text(size = 20)) +
+  theme(axis.text.x = element_text(size = 18, colour = "black")) +
+  theme(axis.text.y = element_text(size = 18, colour = "black")) +
+  # theme(strip.text.x = element_text(size = 20, colour = "black", face = "bold")) +
+  # theme(strip.text.y = element_text(size = 20, colour = "black", face = "bold")) +
+  # theme(strip.background = element_rect(fill = "white", color = "black")) +
+  theme(legend.title = element_text(size = 18, face = "bold")) +
+  theme(legend.text = element_text(size = 18)) +
+  theme(title = element_text(size = 15)) +
+  theme(legend.position = "right")
+
+
+
 # dat <-
 #   dat %>%
 #   filter(init_sample_size == 10)
 
-# 
 # dat$effect <- ifelse(dat$effect < 0, -dat$effect, -dat$effect)
-# 
+#
 # ggplot(data = dat, aes(x = dat$effect)) +
 #   geom_histogram(bins = 50, color = "black", fill = "white", size = 0.3, alpha = 0.8) +
 #   # labs(x = expression(paste("Sampled effect sizes (Cohen's ", italic("d"), ") ")),

### 2020-11-10 scripts/simulation/sim_fixN_Carneiro_replication.R

@@ -5,9 +5,9 @@ rm(list = ls())
 
 # source additional scripts
 
-source("./scripts/simulation/sim_fixN_Carneiro_decision_equivalence_with_SESOI.R")
+# source("./scripts/simulation/sim_fixN_Carneiro_decision_equivalence_with_SESOI.R")
 
-# source("./scripts/simulation/sim_fixN_Carneiro_decision_significance.R")
+source("./scripts/simulation/sim_fixN_Carneiro_decision_significance.R")
 
 dat
 
@@ -18,7 +18,7 @@ sum(dat$ES_true < 0)
 data <-
   dat %>% 
   group_by(init_sample_size, study_id) %>% 
-  filter(selection_equiv == 1)
+  filter(selection_sig == 1)
 
 sum(data$effect < 0)
 sum(data$effect > 0)
@@ -27,7 +27,7 @@ sum(data$effect == 0)
 selected <-
   data %>% 
   group_by(init_sample_size) %>% 
-  summarize(selected = sum(selection_equiv == 1))
+  summarize(selected = sum(selection_sig == 1))
 
 # now estimate sample size for replication study
 # A. Initial effect size (probably the worst solution but good as a floor benchmark, 1 in function)
@@ -39,7 +39,7 @@ for (i in 1:nrow(data)) {
 
   rep_sample_size_std[i] <-
     ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-                             method = 2, SESOI = 1.0, power = 0.5))
+                             method = 1))
 }
 
 data$rep_samp_size_std <- rep_sample_size_std
@@ -80,9 +80,9 @@ replication_data <- list()
 
 rep_exp_no <- 0
 
-# select_experiments <- which(data$selection_sig == 1)
+select_experiments <- which(data$selection_sig == 1)
 
-select_experiments <- which(data$selection_equiv == 1)
+# select_experiments <- which(data$selection_equiv == 1)
 
 select_experiments <- select_experiments[data$effect[select_experiments] >= 0 ]
 
@@ -119,24 +119,68 @@ res_summary_rep <-
              t_value = unlist(map(rep_data_summary, "t_value")),
              p_value = unlist(map(rep_data_summary, "p_value")), #[seq(1, 2*rep_exp_no, 2)],
              effect = unlist(map(rep_data_summary, "effect")),
-             ES_true = data$ES_true[select_experiments],
-             rep_attempts = rep_attempts)
[…]

### 2020-11-13 Manuscript/Manuscript_intro_methods_results.Rmd
merge edits by Ulf

@@
 
-Animal models are designed to enable an informed decision to engage in a clinical trial, so 
-one would expect that they predict efficacy and safety in humans, at least to a certain extent.
+Animal models are designed to enable an informed decision to engage in a clinical trial, so one would expect that they predict efficacy and safety in humans, at least to a certain extent.
 However, decades of failed clinical trials prove otherwise.
@@
 
+\ \ \ \ We explored different approaches to perform preclinical animal experiments via simulations.
+To this end, we modeled a simplified preclinical research trajectory from the exploratory stage to the results of a within-lab replication study (Figure \@ref(fig:trajectory)).
+Along the trajectory, there are different ways to increase the probability of not missing potentially meaningful effects.
+After an initial exploratory study, a first decision identifies experiments for replication.
+In our simulation, we employed two different decision criteria that indicate when one should move from the exploratory stage to the replication [if you use this lingo, it should match the one in the intro] stage.
+If a decision has been made to replicate an initial study, we applied two approaches to determine the sample size for a replication study (smallest effect size of interest (SESOI) and standard power analysis), as outlined in detail below.
+
 ```{r, echo = FALSE, results = FALSE}
@@
 However, in large parts the distribution is in agreement with effect sizes reported to be typical of (some areas of) preclinical research (find good ref here \textcolor{red}{[This could be infectious diseases, usually antibiotics are all or nothing effects]}).
-Moreover, this collection of empirical effect sizes is unique in its size and we reasoned that it constitutes a good refrence (# or something similar to motivate this choice; or maybe mention this only later in discussion as limitation).
-
-~~Moreover, this collection of empirical effect sizes is unique in its size and we reasoned that it constitutes a good refrence (# or something similar to motivate this choice; or maybe mention this only later in discussion as limitation).
-~~
 
 As our study is concerned specifically with preclinical research, we chose a second distribution of empirical effect sizes to represent one field of the preclinical realm.
-Carneiro et al.'s (2018) [@carneiro2018effect] study sytematically examined effect sizes in the rodent fear conditioning literature.
+Carneiro et al.'s (2018) [@carneiro2018effect] study systematically examined effect sizes in the rodent fear conditioning literature.
 Effect sizes were extracted from 410 experiments published in PubMed in 2013.
 The publication included a data file containing all extracted effect sizes.
-After removing NA's, the data set consisted of 336 effect sizes, again, calculated as Cohen's *d*.
+After removing missing values, the data set consisted of 336 effect sizes, again, calculated as Cohen's *d*.
 The effect sizes range from `r round(min(ES_data_Carneiro$D), 2)` to `r round(max(ES_data_Carneiro$D), 2)`, and have a median of `r round(median(ES_data_Carneiro$D), 2)` (Figure \@ref(fig:histograms)b).
@@
 
-\ \ \ \ From each of the two distributions, we drew 10000 samples of effect sizes from which we created 10000 study data sets using R's `rnorm` function (ref or better list of all packages used for simulation with credits).
-Each data set comprised data of two groups consisting of ten experimental units each.
-An experimental unit (EU) is the entity that is randomly and independently assigned to one of the experimental groups [@lazic2018exactly].
+\ \ \ \ From each of the two distributions, we drew 10000 samples of effect sizes from which we created 10000 study data sets.
+Each data set comprised data of two groups consisting of ten experimental units each drawn from a normal distribution.
 We chose a number of ten EUs based on reported sample sizes in preclinical studies [@howells2014bringing].
+[For the limitation section we need to argue that different initial sample sizes like 7 or 15 will not change results dramatically].
 Our simulated design mimics a comparison between two groups where one receives an intervention and the other functions as a control group.
 The study data sets are compared using a two-sided two-sample *t*-test.
-This procedure constituted the exploratory study in our preclinical research trajectory.
-From the exploratory study results, we extracted the *p*-values and 95 percent confidence intervals (CI).
+From these exploratory study results, we extracted the *p*-values and 95 percent confidence intervals (CI).
 We then employed two different criteria based on the *p*-value or 95 percent CI, respectively, to decide whether to continue to a replication.
@@
 ### Decision criteria to proceed to replication
+
 \ \ \ \ The first decision criterion employs the conventional significance threshold  ($\alpha$ = .05) to decide whether to replicate an exploratory study.
@@
 A replication study was performed as a one-sided two-sample *t*-test, where the number of animals in each group was determined by the approach to calculate the sample size.
-For a replication be considered "successful", the *p*-value had to be below the conventional significance threshold ($\alpha$ = .05)
+For a replication be considered "successful", the *p*-value had to be below the conventional significance threshold ($\alpha$ = .05).
 
@@
 Note that in case of trajectories using a SESOI to calculate sample size for replication the number of animals is fixed.", fig.align = 'center'}
+
 plot_mean_N <-
@@
 
+\ \ \ \ text
+
 ```{r data_effect_sizes_after_replication, echo = FALSE, message = FALSE}
@@
 
-\ \ \ \ text
-
 # Discussion

### 2020-11-26 Manuscript/Manuscript_intro_methods_results.Rmd
additional analysis standard trajectory

@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r, include=FALSE}
+# options(tinytex.verbose = TRUE)
 ```
@@
 
+Along the trajectory, we have to make sure that at a given stage (which implies that we are in a specific operating mode), our strategy is aligned with the goal that we have at this stage.
+At early stages, we want to find possibly working interventions.
+In statistical terms this means that we aim for a low Type II error.
+We do not want to miss effects.
+That means that our tests should be sensitive.
+This comes at the cost of specificity.
+We allow for false positives at this stage.
+Dirnagl (2016) [@dirnagl2016thomas]
+
+*In exploratory mode, we allow for more false positives. This happens naturally, as we operate with low prior odds that our hypothesis is true.
+
+* focus in exploration should rather be on proper descriptive statistics, including measures of variance and confidence intervals.
+
+*
+
 \textcolor{red}{[Here you need to focus more.
@@
 ]}
+
 The stark contrast between the importance of preclinical research and its shortcomings demands a change to design preclinical experiments that yield robust evidence to guide decision making towards translation.
@@
 This is especially true if true underlying effects are small.
-When statistical power is low, non-replications will often denote a false negative under the assumption that an underlying effect exists.
-\textcolor{red}{[Under the assumption that an effect exists this will always be a false negative.
-rewrite to remove ambiguity]} An insufficiently powered replication is not suited to support or refute a previously made claim, as success and failure might be accounted for simply by inference errors.
+When statistical power is low, non-replications denote false negatives under the assumption that an underlying effect exists.
+An insufficiently powered replication is not suited to support or refute a previously made claim, as success and failure might be accounted for simply by inference errors.
 

### 2020-11-26 Manuscript/internal_revision/supplemetary_plots.Rmd
additional analysis standard trajectory

@@
 
-```{r, PPVplot_Carneiro, echo = FALSE, fig.width = 45, fig.height = 10, out.width = "100%", fig.align = 'center'}
+```{r, PPVplot_Carneiro, echo = FALSE, out.width = "100%", fig.align = 'center'}
 facet_names <- 
@@
          aes(x = trajectory, y = PPV_pop_prev)) + 
-  geom_point(size = 3) + 
-  facet_wrap(~ SESOI, nrow = 1, ncol = 4,  labeller = labeller(.rows = facet_names)) + 
+  geom_point(size = 1.5) + 
+  facet_wrap(~ SESOI, nrow = 2, ncol = 2,  labeller = labeller(.rows = facet_names)) + 
   labs(x = "Trajectory", y = "Positive predictive value") +
@@
   theme_bw() +
-  theme(axis.title.x = element_text(size = 35)) +
-  theme(axis.title.y = element_text(size = 35)) +
-  theme(axis.text.x = element_text(size = 28, colour = "black")) +
-  theme(axis.text.y = element_text(size = 28, colour = "black")) +
-  theme(strip.text.x = element_text(size = 25, colour = "black", face = "bold")) +
-  theme(strip.text.y = element_text(size = 25, colour = "black", face = "bold")) +
-  theme(strip.background = element_rect(fill = "white", color = "black")) +
-  theme(legend.title = element_text(size = 15, face = "bold")) +
-  theme(legend.text = element_text(size = 14))
+  theme(axis.title.x = element_text(size = 11)) +
+  theme(axis.title.y = element_text(size = 11)) +
+  theme(axis.text.x = element_text(size = 7, colour = "black")) +
+  theme(axis.text.y = element_text(size = 10, colour = "black")) +
+  theme(strip.text.x = element_text(size = 11, colour = "black", face = "bold")) +
+  theme(strip.text.y = element_text(size = 11, colour = "black", face = "bold")) +
+  theme(strip.background = element_rect(fill = "white", color = "black"))
 
@@
 
+\newpage
+
 Optimistic scenario (Szucs distribution)
@@
 
-```{r, PPVplot_Szucs, echo = FALSE, fig.width = 45, fig.height = 10, out.width = "100%", fig.align = 'center'}
+```{r, PPVplot_Szucs, echo = FALSE, out.width = "100%", fig.align = 'center'}
 facet_names <- 

### 2020-11-26 Manuscript/internal_revision/analysis_all_trajectories_Carneiro.R

@@ -30,7 +30,7 @@ plot_PPV <-
   ggplot(data = outcomes_10EU,
          aes(x = trajectory, y = PPV_pop_prev)) + 
   geom_point() + 
-  facet_wrap(~ SESOI, nrow = 1, ncol = 4,  labeller = labeller(.rows = facet_names)) + 
+  facet_wrap(~ SESOI, nrow = 2, ncol = 2,  labeller = labeller(.rows = facet_names)) + 
   labs(x = "Trajectory", y = "Positive predictive value") +
   scale_x_discrete(labels = c("SESOI within CI \nSESOI",
                               "Significance \nSESOI",

### 2020-11-26 scripts/analysis/analysis_effect_size_precision.R

@@ -83,8 +83,8 @@ final_init_samp_size_10$CI_upper <- tes(t = final_init_samp_size_10$t_value,
                         n.1 = final_init_samp_size_10$rep_sample_size, 
                         n.2 = final_init_samp_size_10$rep_sample_size)$u.d
 
-save(final_init_samp_size_10,
-     file = "./Manuscript/effect_sizes_with_CI_after_replication.RData")
+# save(final_init_samp_size_10,
+#      file = "./Manuscript/effect_sizes_with_CI_after_replication.RData")
 
 # final_init_samp_size_10$median_ES_true <- median(final_init_samp_size_10$ES_true)
 # final_init_samp_size_10$median_effect <- median(final_init_samp_size_10$effect)
@@ -177,7 +177,7 @@ for(i in 1:nrow(T1_data)) {
   
   delta_powered[i] <- 
     pwr.t.test(n = T1_data$rep_sample_size[i],
-               d = NULL, sig.level = .05, power = .5,
+               d = NULL, sig.level = .05, power = .8,
                type = "two.sample",
                alternative = "greater")$d
   
@@ -190,20 +190,28 @@ T1_data_optimistic <-
   T1_data %>% 
   filter(distribution == "Optimistic")
 
-
-
 significant_effect <- data.frame(effect = T1_data_optimistic$effect)
 significant_effect$type <- "significant"
 
 powered_effect <- data.frame(effect = T1_data_optimistic$delta_powered)
 powered_effect$type <- "powered"
 
+test <- bind_rows(significant_effect, powered_effect)
+
+ggplot(data = test,
+       aes(x = type,
+           y = effect)) +
+  geom_violin()
+
+ggplot(data = test,
+       aes(x = effect, fill = type)) +
+  geom_density(alpha = 0.7) +
+  theme_bw()
+
 
 T1_data_pessimistic <-
   T1_data %>% 
   filter(distribution == "Pessimistic")
-
-
 
 significant_effect <- data.frame(effect = T1_data_pessimistic$effect)
 significant_effect$type <- "significant"
@@ -223,5 +231,23 @@ ggplot(data = test,
   geom_density(alpha = 0.7) +
   theme_bw()
 
-         
+T1_data_optimistic$sufficiently_powered <-
+  ifelse(T1_data_optimistic$effect < T1_data_optimistic$delta_powered, 1, 0)
 
[…]

### 2021-01-25 Manuscript/MS_Nature_Methods.Rmd
first draft Nat Met MS

---
title: 'MS Nature Methods'
csl: nature-methods.csl
indent: yes
output: bookdown::pdf_document2
toc: false
linestretch: 1.5
fontsize: 11pt
header-includes:
       - \usepackage{subfig}
abstract: \singlespacing Brief Communications begin with a brief unreferenced abstract (3 sentences, no more than 70 words),
bibliography: refs_repro.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# options(tinytex.verbose = TRUE)
```

```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(bookdown)
library(knitr)
library(tidyverse)
```

```{r data, include = FALSE, cache = FALSE}
load("all_outcomes_init_samp_size_10.RData")
load("exploratory_data_SESOI_Carneiro_10.RData")
load("exploratory_data_SESOI_Szucs_10.RData")
load("exploratory_data_significance_Carneiro_10.RData")
load("exploratory_data_significance_Szucs_10.RData")
```

**Main**

\ \ \ \ Preclinical research is essential for identifying promising interventions and to generate robust evidence to support translation to humans. Experiments in a preclinical setting are conducted in two different operating modes [@kimmelman2014distinguishing]. Early-stage preclinical experiments are *exploratory* with the aim to discover potentially effective interventions and generate hypotheses. These are tested at a later stage under more strict conditions in *confirmatory* mode [@landis2012call].

Various publications have called for a distinction between exploration and confirmation in preclinical animal research [@dirnagl2016thomas; @kimmelman2014distinguishing; @landis2012call]. Mogil & Macleod [@mogil2017no] took up this sentiment and proposed that an initial claim must be independently confirmed in order to be published. This suggestion emphasizes confirmation of initial findings. However, the question remains which exploratory results should be contested during confirmation. There is currently no guidance regarding criteria that should be applied to make this decision.

Commonly, the *p*-value is used to make a decision whether or not to further investigate a claim. Given that prior probabilities at this stage of discovery are low and sample sizes are small [@macleod2008evidence; @howells2014bringing], the *p*-value might prematurely discard potentially promising interventions [@dirnagl2020resolving].

Exploration requires sensitive tests to detect rare and possibly small effects [@bonapersona2019repair; @dirnagl2020resolving]. As more sensitive criteria invite more false positive results, confirmation must aim at reducing false positives to ensure that only true effects are carried forward to clinical testing. To increase power and safeguard reliability of results, sample sizes must be increased when switching from exploratory to confirmatory mode. However, ethical, time, and budget constraints prevent preclinical researchers from increasing their sample size and thus complicate the detection and precise estimation of effects. In an effort to prevent false negative results in exploration and reduce false positives during confirmation, it is necessary to devise strategies to move from exploration to confirmation that meet these complementary goals.

Here, we consider how to move from exploratory to confirmatory mode while balancing the number of animals against the likelihood of false negative and false positive outcomes.

To this end, we simulated two preclinical research trajectories comprising an exploratory study and a first confirmatory study (Figure 1?).

We based our simulations on two empirical effect size distributions reflecting an *optimistic* (high pre-study odds) and a *pessimistic* (low pre-study odds) scenario. After an initial exploratory study, a first decision identified experiments that should move from exploratory to confirmatory mode. One trajectory (standard) employed the conventional significance threshold ($\alpha$ = .05) for this decision. The second trajectory (SESOI) used a more lenient threshold based on an *a priori* determined smallest effect size of interest (SESOI).

In the standard trajectory, in the optimistic scenario, `r outcomes_all_distributions_init_samp_size_10$rep_attempts[8] / 100` percent of experiments met the criterion *p* $\leq$ .05. In the pessimistic scenario, `r outcomes_all_distributions_init_samp_size_10$rep_attempts[4] / 100` percent of experiments had a *p*-value $\leq$ .05. A closer look at the effect sizes of the experiments that proceeded to replication reveals that the conventional significance threshold is a conservative filter. Many effect sizes smaller than 1 but larger than 0 are eliminated and are not further investigated (Figure \@ref(fig:histograms)a--b). This demonstrates that the conventional significance threshold is not useful to screen for potentially meaningful effects at the early stages of drug discovery even if pre-study odds are high.

In the SESOI trajectory, we estimated the exploratory effect size and 95 percent confidence interval (CI) around that estimate. We examined whether the CI covered our SESOI (0.5 and 1.0, respectively). If this was the case for an experiment, it advanced to confirmatory mode. Importantly, we did not consider the *p*-value additionally. Applying this decision criterion resulted in `r outcomes_all_distributions_init_samp_size_10$rep_attempts[6] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[5] / 100` percent of experiments moving to confirmation in case of the optimistic distribution and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[2] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[1] / 100` percent for the pessimistic distribution based on an SESOI of 0.5 and 1.0, respectively. Compared to the conventional significance threshold, the range of effect size estimates that proceeded to confirmation shifted and included less extreme values, as well as more values closer to zero (Figure \@ref(fig:histograms)c--f). This decision criterion meets the goal of keeping the false negative rate low better than the conventional significance threshold.

```{r histograms, echo = FALSE, fig.cap = 'Samples (n = 10000) drawn from the two empirical effect size distributions (left: optimistic, right: pessimistic). The panels show the underlying effect sizes that were detected using one of the two decision criteria. Orange shaded bars (panels (a) and (b)) indicate those effect sizes that were identified for replication using the conventional significance threshold (p = .05). Blue shaded bars indicate effect sizes that were selected using a SESOI of 0.5 (panels (c) and (d)) or 1.0 (panels (e) and (f)), respectively. Note that in (a), (c), and (e) 16 values $>$ 10 were removed in order to display the distribution.', fig.subcap = c('', ''), out.width = '.50\\linewidth', fig.ncol = 2, fig.align = 'center'}

exploratory_data_sig_Szucs_10 <-
  exploratory_data_sig_Szucs_10 %>% 
  filter(pval_threshold == 0.05) %>% 
  filter(ES_true <= 10)

ES_histrogram_Szucs_selected_sig <-
  ggplot(data = exploratory_data_sig_Szucs_10, 
       aes(x = ES_true, fill = factor(selection_sig))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Cohen's ", italic("d"))),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#E69F00")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Szucs_selected_sig

exploratory_data_sig_Carneiro_10 <-
  exploratory_data_sig_Carneiro_10 %>% 
  filter(pval_threshold == 0.05)

ES_histrogram_Carneiro_selected_sig <-
  ggplot(data = exploratory_data_sig_Carneiro_10, 
       aes(x = ES_true, fill = factor(selection_sig))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Cohen's ", italic("d"))),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#E69F00")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Carneiro_selected_sig

exploratory_data_Szucs_10_0.5 <-
  exploratory_data_Szucs_10 %>% 
  filter(SESOI == 0.5) %>% 
  filter(ES_true <= 10)

ES_histrogram_Szucs_selected_SESOI_0.5 <-
  ggplot(data = exploratory_data_Szucs_10_0.5, 
       aes(x = ES_true, fill = factor(selection_equiv))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Cohen's ", italic("d"))),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Szucs_selected_SESOI_0.5

exploratory_data_Carneiro_10_0.5 <-
  exploratory_data_Carneiro_10 %>% 
  filter(SESOI == 0.5)

ES_histrogram_Carneiro_selected_SESOI_0.5 <-
  ggplot(data = exploratory_data_Carneiro_10_0.5, 
       aes(x = ES_true, fill = factor(selection_equiv))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Cohen's ", italic("d"))),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Carneiro_selected_SESOI_0.5

exploratory_data_Szucs_10_1.0 <-
  exploratory_data_Szucs_10 %>% 
  filter(SESOI == "1.0") %>% 
  filter(ES_true <= 10)

ES_histrogram_Szucs_selected_SESOI_1.0 <-
  ggplot(data = exploratory_data_Szucs_10_1.0, 
       aes(x = ES_true, fill = factor(selection_equiv))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Cohen's ", italic("d"))),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Szucs_selected_SESOI_1.0

exploratory_data_Carneiro_10_1.0 <-
  exploratory_data_Carneiro_10 %>% 
  filter(SESOI == "1.0")

ES_histrogram_Carneiro_selected_SESOI_1.0 <-
  ggplot(data = exploratory_data_Carneiro_10_1.0, 
       aes(x = ES_true, fill = factor(selection_equiv))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Cohen's ", italic("d"))),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Carneiro_selected_SESOI_1.0
```

```{r ES data, include = FALSE, cache = FALSE}
load("ES_data_Szucs.RData")
load("ES_data_Carneiro.RData")
```

```{r ES distribution, echo = FALSE, results = "hide"}
min(ES_data_Szucs$D)
max(ES_data_Szucs$D)

sum(ES_data_Szucs$D > 20)

median(ES_data_Szucs$D)
mean(ES_data_Szucs$D)

names(ES_data_Carneiro)[1] <- "D"

ES_data_Carneiro$D <- ifelse(ES_data_Carneiro$D < 0, -ES_data_Carneiro$D, -ES_data_Carneiro$D)

min(ES_data_Carneiro$D)
max(ES_data_Carneiro$D)
median(ES_data_Carneiro$D)
mean(ES_data_Carneiro$D)

ES_data_Carneiro <-
  ES_data_Carneiro %>% 
  mutate(distribution = "Pessimistic")

ES_Szucs <-
  ES_data_Szucs %>% 
  select(D) %>% 
  filter(D <= 20) %>% 
  mutate(distribution = "Optimistic")

ES_data <- bind_rows(ES_Szucs, ES_data_Carneiro)
```

In a second step, we calculated the sample size for a first confirmatory study. In the standard trajectory this was done via a standard power analysis using the initial exploratory effect size. The SESOI trajectory used again a pre-defined smallest effect size of interest (SESOI). Note that all sample sizes reported are the number of animals needed in *each* group (control and intervention).

In the standard trajectory, this resulted in a mean number of `r round(outcomes_all_distributions_init_samp_size_10$mean_N[7], 2)` (SD = `r round(outcomes_all_distributions_init_samp_size_10$mean_N[7] - outcomes_all_distributions_init_samp_size_10$mean_N_min[7], 2)`) animals in the optimistic, and `r round(outcomes_all_distributions_init_samp_size_10$mean_N[3], 2) ` (SD = `r round(outcomes_all_distributions_init_samp_size_10$mean_N[3] - outcomes_all_distributions_init_samp_size_10$mean_N_min[3], 2)`) in the pessimistic scenario. These small numbers reflect the large effect sizes that passed on to confirmation and were the basis for sample size calculation (Supplementary Figure 2a--b).

In the SESOI trajectory, the number of animals varied with the SESOI that was chosen. For an SESOI of 1.0, `r round(outcomes_all_distributions_init_samp_size_10$mean_N[1])` animals were needed in the replication in both the optimistic and pessimistic scenario. If the SESOI was 0.5, animal numbers increased to `r round(outcomes_all_distributions_init_samp_size_10$mean_N[2])` (Figure \@ref(fig:animalnumbers)). Note that the sample sizes reported are the number of animals needed in *each* group (control and intervention).

We further calculated the positive predictive value (PPV), false positive rate (FPR), and false negative rate (FNR) across both trajectories. The positive predictive value (PPV) of a study is the post-study probability that a positive finding which is based on statistical significance reflects a true effect [@ioannidis2005most]. The PPV is calculated from the pre-study odds, as well as the sensitivity and specificity of the test. In our study, pre-study odds of an effect of a given size (0.5 and 1.0, respectively) were determined by the empirical effect size distributions. If evidence for an initial claim is strengthened throughout the preclinical research trajectory, we would observe an increased PPV compared to pre-study odds.
In the optimistic scenario, the pre-study odds were `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[5], 2)` for SESOI of 0.5 and 1.0, respectively. In the pessimistic scenario pre-study odds were `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[1], 2)`, respectively. 
Across the standard trajectory, the PPV drops below pre-study odds in both scenarios (Figure \@ref(fig:PPV)). After the within-lab replication, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[8], 2)` and  `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[7], 2)` in the optimistic scenario for SESOI of 0.5 and 1.0, respectively. In the pessimistic scenario, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[4], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[3], 2)`.
In the SESOI trajectory, employing a SESOI at both stages along the decision-making process elevates the PPV above pre-study odds. Given a SESOI of 0.5 and 1.0, respectively, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[5], 2)` in the optimistic scenario, and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[1], 2)` in the pessimistic scenario.
Across the standard trajectory, given the "optimistic" scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[8], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[7], 2)` for SESOI of 0.5 and 1.0, respectively. In the "pessimistic" scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[4], 3)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[3], 2)` for SESOI of 0.5 and 1.0, respectively.
Across the SESOI trajectory, the FPR increased to `r round(outcomes_all_distributions_init_samp_size_10$FPR[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[5], 2)` in the "optimistic" scenario for SESOI of 0.5 and 1.0. Given the "pessimistic" scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[1], 2)` for SESOI set to 0.5 and 1.0. 
Across the standard trajectory, given the "optimistic" scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[8], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[7], 2)` for SESOI set to 0.5 and 1.0, respectively.  In the "pessimistic" scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[4], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[3], 2)` for SESOI of 0.5 and 1.0, respectively.
Across the SESOI trajectory, the FNR decreased to `r round(outcomes_all_distributions_init_samp_size_10$FNR[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[5], 2)` in the "optimistic" scenario for SESOI set to 0.5 and 1.0. Given the "pessimistic" scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[1], 2)` for SESOI of 0.5 and 1.0

```{r, echo = FALSE}

outcomes_all_distributions_init_samp_size_10$traject_SESOI <- 
  interaction(outcomes_all_distributions_init_samp_size_10$trajectory, outcomes_all_distributions_init_samp_size_10$SESOI)

plot_data_N <-
  outcomes_all_distributions_init_samp_size_10 %>% 
  filter(init_sample_size == 10) %>% 
  filter(traject_SESOI != "T1.1") %>% 
  group_by(distribution, SESOI)

plot_data <-
  outcomes_all_distributions_init_samp_size_10 %>% 
  filter(init_sample_size == 10) %>% 
  group_by(distribution, SESOI)
```

```{r results, echo = FALSE, message = FALSE, fig.cap = '(a) Number of animals needed in the first confirmatory study. In the standard trajectory, sample sizes are low, as they are based on large exploratory effect sizes. Error bars represent standard deviations. In case of trajectories using a SESOI, the number of animals is fixed. (b) Positive predictive value across trajectory. Dashed lines indicate pre-study odds based on empirical effect size distributions.', fig.subcap = c('', ''), out.width = '.50\\linewidth', fig.ncol = 2, fig.align = 'center'}

plot_mean_N <-
  ggplot(data = plot_data_N,
         aes(x = factor(traject_SESOI), y = mean_N, fill = factor(trajectory))) +
  geom_bar(stat = "identity", position = "dodge",
           size = .3,
           color = "black",
           alpha = 0.8) +
  geom_errorbar(aes(ymax = mean_N_max, ymin = mean_N_min), width = 0.1,
                position = position_dodge(width = 0.9)) +
  facet_grid(~ distribution) +
  # ggtitle("Mean number of animals for each trajectory") +
  labs(x = " ", y = "Mean # of animals in replication",
       fill = "Trajectory") +
  scale_x_discrete(labels = c("Standard", "SESOI = 0.5", "SESOI = 1.0")) +
  scale_fill_manual(labels = c("Standard", "SESOI"), values = c("#E69F00", "#0072B2")) + 
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 14, colour = "black",
                                   angle = 45, hjust = 1)) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  # theme(axis.title.x = element_text(size = 14)) +
  # theme(axis.title.y = element_text(size = 14)) +
  # theme(axis.text.x = element_text(size = 12, colour = "black")) +
  # theme(axis.text.y = element_text(size = 12, colour = "black")) +
  theme(strip.text.x = element_text(size = 14, colour = "black")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.title = element_text(size = 13, face = "bold")) +
  theme(legend.text = element_text(size = 12)) +
  theme(legend.position = "none")
  # theme(title = element_text(size = 15))

plot_mean_N

facet_names <- 
  c("0.5" = "SESOI = 0.5",
    "1" = "SESOI = 1.0")

plot_PPV <-
  ggplot(data = plot_data, 
         aes(x = factor(trajectory), 
             y = PPV_pop_prev,
             color = trajectory)) +
  geom_point(size = 3) +
  facet_grid(SESOI ~ distribution, labeller = labeller(.rows = facet_names)) +
  labs(x = "Trajectory", y = "Positive predictive value") +
  scale_x_discrete(labels = c("Standard", "SESOI")) +
  scale_color_manual(breaks = c("T1", "T2"),
                    values = c("#E69F00", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(strip.text.x = element_text(size = 14, colour = "black")) +
  theme(strip.text.y = element_text(size = 14, colour = "black")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.position = "none")

hlines <- data.frame(pre_study_odds = c(plot_data$Prevalence[1], plot_data$Prevalence[2],
                                        plot_data$Prevalence[5], plot_data$Prevalence[6]), 
                     distribution   = c(rep(plot_data$distribution[1], 2), 
                                        rep(plot_data$distribution[5], 2)),
                     SESOI          = rep(c("1", "0.5")))

plot_PPV <- 
  plot_PPV + 
  geom_hline(data = hlines, 
             aes(yintercept = pre_study_odds),
             color = "black", lty = 2, size = .5)

plot_PPV

facet_names <- c(
  "0.5" = "SESOI = 0.5",
  "1" = "SESOI = 1.0")

plot_FPR <-
  ggplot(data = plot_data, 
         aes(x = factor(trajectory), 
             y = FPR,
             color = trajectory)) +
  geom_point(size = 3, alpha = .8) +
  facet_grid(SESOI ~ distribution, labeller = labeller(.rows = facet_names)) +
  labs(x = "Trajectory", y = "False positive rate") +
  scale_x_discrete(labels = c("Standard", "SESOI")) +
  scale_color_manual(breaks = c("T1", "T2"),
                    values = c("#E69F00", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(strip.text.x = element_text(size = 14, colour = "black")) +
  theme(strip.text.y = element_text(size = 14, colour = "black")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.position = "none")

plot_FPR

facet_names <- c(
  "0.5" = "SESOI = 0.5",
  "1" = "SESOI = 1.0")

plot_FNR <-
  ggplot(data = plot_data, 
         aes(x = factor(trajectory), 
             y = FNR,
             color = trajectory)) +
  geom_point(size = 3, alpha = .8) +
  facet_grid(SESOI ~ distribution, labeller = labeller(.rows = facet_names)) +
  labs(x = "Trajectory", y = "False negative rate") +
  scale_x_discrete(labels = c("Standard", "SESOI")) +
  scale_color_manual(breaks = c("T1", "T2"),
                    values = c("#E69F00", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(strip.text.x = element_text(size = 14, colour = "black")) +
  theme(strip.text.y = element_text(size = 14, colour = "black")) +
  theme(strip.background = element_rect(fill = "white", color = "black")) +
  theme(legend.position = "none")

plot_FNR
```

\newpage

**Methods**

\scriptsize

\ \ \ \ **Simulation.**We explored different approaches to perform preclinical animal experiments via simulations. To this end, we modeled a simplified preclinical research trajectory from the exploratory stage to the results of a within-lab replication study (Figure \@ref(fig:trajectory)). Along the trajectory, there are different ways to increase the probability of not missing potentially meaningful effects. After an initial exploratory study, a first decision identifies experiments for replication. In our simulation, we employed two different decision criteria that indicate when one should move from the exploratory to confirmatory mode. If a decision has been made to replicate an initial study, we applied two approaches to determine the sample size for a replication study (smallest effect size of interest (SESOI) and standard power analysis), as outlined in detail below.

\ \ \ \ We explored different approaches to perform preclinical animal experiments via simulations. To this end, we modeled a simplified preclinical research trajectory from the exploratory stage to the results of a within-lab replication study (Figure \@ref(fig:trajectory) \textcolor{red}{make new figure and include two trajectories in MS and all 4 in Supplement}). Along the trajectory, there are different ways to increase the probability of not missing potentially meaningful effects. After an initial exploratory study, a first decision identifies experiments for replication. In our simulation, we employed two different decision criteria that indicate when one should move from the exploratory stage to the replication [if you use this lingo, it should match the one in the intro] stage. If a decision has been made to replicate an initial study, we applied two approaches to determine the sample size for a replication study (smallest effect size of interest (SESOI) and standard power analysis), as outlined in detail below.

```{r, echo = FALSE, results = FALSE}
library(png)
library(knitr)
img1_path <- "trajectory_proposal.png"
img1 <- readPNG(img1_path, native = TRUE, info = TRUE)
attr(img1, "info")
```

```{r trajectory, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "A preclinical research trajectory from the exploratory stage to a within-lab replication. The four panels along the arrow display four possible combinations of decision criteria and approaches to calculate the sample size for replication that can be employed throughout the trajectory.", fig.align = 'center'}
include_graphics(img1_path)
```

\ \ \ \ *Empirical effect size distributions.*Simulations were based on empirical effect size distributions from the recently published literature  [@szucs2017empirical; @carneiro2018effect]. This enabled us to determine the prior probability (pre-study odds) of a certain alternative hypothesis ($H_{1}$) which we defined as an effect of a given size (e.g. a Cohen's *d* of 0.5). The two distributions reflect different research fields.

The distribution of effect sizes extracted from Szucs & Ioannidis (2017) [@szucs2017empirical] contains 26841 effect sizes from the cognitive neuroscience and psychology literature published between January 2011 and August 2014. All effect sizes are calculated as the standardized difference in means (Cohen's *d*). Effect size estimates range from `r round(min(ES_data_Szucs$D), 2)` to `r round(max(ES_data_Szucs$D), 2)`, and have a median of `r round(median(ES_data_Szucs$D), 2)`. As the pre-study odds of a medium effect of 0.5 are rather large (`r round(outcomes_all_distributions_init_samp_size_10$prev_pop[6], 2)`), we will refer to this distribution as "optimistic". We acknowledge that the effect sizes were mainly extracted from human studies. However, in large parts the distribution is in agreement with effect sizes reported to be typical of (some areas of) preclinical research (find good ref here \textcolor{red}{[This could be infectious diseases, usually antibiotics are all or nothing effects]}).

As our study is concerned specifically with preclinical research, we chose a second distribution of empirical effect sizes to represent one field of the preclinical realm. Carneiro et al.'s (2018) [@carneiro2018effect] study systematically examined effect sizes in the rodent fear conditioning literature. Effect sizes were extracted from 410 experiments published in PubMed in 2013. The publication included a data file containing all extracted effect sizes. After removing missing values, the data set consisted of 336 effect sizes, again, calculated as Cohen's *d*. The effect sizes range from `r round(min(ES_data_Carneiro$D), 2)` to `r round(max(ES_data_Carneiro$D), 2)`, and have a median of `r round(median(ES_data_Carneiro$D), 2)`. The prior probability of observing an effect of 0.5 is `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[2], 2)`. We will therefore refer to this distribution as "pessimistic". [in the limitations section you could cite the Bioarxiv paper and compare the two distributions to support your choice and that it may not be overly pessimistic. Particular in fields where there was no progress like neurodegenerative diseases]

\ \ \ \ *Exploratory stage.*From each of the two distributions, we drew 10000 samples of effect sizes from which we created 10000 study data sets. Each data set comprised data of two groups consisting of ten experimental units each drawn from a normal distribution. We chose a number of ten EUs based on reported sample sizes in preclinical studies [@howells2014bringing]. [For the limitation section we need to argue that different initial sample sizes like 7 or 15 will not change results dramatically]. Our simulated design mimics a comparison between two groups where one receives an intervention and the other functions as a control group. The study data sets are compared using a two-sided two-sample *t*-test. From these exploratory study results, we extracted the *p*-values and 95 percent confidence intervals (CI). We then employed two different criteria based on the *p*-value or 95 percent CI, respectively, to decide whether to continue to a replication.

\ \ \ \ *Decision criteria to proceed to replication.*The first decision criterion employs the conventional significance threshold  ($\alpha$ = .05) to decide whether to replicate an exploratory study. If a *p*-value extracted from a two-sided two-sample *t*-test is $\leq$ .05, this study will proceed to the replication stage. If not, the trajectory is terminated after the exploratory study. We chose this decision criterion as our reference, as this is what we consider to be current practice.

As an alternative to this approach, we propose to set a smallest effect size of interest (SESOI) and examine whether the 95 percent CI around the exploratory effect size estimate covers this SESOI. A SESOI is the effect size that the researcher based their knowledge of the literature in their respective field (domain knowledge) and given practical constraints considers biologically and clinically meaningful [@lakens2018equivalence]. In our simulation, we used 0.5 and 1.0 as SESOI. This approach emphasizes the importance of effect sizes rather than statistical significance to evaluate an intervention's effect. Further, we expected this approach to be more lenient than statistical significance (at least if the significance threshold was set at $\alpha$ = .05) and to allow a broader range of effect sizes to pass on to be further investigated.

\ \ \ \ *Approaches to determine sample size for replication.* Once the decision to continue to replication has been made, we employed two different approaches to determine the sample size for the replication study. After having conducted an exploratory study, we have an estimate of the direction of the effect. Only effect sizes that showed an effect that favors the treatment over the control group were considered for further investigation. Thus, for the replication study, a one-sided two-sample *t*-test was performed. The desired power level for replication was set to .80, $\alpha$ was set to .05. In order to calculate the sample size for replication given power and $\alpha$, an effect size estimate is required. In one approach, we used the exploratory effect size estimate to compute the replication sample size. In an alternative approach, we employed the same SESOI used as decision criterion earlier. In statistical terms, our SESOI was set such that the replication study would have a power of .50 to detect an effect of this size (# explain in more detail and find good ref to motivate this: Lakens? \textcolor{red}{[This is mainly to ensure that the likelihood of a type I error below this threshold is negligible. The goal is to reduce Type I error in this second phase.]}). Consequently, in the first approach, the replication sample size was dependent on the outcome of the exploratory study, whereas using a SESOI always yielded the same sample size regardless of the exploratory effect size (e.g. 23 EUs in each group for a SESOI of 0.5).

\ \ \ \ *Replication stage.* For each of the studies that met the decision criterion after the exploratory study (either *p* $\leq$ .05 or SESOI within the 95 percent CI of the exploratory effect size estimate), a replication study was performed. The number of replication studies conducted varied with the decision criterion used and, in case of the criterion employing a SESOI, also with the SESOI (0.5 and 1.0). A replication study was performed as a one-sided two-sample *t*-test, where the number of animals in each group was determined by the approach to calculate the sample size. For a replication be considered "successful", the *p*-value had to be below the conventional significance threshold ($\alpha$ = .05).

\ \ \ \ *Trajectories.* We compared the two trajectories (Standard and SESOI) regarding the number of experiments proceeding to the replication stage, number of animals needed in the replication, and positive predictive value across the trajectory. Secondary outcomes are the false positive rate, false negative rate, and effect size precision. Outcome variables are outlined in more detail in the following section. The standard trajectory constitutes our reference, as we consider it to be closest to current practice. We have stored data, results, and figures of all four trajectories in an online repository (insert URL here).

\newpage

**References**
\ \ \ \ 
\singlespacing

### 2021-01-25 scripts/analysis/analysis_effect_size_precision.R

@@ -160,7 +160,7 @@ ggplot(data = T1_data,
   geom_violin(position = position_dodge(width = 0.9)) +
   geom_rect(fill = "lightgrey", alpha = .01, 
             aes(xmin = -Inf, xmax = Inf, 
-                ymin = powered_d_50, ymax = Inf)) +
+                ymin = powered_d_80, ymax = Inf)) +
   # geom_rect(fill = "lightpink", alpha = .01, 
   #           aes(xmin = -Inf, xmax = Inf, 
   #               ymin = powered_d_50, ymax = powered_d_80)) +
@@ -182,7 +182,6 @@ for(i in 1:nrow(T1_data)) {
                alternative = "greater")$d
   
 }
-  
 
 T1_data$delta_powered <- delta_powered
 
@@ -236,7 +235,6 @@ T1_data_optimistic$sufficiently_powered <-
 
 sum(T1_data_optimistic$sufficiently_powered == 1) / nrow(T1_data_optimistic)
   
-    
 T1_data_pessimistic$sufficiently_powered <-
   ifelse(T1_data_pessimistic$effect < T1_data_pessimistic$delta_powered, 1, 0)
 

### 2021-01-25 scripts/analysis/analysis_exploratory_stage_Carneiro.R

@@ -1,0 +1,100 @@
+library(tidyverse)
+
+setwd("~/Documents/SimulateTranslation/")
+
+
+extractor <- function(k) {
+  
+  test_data <- read_csv(paste("./data/Carneiro_distribution/Frequentist_analysis/exploratory_stage_equiv_", k, sep = ""))
+}
+
+sets_equiv <- c("0.1", "0.3", "0.5", "0.7", "1.0")
+
+exploratory_data <- list()
+
+for (i in sets_equiv) {
+  
+  exploratory_data[[i]] <- extractor(i)
+  print(i)
+  
+}
+
+
+sum(exploratory_data$`0.1`$selection_equiv == 1)
+sum(exploratory_data$`0.3`$selection_equiv == 1)
+sum(exploratory_data$`0.5`$selection_equiv == 1)
+sum(exploratory_data$`0.7`$selection_equiv == 1)
+sum(exploratory_data$`1.0`$selection_equiv == 1)
+
+
+exploratory_data <-
+  bind_rows(exploratory_data$`0.1`,
+            exploratory_data$`0.3`,
+            exploratory_data$`0.5`,
+            exploratory_data$`0.7`,
+            exploratory_data$`1.0`)
+
+
+exploratory_data$SESOI <-
+  rep(sets_equiv, each = 30000)
+
+exploratory_data_Carneiro_10 <-
+  exploratory_data %>% 
+  filter(init_sample_size == 10) %>% 
+  filter(SESOI == "0.5") %>% 
+  filter(selection_equiv == 1) %>% 
+  filter(effect <= 0)
+
+# save(exploratory_data_Carneiro_10, file = "./data/exploratory_data_SESOI_Carneiro_10.RData")
+
+ggplot(data = exploratory_data_Carneiro_10,
+       aes(x = ES_true, 
+           color = factor(selection_equiv), 
+           fill = factor(selection_equiv))) +
+  geom_density(alpha = 0.2) +
+  facet_wrap(~ SESOI, nrow = 1, ncol = 5) +
+  theme_bw()
+  # geom_vline(xintercept = SESOI, color = "red", lty = 2)
+
+extractor2 <- function(k) {
+  
+  test_data_sig <- read_csv(paste("./data/Carneiro_distribution/Frequentist_analysis/exploratory_stage_sig_", k, sep = ""))
+}
+
+sets_sig <- c("0.05", "0.1")
+
+exploratory_data_sig <- list()
+
[…]

### 2021-01-25 scripts/analysis/analysis_exploratory_stage_Szucs.R

@@ -1,0 +1,102 @@
+library(tidyverse)
+
+setwd("~/Documents/SimulateTranslation/")
+
+
+extractor <- function(k) {
+  
+  test_data <- read_csv(paste("./data/Szucs_distribution/Frequentist_analysis/exploratory_stage_equiv_", k, sep = ""))
+}
+
+sets_equiv <- c("0.1", "0.3", "0.5", "0.7", "1.0")
+
+exploratory_data <- list()
+
+for (i in sets_equiv) {
+  
+  exploratory_data[[i]] <- extractor(i)
+  print(i)
+  
+}
+
+sum(exploratory_data$`0.1`$selection_equiv == 1)
+sum(exploratory_data$`0.3`$selection_equiv == 1)
+sum(exploratory_data$`0.5`$selection_equiv == 1)
+sum(exploratory_data$`0.7`$selection_equiv == 1)
+sum(exploratory_data$`1.0`$selection_equiv == 1)
+
+
+exploratory_data <-
+  bind_rows(exploratory_data$`0.1`,
+            exploratory_data$`0.3`,
+            exploratory_data$`0.5`,
+            exploratory_data$`0.7`,
+            exploratory_data$`1.0`)
+
+
+exploratory_data$SESOI <-
+  rep(sets_equiv, each = 30000)
+
+exploratory_data_Szucs_10 <-
+  exploratory_data %>% 
+  filter(init_sample_size == 10) %>% 
+  filter(SESOI == "1.0") %>% 
+  filter(selection_equiv == 1) %>%
+  filter(effect <= 0)
+
+# save(exploratory_data_Szucs_10, file = "./data/exploratory_data_SESOI_Szucs_10.RData")
+
+
+ggplot(data = exploratory_data_Szucs_10,
+       aes(x = ES_true, 
+           color = factor(selection_equiv), 
+           fill = factor(selection_equiv))) +
+  geom_density(alpha = 0.2) +
+  facet_wrap(~ SESOI, nrow = 1, ncol = 5) +
+  theme_bw()
+# geom_vline(xintercept = SESOI, color = "red", lty = 2)
+
+
+
+extractor2 <- function(k) {
+  
+  test_data_sig <- read_csv(paste("./data/Szucs_distribution/Frequentist_analysis/exploratory_stage_sig_", k, sep = ""))
+}
+
+sets_sig <- c("0.05", "0.1")
+
+exploratory_data_sig <- list()
+
+for (i in sets_sig) {
+  
[…]

### 2021-01-25 scripts/analysis/analysis_simulation_results_across_trajectory.R

@@ -3,16 +3,16 @@ setwd("~/Documents/QUEST/PhD/R/SimulateTranslation/")
 rm(list = ls())
 library(tidyverse)
 
-# source("./scripts/analysis/prior_odds_for_analysis_Szucs.R")
+source("./scripts/analysis/prior_odds_for_analysis_Szucs.R")
 
-source("./scripts/analysis/prior_odds_for_analysis_Carneiro.R")
+# source("./scripts/analysis/prior_odds_for_analysis_Carneiro.R")
 
 ### read in data sets from the different trajectories 
 ### using fixed-N design in replication study
 
-final <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_1.0")
+# final <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_1.0")
 
-# final <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1")
+final <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1")
 
 
 hist(final$effect, breaks = 100)
@@ -21,10 +21,10 @@ hist(final$ES_true, breaks = 100)
 
 
 ### add column that codes decision criterion from exploratory stage to confirmatory stage
-final$decision_crit <- "equivalence"
+final$decision_crit <- "significance"
 
 ### add column that codes approach to determine sample size for confirmatory study
-final$sampsize_approach <- "SESOI"
+final$sampsize_approach <- "standard"
 
 ### add column for outcome significant / not significant to match outcome column
 ### of sequential design data set
@@ -68,7 +68,7 @@ true_positives <-
            decision_crit, sampsize_approach, H0, 
            all_positives, all_negatives, 
            prev_pop, rep_attempts) %>%
-  filter(H0 == 2 & ES_true > 1) %>%
+  filter(H0 == 2 & ES_true > 0) %>%
   summarize(true_pos = n())
 
 false_negatives <-
@@ -77,7 +77,7 @@ false_negatives <-
            decision_crit, sampsize_approach, H0, 
            all_positives, all_negatives, 
            prev_pop, rep_attempts) %>%
-  filter(H0 == 1 & ES_true > 1) %>%
+  filter(H0 == 1 & ES_true > 0) %>%
[…]

### 2021-01-25 scripts/simulation/sim_fixN_Carneiro_decision_equivalence_with_SESOI.R

@@ -1,4 +1,4 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+setwd("~/Documents/SimulateTranslation")
 
 rm(list = ls())
 
@@ -53,11 +53,11 @@ sum(current_ES < 0)
 sum(current_ES == 0)
 
 #how many hypothesis over SESOI threshold
-SESOI       <- c(.5, 1)
+SESOI <- c(.1, .3, .5, .7, 1)
 
 mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
               dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
-                              c(.5, 1)))
+                              c(.1, .3, .5, .7, 1)))
 
 prev_pop      <- vector()
 all_positives <- vector()
@@ -134,7 +134,7 @@ selection_equiv <- list()
 for (i in 1:length(samp_size_vector)) {
   
   selection_equiv[[i]] <- future_map(exploratory_data_summary[[i]], get_decision_equiv,
-                                     SESOI = 0.5)
+                                     SESOI = 1.0)
   
 }
 
@@ -158,10 +158,36 @@ df_equiv <- as_tibble(matrix(unlist(selection_equiv),
 
 dat <- bind_cols(df, df_equiv)
 
-dat$ES_true <- current_ES
+dat$ES_true <- rep(current_ES, 3)
 
+# write.csv(dat, file = "./data/Carneiro_distribution/Frequentist_analysis/exploratory_stage_equiv_1.0")
 
 hist(dat$ES_true, breaks = 100)
+
+dat <-
+  dat %>%
+  filter(init_sample_size == 10) %>% 
+  filter(ES_true > 0)
+
+dat_selected <-
+  dat %>%
+  filter(selection_equiv == 1)
+
+dat_not_selected <-
+  dat %>% 
+  filter(selection_equiv == 0)
+
+false_positives <-
+  dat_selected %>% 
+  filter(selection_equiv == 1 & ES_true < 0.5)
+
+false_negatives <-
+  dat_not_selected %>% 
+  filter(selection_equiv == 0 & ES_true >= 0.5)
+
+FPR <- nrow(false_positives) / nrow(dat_selected)
+
+FNR <- nrow(false_negatives) / nrow(dat_not_selected)
 
 # dat <-
 #   dat %>%

### 2021-01-25 scripts/simulation/sim_fixN_Carneiro_decision_significance.R

@@ -47,16 +47,17 @@ current_ES <- sample(ES_true, n_exp, replace = TRUE)
 hist(current_ES, breaks = 200)
 median(current_ES)
 mean(current_ES)
+
 sum(current_ES > 0)
 sum(current_ES < 0)
 sum(current_ES == 0)
 
 #how many hypothesis over SESOI threshold
-SESOI <- c(.5, 1)
+SESOI <- c(.1, .3, .5, .7, 1)
 
 mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
               dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
-                              c(.5, 1)))
+                              c(.1, .3, .5, .7, 1)))
 
 prev_pop      <- vector()
 all_positives <- vector()
@@ -133,7 +134,9 @@ selection_sig <- list()
 for (i in 1:length(samp_size_vector)) {
   
   selection_sig[[i]] <- 
-    future_map(exploratory_data_summary[[i]], get_decision_sig, pval_threshold = 0.05)
+    future_map(exploratory_data_summary[[i]],
+               get_decision_sig,
+               pval_threshold = 0.05)
   
 }
 
@@ -155,49 +158,96 @@ df_sig <- as_tibble(matrix(unlist(selection_sig),
                                            c(col_name))))
 
 
-
 dat <- bind_cols(df, df_sig)
 
-dat$ES_true <- current_ES
+dat$ES_true <- rep(current_ES, 3)
+# 
+# # write.csv(dat, file = "./data/Carneiro_distribution/Frequentist_analysis/exploratory_stage_sig_0.05")
+# 
+# 
+# dat <-
+#   dat %>%
+#   group_by(init_sample_size)
+#   # filter(init_sample_size == 10) %>% 
+#   # filter(ES_true > 0)
+# 
+# dat$effect <- ifelse(dat$effect < 0, -dat$effect, -dat$effect)
+# 
+# ggplot(data = dat,
+#        aes(x = ES_true, y = effect, color = factor(selection_sig))) +
+#   geom_point(alpha = 0.4, size = 0.5) +
+#   facet_wrap(~ init_sample_size) +
+#   theme_bw()
+# 
+# dat_selected <-
+#   dat %>%
+#   filter(selection_sig == 1)
+# 
+# dat_not_selected <-
+#   dat %>% 
+#   filter(selection_sig == 0)
+# 
+# false_positives <-
[…]

### 2021-01-25 scripts/simulation/sim_fixN_Szucs_decision_equivalence_with_SESOI.R

@@ -1,4 +1,4 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+setwd("~/Documents/SimulateTranslation")
 
 rm(list = ls())
 
@@ -34,11 +34,11 @@ hist(ES_true, breaks = 200)
 hist(current_ES, breaks = 200)
 
 #how many hypothesis over SESOI threshold
-SESOI       <- c(.5, 1)
+SESOI <- c(.1, .3, .5, .7, 1)
 
 mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
               dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
-                              c(.5, 1)))
+                              c(.1, .3, .5, .7, 1)))
 
 prev_pop      <- vector()
 all_positives <- vector()
@@ -103,7 +103,6 @@ for (i in 1:length(samp_size_vector)) {
   
 }
 
-
 # decision to go on
 # this decision depends on whether our SESOI is within the 95 % CI
 # select studies for replication if SESOI lies within the 95 % CI
@@ -115,7 +114,6 @@ for (i in 1:length(samp_size_vector)) {
                                      SESOI = 1.0)
   
 }
-
 
 row_names <- NULL
 col_names <- c("init_sample_size", "study_id", "t_value",
@@ -136,7 +134,8 @@ df_equiv <- as_tibble(matrix(unlist(selection_equiv),
 
 dat <- bind_cols(df, df_equiv)
 
-dat$ES_true <- current_ES
+dat$ES_true <- rep(current_ES, 3)
 
+write.csv(dat, file = "./data/Szucs_distribution/Frequentist_analysis/exploratory_stage_equiv_1.0")
 
 hist(dat$ES_true, breaks = 100)

### 2021-01-25 scripts/simulation/sim_fixN_Szucs_decision_significance.R

@@ -1,4 +1,4 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+setwd("~/Documents/SimulateTranslation")
 
 rm(list = ls())
 
@@ -25,7 +25,7 @@ library(doParallel)
 registerDoParallel()
 getDoParWorkers()
 
-n_exp <- 3
+n_exp <- 10000
 ES_true <- ES_data_Szucs$D
 
 set.seed(4321)
@@ -33,12 +33,14 @@ current_ES <- sample(ES_true, n_exp)
 hist(ES_true, breaks = 200)
 hist(current_ES, breaks = 200)
 
+max(current_ES)
+
 #how many hypothesis over SESOI threshold
-SESOI       <- c(.5, 1)
+SESOI <- c(.1, .3, .5, .7, 1)
 
 mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
               dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
-                              c(.5, 1)))
+                              c(.1, .3, .5, .7, 1)))
 
 prev_pop      <- vector()
 all_positives <- vector()
@@ -111,7 +113,9 @@ selection_sig <- list()
 for (i in 1:length(samp_size_vector)) {
   
   selection_sig[[i]] <- 
-    future_map(exploratory_data_summary[[i]], get_decision_sig, pval_threshold = 0.05)
+    future_map(exploratory_data_summary[[i]],
+               get_decision_sig,
+               pval_threshold = 0.1)
   
 }
 
@@ -136,8 +140,8 @@ df_sig <- as_tibble(matrix(unlist(selection_sig),
 
 dat <- bind_cols(df, df_sig)
 
-dat$ES_true <- current_ES
+dat$ES_true <- rep(current_ES, 3)
 
-# write.csv(dat, file = "./data/Szucs_distribution_only_exploratory_stage")
+# write.csv(dat, file = "./data/Szucs_distribution/Frequentist_analysis/exploratory_stage_sig_0.1")
 
 

### 2021-01-26 Manuscript/MS_Nature_Methods.Rmd
changes in Nat Met MS

@@
 
-We based our simulations on two empirical effect size distributions reflecting an *optimistic* (high pre-study odds) and a *pessimistic* (low pre-study odds) scenario.
+```{r experiments proceeding, echo = FALSE, results = "hide"}
+
+exploratory_data_sig_Carneiro_10_0.05 <-
+  exploratory_data_sig_Carneiro_10 %>% 
+  filter(pval_threshold == 0.05) %>% 
+  filter(selection_sig == 1) %>% 
+  #filter(effect < 0) %>% 
+  summarise(N = n())
+
+exploratory_data_sig_Carneiro_10_0.05$N / 100
+
+exploratory_data_Carneiro_10_0.5 <-
+  exploratory_data_Carneiro_10 %>% 
+  filter(SESOI == 0.5) %>% 
+  filter(selection_equiv == 1) %>%
+  # filter(effect < 0) %>% 
+  summarise(N = n())
+
+exploratory_data_Carneiro_10_0.5$N / 100
+
+exploratory_data_Carneiro_10_1.0 <-
+  exploratory_data_Carneiro_10 %>% 
+  filter(SESOI == "1.0") %>% 
+  filter(selection_equiv == 1) %>% 
+  summarise(N = n())
+
+exploratory_data_Carneiro_10_1.0$N / 100
+
+exploratory_data_sig_Szucs_10_0.05 <-
+  exploratory_data_sig_Szucs_10 %>% 
+  filter(pval_threshold == 0.05) %>% 
+  filter(selection_sig == 1) %>% 
+  # filter(effect < 0) %>%
+  summarise(N = n())
+
+exploratory_data_sig_Szucs_10_0.05$N / 100
+
+exploratory_data_Szucs_10_0.5 <-
+  exploratory_data_Szucs_10 %>% 
+  filter(SESOI == 0.5) %>% 
+  filter(selection_equiv == 1) %>%
+  # filter(effect < 0) %>% 
+  summarise(N = n())
+
+exploratory_data_Szucs_10_0.5$N / 100
+
+exploratory_data_Szucs_10_1.0 <-
+  exploratory_data_Szucs_10 %>% 
+  filter(SESOI == "1.0") %>% 
+  filter(selection_equiv == 1) %>% 
+  summarise(N = n())
+
+exploratory_data_Szucs_10_1.0$N / 100
+```
+
+We based our simulations on two empirical effect size distributions reflecting an *optimistic* (high pre-study odds) and a *pessimistic* (low pre-study odds) scenario (Supplementary Figure 1).
 After an initial exploratory study, a first decision identified experiments that should move from exploratory to confirmatory mode.
@@
 
-In the standard trajectory, in the optimistic scenario, `r outcomes_all_distributions_init_samp_size_10$rep_attempts[8] / 100` percent of experiments met the criterion *p* $\leq$ .05.
-In the pessimistic scenario, `r outcomes_all_distributions_init_samp_size_10$rep_attempts[4] / 100` percent of experiments had a *p*-value $\leq$ .05.
-A closer look at the effect sizes of the experiments that proceeded to replication reveals that the conventional significance threshold is a conservative filter.
-Many effect sizes smaller than 1 but larger than 0 are eliminated and are not further investigated (Figure \@ref(fig:histograms)a--b).
+In the standard trajectory, in the optimistic scenario, `r exploratory_data_sig_Szucs_10_0.05$N / 100` percent of experiments met the criterion *p* $\leq$ .05.
+In the pessimistic scenario, `r exploratory_data_sig_Carneiro_10_0.05$N / 100` percent of experiments had a *p*-value $\leq$ .05.
+A closer look at the effect sizes of the experiments that proceeded to confirmation reveals that the conventional significance threshold is a conservative filter.
+Many effect sizes $\neq$ 0 are eliminated and not further investigated (Figure \@ref(fig:histograms)a--b).
 This demonstrates that the conventional significance threshold is not useful to screen for potentially meaningful effects at the early stages of drug discovery even if pre-study odds are high.
@@
 In the SESOI trajectory, we estimated the exploratory effect size and 95 percent confidence interval (CI) around that estimate.
-We examined whether the CI covered our SESOI (0.5 and 1.0, respectively).
+We examined whether the CI covered our SESOI (Cohen's *d* of 0.5 and 1.0, respectively).
 If this was the case for an experiment, it advanced to confirmatory mode.
 Importantly, we did not consider the *p*-value additionally.
-Applying this decision criterion resulted in `r outcomes_all_distributions_init_samp_size_10$rep_attempts[6] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[5] / 100` percent of experiments moving to confirmation in case of the optimistic distribution and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[2] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[1] / 100` percent for the pessimistic distribution based on an SESOI of 0.5 and 1.0, respectively.
+Applying this decision criterion resulted in `r exploratory_data_Szucs_10_0.5$N / 100` and `r exploratory_data_Szucs_10_1.0$N / 100` percent of experiments moving to confirmation in case of the optimistic distribution and `r exploratory_data_Carneiro_10_0.5$N / 100` and `r exploratory_data_Carneiro_10_1.0$N / 100` percent for the pessimistic distribution based on a SESOI of 0.5 and 1.0, respectively.
 Compared to the conventional significance threshold, the range of effect size estimates that proceeded to confirmation shifted and included less extreme values, as well as more values closer to zero (Figure \@ref(fig:histograms)c--f).
-This decision criterion meets the goal of keeping the false negative rate low better than the conventional significance threshold.
+This decision criterion meets the goal of preventing false negatives better than the conventional significance threshold.
 
@@
 
+From the experiments that were identified for confirmation only those that showed exploratory effect sizes in favor of the treatment were further investigated.
 In a second step, we calculated the sample size for a first confirmatory study.
@@
 The SESOI trajectory used again a pre-defined smallest effect size of interest (SESOI).
-Note that all sample sizes reported are the number of animals needed in *each* group (control and intervention).
 
-In the standard trajectory, this resulted in a mean number of `r round(outcomes_all_distributions_init_samp_size_10$mean_N[7], 2)` (SD = `r round(outcomes_all_distributions_init_samp_size_10$mean_N[7] - outcomes_all_distributions_init_samp_size_10$mean_N_min[7], 2)`) animals in the optimistic, and `r round(outcomes_all_distributions_init_samp_size_10$mean_N[3], 2) ` (SD = `r round(outcomes_all_distributions_init_samp_size_10$mean_N[3] - outcomes_all_distributions_init_samp_size_10$mean_N_min[3], 2)`) in the pessimistic scenario.
+In the standard trajectory, this resulted in a mean number of `r round(outcomes_all_distributions_init_samp_size_10$mean_N[7], 2)` (SD = `r round(outcomes_all_distributions_init_samp_size_10$mean_N[7] - outcomes_all_distributions_init_samp_size_10$mean_N_min[7], 2)`) animals in the optimistic, and `r round(outcomes_all_distributions_init_samp_size_10$mean_N[3], 2) ` (SD = `r round(outcomes_all_distributions_init_samp_size_10$mean_N[3] - outcomes_all_distributions_init_samp_size_10$mean_N_min[3], 2)`) in the pessimistic scenario (Figure \@ref(fig:results)a).
 These small numbers reflect the large effect sizes that passed on to confirmation and were the basis for sample size calculation (Supplementary Figure 2a--b).
@@
 In the SESOI trajectory, the number of animals varied with the SESOI that was chosen.
-For an SESOI of 1.0, `r round(outcomes_all_distributions_init_samp_size_10$mean_N[1])` animals were needed in the replication in both the optimistic and pessimistic scenario.
-If the SESOI was 0.5, animal numbers increased to `r round(outcomes_all_distributions_init_samp_size_10$mean_N[2])` (Figure \@ref(fig:animalnumbers)).
-Note that the sample sizes reported are the number of animals needed in *each* group (control and intervention).
+For an SESOI of 1.0, `r round(outcomes_all_distributions_init_samp_size_10$mean_N[1])` animals were needed in the confirmatory study in both the optimistic and pessimistic scenario.
+If the SESOI was 0.5, animal numbers increased to `r round(outcomes_all_distributions_init_samp_size_10$mean_N[2])` (Figure \@ref(fig:results)a).
 
@@
 If evidence for an initial claim is strengthened throughout the preclinical research trajectory, we would observe an increased PPV compared to pre-study odds.
+
 In the optimistic scenario, the pre-study odds were `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[5], 2)` for SESOI of 0.5 and 1.0, respectively.
 In the pessimistic scenario pre-study odds were `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[1], 2)`, respectively.
-Across the standard trajectory, the PPV drops below pre-study odds in both scenarios (Figure \@ref(fig:PPV)).
+
+Across the standard trajectory, the PPV drops below pre-study odds in both scenarios (Figure \@ref(fig:results)b).
 After the within-lab replication, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[8], 2)` and  `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[7], 2)` in the optimistic scenario for SESOI of 0.5 and 1.0, respectively.
 In the pessimistic scenario, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[4], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[3], 2)`.
-In the SESOI trajectory, employing a SESOI at both stages along the decision-making process elevates the PPV above pre-study odds.
+
+In the SESOI trajectory, employing a SESOI at both stages along the decision-making process elevates the PPV above pre-study odds (Figure \@ref(fig:results)b).
 Given a SESOI of 0.5 and 1.0, respectively, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[5], 2)` in the optimistic scenario, and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[1], 2)` in the pessimistic scenario.
-Across the standard trajectory, given the "optimistic" scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[8], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[7], 2)` for SESOI of 0.5 and 1.0, respectively.
-In the "pessimistic" scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[4], 3)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[3], 2)` for SESOI of 0.5 and 1.0, respectively.
-Across the SESOI trajectory, the FPR increased to `r round(outcomes_all_distributions_init_samp_size_10$FPR[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[5], 2)` in the "optimistic" scenario for SESOI of 0.5 and 1.0.
-Given the "pessimistic" scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[1], 2)` for SESOI set to 0.5 and 1.0.
-Across the standard trajectory, given the "optimistic" scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[8], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[7], 2)` for SESOI set to 0.5 and 1.0, respectively.
-In the "pessimistic" scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[4], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[3], 2)` for SESOI of 0.5 and 1.0, respectively.
-Across the SESOI trajectory, the FNR decreased to `r round(outcomes_all_distributions_init_samp_size_10$FNR[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[5], 2)` in the "optimistic" scenario for SESOI set to 0.5 and 1.0.
-Given the "pessimistic" scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[1], 2)` for SESOI of 0.5 and 1.0
+
+Across the standard trajectory, given the optimistic scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[8], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[7], 2)` for SESOI of 0.5 and 1.0, respectively.
+In the pessimistic scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[4], 3)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[3], 2)` for SESOI of 0.5 and 1.0, respectively (Figure \@ref(fig:results)c).
+
+Across the SESOI trajectory, the FPR increased to `r round(outcomes_all_distributions_init_samp_size_10$FPR[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[5], 2)` in the optimistic scenario for SESOI of 0.5 and 1.0.
+Given the pessimistic scenario, the FPR was `r round(outcomes_all_distributions_init_samp_size_10$FPR[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FPR[1], 2)` for SESOI set to 0.5 and 1.0 (Figure \@ref(fig:results)c).
+
+Across the standard trajectory, given the optimistic scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[8], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[7], 2)` for SESOI set to 0.5 and 1.0, respectively.
+In the pessimistic scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[4], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[3], 2)` for SESOI of 0.5 and 1.0, respectively (Figure \@ref(fig:results)d).
+
+Across the SESOI trajectory, the FNR decreased to `r round(outcomes_all_distributions_init_samp_size_10$FNR[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[5], 2)` in the optimistic scenario for SESOI set to 0.5 and 1.0.
+Given the pessimistic scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[1], 2)` for SESOI of 0.5 and 1.0 (Figure \@ref(fig:results)d).
+
+Overall, our simulation shows that current practice reflected by the standard trajectory does not meet the complementary goals of exploration and confirmation along the preclinical research trajectory.
+In the standard trajectory, the switch from exploratory to confirmatory mode is extremely costly, as it eliminates numerous potentially meaningful effects.
+This shortcoming is also reflected in the PPV where even in the best-case scenario only 40 out 100 significant results reflect an underlying true effect (compared to 73/100 in the SESOI trajectory).
+In light of our findings, we advise a rejection of the standard approach.
+We present here an easily applicable alternative to current practice.
+Optimizing decision criteria and sample size calculations by employing a SESOI increases chances to detect true effects.
 
@@
 In case of trajectories using a SESOI, the number of animals is fixed.
-(b) Positive predictive value across trajectory.
+Note that all sample sizes displayed and reported in the text are the number of animals needed in *each* group (control and intervention).(b) Positive predictive value across trajectory.
 Dashed lines indicate pre-study odds based on empirical effect size distributions.', fig.subcap = c('', ''), out.width = '.50\\linewidth', fig.ncol = 2, fig.align = 'center'}
@@
 
-\ \ \ \ **Simulation.**We explored different approaches to perform preclinical animal experiments via simulations.
-To this end, we modeled a simplified preclinical research trajectory from the exploratory stage to the results of a within-lab replication study (Figure \@ref(fig:trajectory)).
+\ \ \ \ **Simulation.** We explored different approaches to perform preclinical animal experiments via simulations.
+To this end, we modeled a simplified preclinical research trajectory from the exploratory stage to the results of a first confirmatory study (within-lab replication; Figure \@ref(fig:trajectory)) (\textcolor{red}{make new figure and include two trajectories in MS and all 4 in Supplement}).
 Along the trajectory, there are different ways to increase the probability of not missing potentially meaningful effects.
@@
 
-\ \ \ \ We explored different approaches to perform preclinical animal experiments via simulations.
-To this end, we modeled a simplified preclinical research trajectory from the exploratory stage to the results of a within-lab replication study (Figure \@ref(fig:trajectory) \textcolor{red}{make new figure and include two trajectories in MS and all 4 in Supplement}).
-Along the trajectory, there are different ways to increase the probability of not missing potentially meaningful effects.
-After an initial exploratory study, a first decision identifies experiments for replication.
-In our simulation, we employed two different decision criteria that indicate when one should move from the exploratory stage to the replication [if you use this lingo, it should match the one in the intro] stage.
-If a decision has been made to replicate an initial study, we applied two approaches to determine the sample size for a replication study (smallest effect size of interest (SESOI) and standard power analysis), as outlined in detail below.
-
-```{r, echo = FALSE, results = FALSE}
-library(png)
-library(knitr)
-img1_path <- "trajectory_proposal.png"
-img1 <- readPNG(img1_path, native = TRUE, info = TRUE)
-attr(img1, "info")
-```
-
-```{r trajectory, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "A preclinical research trajectory from the exploratory stage to a within-lab replication.
-The four panels along the arrow display four possible combinations of decision criteria and approaches to calculate the sample size for replication that can be employed throughout the trajectory.", fig.align = 'center'}
-include_graphics(img1_path)
-```
-
-\ \ \ \ *Empirical effect size distributions.*Simulations were based on empirical effect size distributions from the recently published literature  [@szucs2017empirical; @carneiro2018effect].
+\ \ \ \ *Empirical effect size distributions.* Simulations were based on empirical effect size distributions from the recently published literature  [@szucs2017empirical; @carneiro2018effect].
 This enabled us to determine the prior probability (pre-study odds) of a certain alternative hypothesis ($H_{1}$) which we defined as an effect of a given size (e.g.
 a Cohen's *d* of 0.5).
-The two distributions reflect different research fields.
 
@@
 Effect size estimates range from `r round(min(ES_data_Szucs$D), 2)` to `r round(max(ES_data_Szucs$D), 2)`, and have a median of `r round(median(ES_data_Szucs$D), 2)`.
-As the pre-study odds of a medium effect of 0.5 are rather large (`r round(outcomes_all_distributions_init_samp_size_10$prev_pop[6], 2)`), we will refer to this distribution as "optimistic".
-We acknowledge that the effect sizes were mainly extracted from human studies.
-However, in large parts the distribution is in agreement with effect sizes reported to be typical of (some areas of) preclinical research (find good ref here \textcolor{red}{[This could be infectious diseases, usually antibiotics are all or nothing effects]}).
+As the pre-study odds of a medium effect of 0.5 are rather large (`r round(outcomes_all_distributions_init_samp_size_10$prev_pop[6], 2)`), we will refer to this distribution as *optimistic*.
 
-As our study is concerned specifically with preclinical research, we chose a second distribution of empirical effect sizes to represent one field of the preclinical realm.
-Carneiro et al.'s (2018) [@carneiro2018effect] study systematically examined effect sizes in the rodent fear conditioning literature.
+Our second empirical distribution was taken from Carneiro et al.'s (2018) [@carneiro2018effect] study which systematically examined effect sizes in the rodent fear conditioning literature.
 Effect sizes were extracted from 410 experiments published in PubMed in 2013.
@@
 The prior probability of observing an effect of 0.5 is `r round(outcomes_all_distributions_init_samp_size_10$prev_pop[2], 2)`.
-We will therefore refer to this distribution as "pessimistic".
-[in the limitations section you could cite the Bioarxiv paper and compare the two distributions to support your choice and that it may not be overly pessimistic.
-Particular in fields where there was no progress like neurodegenerative diseases]
+We will therefore refer to this distribution as *pessimistic*.
 
-\ \ \ \ *Exploratory stage.*From each of the two distributions, we drew 10000 samples of effect sizes from which we created 10000 study data sets.
+\ \ \ \ *Exploratory mode.* From each of the two distributions, we drew 10000 samples of effect sizes from which we created 10000 study data sets.
 Each data set comprised data of two groups consisting of ten experimental units each drawn from a normal distribution.
 We chose a number of ten EUs based on reported sample sizes in preclinical studies [@howells2014bringing].
-[For the limitation section we need to argue that different initial sample sizes like 7 or 15 will not change results dramatically].
 Our simulated design mimics a comparison between two groups where one receives an intervention and the other functions as a control group.
@@
 From these exploratory study results, we extracted the *p*-values and 95 percent confidence intervals (CI).
-We then employed two different criteria based on the *p*-value or 95 percent CI, respectively, to decide whether to continue to a replication.
+We then employed two different criteria based on the *p*-value or 95 percent CI, respectively, to decide whether to continue to a confirmatory mode.
 
-\ \ \ \ *Decision criteria to proceed to replication.*The first decision criterion employs the conventional significance threshold  ($\alpha$ = .05) to decide whether to replicate an exploratory study.
-If a *p*-value extracted from a two-sided two-sample *t*-test is $\leq$ .05, this study will proceed to the replication stage.
+\ \ \ \ *Decision criteria to proceed to replication.* The first decision criterion employs the conventional significance threshold  ($\alpha$ = .05) to decide whether to replicate an exploratory study.
+If a *p*-value extracted from a two-sided two-sample *t*-test is $\leq$ .05, this study will proceed to confirmation.
 If not, the trajectory is terminated after the exploratory study.
@@
 As an alternative to this approach, we propose to set a smallest effect size of interest (SESOI) and examine whether the 95 percent CI around the exploratory effect size estimate covers this SESOI.
-A SESOI is the effect size that the researcher based their knowledge of the literature in their respective field (domain knowledge) and given practical constraints considers biologically and clinically meaningful [@lakens2018equivalence].
+A SESOI is the effect size that the researcher based their domain knowledge and given practical constraints considers biologically and clinically meaningful [@lakens2018equivalence].
 In our simulation, we used 0.5 and 1.0 as SESOI.
@@
 
-\ \ \ \ *Approaches to determine sample size for replication.* Once the decision to continue to replication has been made, we employed two different approaches to determine the sample size for the replication study.
-After having conducted an exploratory study, we have an estimate of the direction of the effect.
-Only effect sizes that showed an effect that favors the treatment over the control group were considered for further investigation.
-Thus, for the replication study, a one-sided two-sample *t*-test was performed.
-The desired power level for replication was set to .80, $\alpha$ was set to .05.
-In order to calculate the sample size for replication given power and $\alpha$, an effect size estimate is required.
-In one approach, we used the exploratory effect size estimate to compute the replication sample size.
-In an alternative approach, we employed the same SESOI used as decision criterion earlier.
-In statistical terms, our SESOI was set such that the replication study would have a power of .50 to detect an effect of this size (# explain in more detail and find good ref to motivate this: Lakens?
-\textcolor{red}{[This is mainly to ensure that the likelihood of a type I error below this threshold is negligible.
-The goal is to reduce Type I error in this second phase.]}).
-Consequently, in the first approach, the replication sample size was dependent on the outcome of the exploratory study, whereas using a SESOI always yielded the same sample size regardless of the exploratory effect size (e.g.
-23 EUs in each group for a SESOI of 0.5).
+\ \ \ \ *Approaches to determine sample size for replication.* Once the decision to continue to confirmatory mode has been made, we employed two different approaches to determine the sample size for the first confirmatory study.
+After the exploratory study, we have an estimate of the direction of the effect.
+Only effect sizes that showed an effect in favor of the treatment were considered for further investigation.
+Thus, for the confirmatory study, a one-sided two-sample *t*-test was performed.
+In the standard trajectory, the desired power level for replication was set to .80, $\alpha$ was set to .05.
+To calculate the sample size given power and $\alpha$, we used the exploratory effect size estimate.
+In the SESOI trajectory, we employed the same SESOI used as decision criterion earlier.
+Our SESOI was set such that the confirmatory study would have a power of .50 to detect an effect of this size.
+This power level was chosen to ensure that the likelihood of a false positive finding below the threshold determined by our SESOI is negligible.
+The aim during confirmation is to weed out false positives.
 
-\ \ \ \ *Replication stage.* For each of the studies that met the decision criterion after the exploratory study (either *p* $\leq$ .05 or SESOI within the 95 percent CI of the exploratory effect size estimate), a replication study was performed.
-The number of replication studies conducted varied with the decision criterion used and, in case of the criterion employing a SESOI, also with the SESOI (0.5 and 1.0).
-A replication study was performed as a one-sided two-sample *t*-test, where the number of animals in each group was determined by the approach to calculate the sample size.
-For a replication be considered "successful", the *p*-value had to be below the conventional significance threshold ($\alpha$ = .05).
+\ \ \ \ *Confirmatory mode.* For each of the studies that met the decision criterion after the exploratory study (either *p* $\leq$ .05 or SESOI within the 95 percent CI of the exploratory effect size estimate), a confirmatory study was performed.
+The number of studies conducted varied with the decision criterion used and, in case of the criterion employing a SESOI, also with the SESOI (0.5 and 1.0).
+A confirmatory study was performed as a one-sided two-sample *t*-test, where the number of animals in each group was determined by the approach to calculate the sample size.
+For a confirmation to be considered "successful", the *p*-value had to be below the conventional significance threshold ($\alpha$ = .05).
 
-\ \ \ \ *Trajectories.* We compared the two trajectories (Standard and SESOI) regarding the number of experiments proceeding to the replication stage, number of animals needed in the replication, and positive predictive value across the trajectory.
-Secondary outcomes are the false positive rate, false negative rate, and effect size precision.
-Outcome variables are outlined in more detail in the following section.
-The standard trajectory constitutes our reference, as we consider it to be closest to current practice.
-We have stored data, results, and figures of all four trajectories in an online repository (insert URL here).
+\ \ \ \ *Outcome variables* We compared the two trajectories (standard and SESOI) regarding the number of experiments proceeding to confirmatory mode, number of animals needed in the confirmatory study, and positive predictive value (PPV), false positive rate (FPR), and false negative rate (FNR) across the trajectory.
 

### 2021-01-26 scripts/analysis/analysis_simulation_results_across_trajectory.R

@@ -1,4 +1,4 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation/")
+setwd("~/Documents/SimulateTranslation/")
 
 rm(list = ls())
 library(tidyverse)
@@ -10,9 +10,9 @@ source("./scripts/analysis/prior_odds_for_analysis_Szucs.R")
 ### read in data sets from the different trajectories 
 ### using fixed-N design in replication study
 
-# final <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_1.0")
+final <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_0.1")
 
-final <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1")
+# final <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1")
 
 
 hist(final$effect, breaks = 100)

### 2021-01-26 scripts/simulation/sim_fixN_Carneiro_decision_equivalence_with_SESOI.R

@@ -134,7 +134,7 @@ selection_equiv <- list()
 for (i in 1:length(samp_size_vector)) {
   
   selection_equiv[[i]] <- future_map(exploratory_data_summary[[i]], get_decision_equiv,
-                                     SESOI = 1.0)
+                                     SESOI = 0.7)
   
 }
 
@@ -163,96 +163,3 @@ dat$ES_true <- rep(current_ES, 3)
 # write.csv(dat, file = "./data/Carneiro_distribution/Frequentist_analysis/exploratory_stage_equiv_1.0")
 
 hist(dat$ES_true, breaks = 100)
-
-dat <-
-  dat %>%
-  filter(init_sample_size == 10) %>% 
-  filter(ES_true > 0)
-
-dat_selected <-
-  dat %>%
-  filter(selection_equiv == 1)
-
-dat_not_selected <-
-  dat %>% 
-  filter(selection_equiv == 0)
-
-false_positives <-
-  dat_selected %>% 
-  filter(selection_equiv == 1 & ES_true < 0.5)
-
-false_negatives <-
-  dat_not_selected %>% 
-  filter(selection_equiv == 0 & ES_true >= 0.5)
-
-FPR <- nrow(false_positives) / nrow(dat_selected)
-
-FNR <- nrow(false_negatives) / nrow(dat_not_selected)
-
-# dat <-
-#   dat %>%
-#   filter(init_sample_size == 10)
-# 
-# # dat$effect <- ifelse(dat$effect < 0, -dat$effect, -dat$effect)
-# 
-# dat_large_ES <-
-#   dat %>%
-#   filter(ES_true > 0)
-# 
-# dat_selected <-
-#   dat_large_ES %>% 
-#   filter(selection_equiv == 1)
-# 
-# 6618/6844*100
-# 
-# ggplot(data = dat_large_ES, aes(x = dat_large_ES$ES_true, fill = factor(selection_equiv))) +
-#   geom_histogram(bins = 50, color = "black", size = 0.3, alpha = 0.8) +
-#   # geom_histogram(data = dat_large_ES, aes(x = ),
-#                  # bins = 50, color = "black", fill = "white")
-#   labs(x = expression(paste("Sampled effect sizes (Cohen's ", italic("d"), ") ")),
-#        y = "Frequency",
-#        fill = "Selected for \nreplication") +
-#   scale_fill_manual(breaks = c("0", "1"),
-#                     labels = c("no",
-#                                "yes"),
[…]

### 2021-01-26 scripts/simulation/sim_fixN_Carneiro_decision_significance.R

@@ -136,7 +136,7 @@ for (i in 1:length(samp_size_vector)) {
   selection_sig[[i]] <- 
     future_map(exploratory_data_summary[[i]],
                get_decision_sig,
-               pval_threshold = 0.05)
+               pval_threshold = 0.1)
   
 }
 
@@ -161,128 +161,7 @@ df_sig <- as_tibble(matrix(unlist(selection_sig),
 dat <- bind_cols(df, df_sig)
 
 dat$ES_true <- rep(current_ES, 3)
-# 
-# # write.csv(dat, file = "./data/Carneiro_distribution/Frequentist_analysis/exploratory_stage_sig_0.05")
-# 
-# 
-# dat <-
-#   dat %>%
-#   group_by(init_sample_size)
-#   # filter(init_sample_size == 10) %>% 
-#   # filter(ES_true > 0)
-# 
-# dat$effect <- ifelse(dat$effect < 0, -dat$effect, -dat$effect)
-# 
-# ggplot(data = dat,
-#        aes(x = ES_true, y = effect, color = factor(selection_sig))) +
-#   geom_point(alpha = 0.4, size = 0.5) +
-#   facet_wrap(~ init_sample_size) +
-#   theme_bw()
-# 
-# dat_selected <-
-#   dat %>%
-#   filter(selection_sig == 1)
-# 
-# dat_not_selected <-
-#   dat %>% 
-#   filter(selection_sig == 0)
-# 
-# false_positives <-
-#   dat_selected %>% 
-#   filter(selection_sig == 1 & ES_true < 0)
-# 
-# ggplot(data = false_positives,
-#        aes(x = ES_true, y = effect)) +
-#   geom_point(alpha = 0.4, size = 0.5) +
-#   facet_wrap(~ init_sample_size) +
-#   theme_bw()
-# 
-# false_negatives <-
-#   dat_not_selected %>% 
-#   filter(selection_sig == 0 & ES_true > 0)
-# 
-# ggplot(data = false_negatives,
-#        aes(x = ES_true, y = effect)) +
-#   geom_point(alpha = 0.4, size = 0.5) +
-#   facet_wrap(~ init_sample_size) +
-#   theme_bw()
-# 
-# FPR <- nrow(false_positives) / nrow(dat_selected)
-# 
-# FNR <- nrow(false_negatives) / nrow(dat_not_selected)
 
[…]

### 2021-01-26 scripts/simulation/sim_fixN_Carneiro_replication.R

@@ -1,4 +1,4 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+setwd("~/Documents/SimulateTranslation")
 
 rm(list = ls())
 
@@ -7,13 +7,18 @@ rm(list = ls())
 
 # source("./scripts/simulation/sim_fixN_Carneiro_decision_equivalence_with_SESOI.R")
 
-source("./scripts/simulation/sim_fixN_Carneiro_decision_significance.R")
+# source("./scripts/simulation/sim_fixN_Carneiro_decision_significance.R")
 
 dat
 
 sum(dat$effect < 0) # empirical effect sizes are negative because t.test function takes control - treat
 sum(dat$effect > 0) 
 sum(dat$ES_true < 0)
+
+# data <-
+#   dat %>% 
+#   group_by(init_sample_size, study_id) %>% 
+#   filter(selection_equiv == 1)
 
 data <-
   dat %>% 
@@ -23,6 +28,11 @@ data <-
 sum(data$effect < 0)
 sum(data$effect > 0)
 sum(data$effect == 0)
+
+# selected <-
+#   data %>% 
+#   group_by(init_sample_size) %>% 
+#   summarize(selected = sum(selection_equiv == 1))
 
 selected <-
   data %>% 
@@ -40,6 +50,10 @@ for (i in 1:nrow(data)) {
   rep_sample_size_std[i] <-
     ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
                              method = 1))
+  
+  # rep_sample_size_std[i] <-
+  #   ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
+  #                            method = 2, SESOI = 0.5, power = .5))
 }
 
 data$rep_samp_size_std <- rep_sample_size_std
@@ -125,62 +139,19 @@ res_summary_rep <-
 hist(res_summary_rep$effect, breaks = 200)
 
 
+res_summary_rep$effect <- ifelse(res_summary_rep$effect < 0,
+                                 -res_summary_rep$effect, -res_summary_rep$effect)
 
-
-# res_summary_rep$effect <- ifelse(res_summary_rep$effect < 0,
-#                                  -res_summary_rep$effect, -res_summary_rep$effect)
-# 
 # write.csv(res_summary_rep,
-#           file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_1.0")
-# 
-# res_summary_rep <-
[…]

### 2021-01-26 scripts/simulation/sim_fixN_Szucs_decision_equivalence_with_SESOI.R

@@ -136,6 +136,6 @@ dat <- bind_cols(df, df_equiv)
 
 dat$ES_true <- rep(current_ES, 3)
 
-write.csv(dat, file = "./data/Szucs_distribution/Frequentist_analysis/exploratory_stage_equiv_1.0")
+# write.csv(dat, file = "./data/Szucs_distribution/Frequentist_analysis/exploratory_stage_equiv_1.0")
 
 hist(dat$ES_true, breaks = 100)

### 2021-01-26 scripts/simulation/sim_fixN_Szucs_decision_significance.R

@@ -106,8 +106,8 @@ for (i in 1:length(samp_size_vector)) {
 }
 
 # decision to go on
-# this decision depends on whether exploratory result is significant (p <= .05) or not
-# select studies for replication if p-value < .05
+# this decision depends on whether exploratory result is significant (p <= .05 / p <= .1) or not
+# select studies for replication if p-value < .05 or < .1
 selection_sig <- list()
 
 for (i in 1:length(samp_size_vector)) {
@@ -115,7 +115,7 @@ for (i in 1:length(samp_size_vector)) {
   selection_sig[[i]] <- 
     future_map(exploratory_data_summary[[i]],
                get_decision_sig,
-               pval_threshold = 0.1)
+               pval_threshold = 0.05)
   
 }
 

### 2021-01-26 scripts/simulation/sim_fixN_Szucs_replication.R

@@ -1,4 +1,4 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+setwd("~/Documents/SimulateTranslation")
 
 rm(list = ls())
 
@@ -15,30 +15,30 @@ sum(dat$effect < 0) # empirical effect sizes are negative because t.test functio
 sum(dat$effect > 0) 
 sum(dat$ES_true < 0)
 
+# data <-
+#   dat %>%
+#   group_by(init_sample_size, study_id) %>%
+#   filter(selection_sig == 1)
+
 data <-
   dat %>%
   group_by(init_sample_size, study_id) %>%
-  filter(selection_sig == 1)
-
-# data <-
-#   dat %>% 
-#   group_by(init_sample_size, study_id) %>% 
-#   filter(selection_equiv == 1)
+  filter(selection_equiv == 1)
 
 sum(data$effect < 0)
 sum(data$effect > 0)
 sum(data$effect == 0)
 
 
+# selected <-
+#   data %>%
+#   group_by(init_sample_size) %>%
+#   summarize(selected = sum(selection_sig == 1))
+
 selected <-
   data %>%
   group_by(init_sample_size) %>%
-  summarize(selected = sum(selection_sig == 1))
-
-# selected <-
-#   data %>% 
-#   group_by(init_sample_size) %>% 
-#   summarize(selected = sum(selection_equiv == 1))
+  summarize(selected = sum(selection_equiv == 1))
 
 
 # now estimate sample size for replication study
@@ -49,13 +49,13 @@ rep_sample_size_std <- NULL
 
 for (i in 1:nrow(data)) {
   
-  rep_sample_size_std[i] <-
-    ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-                             method = 1))
-  
   # rep_sample_size_std[i] <-
   #   ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-  #                            method = 2, SESOI = 0.5, power = .5))
+  #                            method = 1))
+  
+  rep_sample_size_std[i] <-
+    ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
+                             method = 2, SESOI = 1.0, power = .5))
 }
 
 data$rep_samp_size_std <- rep_sample_size_std
@@ -98,9 +98,9 @@ rep_attempts <-
 replication_data <- list()
 rep_exp_no <- 0
 
[…]

### 2021-02-03 Manuscript/MS_Nature_Methods.Rmd
changed figure Nat Met MS

@@
 ---
-title: 'MS Nature Methods'
+title: 'Increasing discovery rates in preclinical research through optimised statistical decision criteria'
 csl: nature-methods.csl
@@
        - \usepackage{subfig}
+         \usepackage{ulem}
 abstract: \singlespacing Brief Communications begin with a brief unreferenced abstract (3 sentences, no more than 70 words),
@@
 \ \ \ \ Preclinical research is essential for identifying promising interventions and to generate robust evidence to support translation to humans.
-Experiments in a preclinical setting are conducted in two different operating modes [@kimmelman2014distinguishing].
+To achieve these goals, experiments are conducted in two different operating modes [@kimmelman2014distinguishing].
 Early-stage preclinical experiments are *exploratory* with the aim to discover potentially effective interventions and generate hypotheses.
@@
 
-Various publications have called for a distinction between exploration and confirmation in preclinical animal research [@dirnagl2016thomas; @kimmelman2014distinguishing; @landis2012call].
-Mogil & Macleod [@mogil2017no] took up this sentiment and proposed that an initial claim must be independently confirmed in order to be published.
-This suggestion emphasizes confirmation of initial findings.
-However, the question remains which exploratory results should be contested during confirmation.
-There is currently no guidance regarding criteria that should be applied to make this decision.
+This sequential approach will increase the likelihood that generated evidence is robust and potentially reduce translational failure [@dirnagl2016thomas; @kimmelman2014distinguishing; @landis2012call;@mogil2017no][ADD elife paper here and cut landis and kimmelman].
+Before conducting confirmatory experiments, researchers need to decide which exploratory results should be contested.
+Commonly, such a decision is based on a statistical test and associated *p*-value threshold in an exploratory study [do we have a good citation here?].
+Here, we argue that this standard approach is detrimental to the goal to actually identify efficacious treatments.
+We propose an alternative that is, mimicking clinical trials, focused on a smallest effect size of interest.
+We show through simulations that this will increase the probability that a confirmatory study identifies a true effect, i.e.
+the positive predictive value.[perhaps cite Ionanidis 2005 here].
+[and will increaeGiven that prior probabilities at this stage of discovery are low and sample sizes are small [@macleod2008evidence; @howells2014bringing], the *p*-value might prematurely discard potentially promising interventions [@dirnagl2020resolving].]
 
-Commonly, the *p*-value is used to make a decision whether or not to further investigate a claim.
-Given that prior probabilities at this stage of discovery are low and sample sizes are small [@macleod2008evidence; @howells2014bringing], the *p*-value might prematurely discard potentially promising interventions [@dirnagl2020resolving].
+Our approach is based on the notion that exploration and confirmation fulfill different roles in scientific discovery.
+Exploration detects via sensitive tests potentially relevant effects among many hypotheses that have a low prior probability of being true (in the following \textit{pre-study odds}).
+[@bonapersona2019repair; @dirnagl2020resolving].
+As more sensitive criteria invite more false positive results, confirmation must aim at reducing false positives to ensure that only true effects are carried forward to subsequent testing.
+[To increase power and safeguard reliability of results, sample sizes must be increased when switching from exploratory to confirmatory mode.
+suggest delete] To complicate matters, ethical, time, and budget constraints limit degrees of freedom in experimental design.
+Consequently, to prevent false negative results in exploration and to reduce false positives during confirmation, it is necessary to devise strategies to move from exploration to confirmation that meet these complementary goals.
 
-Exploration requires sensitive tests to detect rare and possibly small effects [@bonapersona2019repair; @dirnagl2020resolving].
-As more sensitive criteria invite more false positive results, confirmation must aim at reducing false positives to ensure that only true effects are carried forward to clinical testing.
-To increase power and safeguard reliability of results, sample sizes must be increased when switching from exploratory to confirmatory mode.
-However, ethical, time, and budget constraints prevent preclinical researchers from increasing their sample size and thus complicate the detection and precise estimation of effects.
-In an effort to prevent false negative results in exploration and reduce false positives during confirmation, it is necessary to devise strategies to move from exploration to confirmation that meet these complementary goals.
+Specifically, the likelihood of false negative and false positive outcomes needs to be balanced against the number of animals. [Perhaps can give a small example here if space allows]
 
-Here, we consider how to move from exploratory to confirmatory mode while balancing the number of animals against the likelihood of false negative and false positive outcomes.
-
-To this end, we simulated two preclinical research trajectories comprising an exploratory study and a first confirmatory study (Figure 1?).
+To this end, we simulated two preclinical research trajectories each comprising an exploratory study and a first confirmatory study (Figure 1?).
 
@@
 
-We based our simulations on two empirical effect size distributions reflecting an *optimistic* (high pre-study odds) and a *pessimistic* (low pre-study odds) scenario (Supplementary Figure 1).
-After an initial exploratory study, a first decision identified experiments that should move from exploratory to confirmatory mode.
+We based our simulations on two published empirical effect size distributions.
+reflecting an *optimistic* (high pre-study odds) and a *pessimistic* (low pre-study odds) scenario (Supplementary Figure 1).
+After an initial exploratory study, a decision criterion identified experiments that should move from exploratory to confirmatory mode.
 One trajectory (standard) employed the conventional significance threshold ($\alpha$ = .05) for this decision.
-The second trajectory (SESOI) used a more lenient threshold based on an *a priori* determined smallest effect size of interest (SESOI).
+The second trajectory (SESOI) used a more lenient threshold based on an *a priori* determined smallest effect size of interest (SESOI), a practice also employed in clinical trials [CITATION].
+[copied from below] For this, we estimated the effect size of each exploratory study and associated 95 percent confidence interval (CI).
+We examined whether this CI covered our SESOI (Cohen's *d* of 0.5 and 1.0, respectively).
 
-In the standard trajectory, in the optimistic scenario, `r exploratory_data_sig_Szucs_10_0.05$N / 100` percent of experiments met the criterion *p* $\leq$ .05.
+\sout{In the standard trajectory, under optimistic pre-study odds, `r exploratory_data_sig_Szucs_10_0.05$N / 100` percent of experiments met the criterion *p* $\leq$ .05.
 In the pessimistic scenario, `r exploratory_data_sig_Carneiro_10_0.05$N / 100` percent of experiments had a *p*-value $\leq$ .05.
-A closer look at the effect sizes of the experiments that proceeded to confirmation reveals that the conventional significance threshold is a conservative filter.
+Effect sizes of experiments that proceeded to confirmation reveals that the conventional significance threshold is a conservative filter.
 Many effect sizes $\neq$ 0 are eliminated and not further investigated (Figure \@ref(fig:histograms)a--b).
+This demonstrates that the conventional significance threshold is not useful to screen for potentially meaningful effects at the early stages of drug discovery even if pre-study odds are high.}
+
+\emph{I rephrased the entire previous paragraph for readability.
+The main take home message in this rewrite is that you currently do not guide the reader through the results.
+You state the results but do not imply an interpretation.}
+
+As expected, more experiments transitioned to the confirmatory phase in the SESOI trajectory compared to the standard trajectory.
+This was true for both the optimistic and pessimistic pre-study odds (PUT STATS HERE).
+This was also reflected in effect sizes where many effect sizes $\neq$ 0 are eliminated and not further investigated (Figure \@ref(fig:histograms)a--b).
 This demonstrates that the conventional significance threshold is not useful to screen for potentially meaningful effects at the early stages of drug discovery even if pre-study odds are high.
+Contrary, the range of effect size estimates in the SESOI trajectory that proceeded to confirmation shifted[shifted where?]  and included less extreme values, as well as more values closer to zero (Figure \@ref(fig:histograms)c--f).
+This decision criterion thus reduces false negatives compared to the conventional significance threshold.
 
-In the SESOI trajectory, we estimated the exploratory effect size and 95 percent confidence interval (CI) around that estimate.
+\sout{In the SESOI trajectory, we estimated the exploratory effect size and 95 percent confidence interval (CI) around that estimate.
 We examined whether the CI covered our SESOI (Cohen's *d* of 0.5 and 1.0, respectively).
@@
 Compared to the conventional significance threshold, the range of effect size estimates that proceeded to confirmation shifted and included less extreme values, as well as more values closer to zero (Figure \@ref(fig:histograms)c--f).
-This decision criterion meets the goal of preventing false negatives better than the conventional significance threshold.
+This decision criterion meets the goal of preventing false negatives better than the conventional significance threshold.}
 
@@
 
+[Please adjust the following paragraphs in the same way I have adjusted the one above.
+State the result and give descriptive stats in brackets.
+As it is this is way too long and does not read smoothly.]
+
 From the experiments that were identified for confirmation only those that showed exploratory effect sizes in favor of the treatment were further investigated.
@@
 Given the pessimistic scenario, the FNR was `r round(outcomes_all_distributions_init_samp_size_10$FNR[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$FNR[1], 2)` for SESOI of 0.5 and 1.0 (Figure \@ref(fig:results)d).
-
-Overall, our simulation shows that current practice reflected by the standard trajectory does not meet the complementary goals of exploration and confirmation along the preclinical research trajectory.
-In the standard trajectory, the switch from exploratory to confirmatory mode is extremely costly, as it eliminates numerous potentially meaningful effects.
-This shortcoming is also reflected in the PPV where even in the best-case scenario only 40 out 100 significant results reflect an underlying true effect (compared to 73/100 in the SESOI trajectory).
-In light of our findings, we advise a rejection of the standard approach.
-We present here an easily applicable alternative to current practice.
-Optimizing decision criteria and sample size calculations by employing a SESOI increases chances to detect true effects.
 
@@
 
+Overall, our simulation shows that current practice reflected by the standard trajectory does not meet the complementary goals of exploration and confirmation along a preclinical research trajectory.
+In the standard trajectory, the switch from exploratory to confirmatory mode eliminates numerous potentially meaningful effects.
+This shortcoming is also reflected in the PPV where only 40 out 100 significant results reflect an underlying true effect (compared to 73/100 in the SESOI trajectory).
+In light of our findings, we advise a rejection of the standard approach.
+[After shortening the results part significantly here we can drop a few lines more on how to apply this.
+Adjust the SESOI to the feasibility in the lab.
+This will raise awareness for the limitations of the study!
+We could also hint at the number of initial animals and how these influence our results.
+We should also emphasise the deviation from one experiment one result practice but towards a systemic investigation of true effects.
+Also hint at the possibility for multi center trials and how much costs are saved downstream].We present here an easily applicable alternative to current practice.
+Optimizing decision criteria and sample size calculations by employing a SESOI increases chances to detect true effects while keeping the number of animals reasonably low.
+
 \newpage

### 2021-02-05 Manuscript/Discussion.Rmd
updated data files after R update

---
title: 'Identifying optimized decision criteria and experimental designs by simulating preclinical experiments in silico'
csl: biomed-central.csl
indent: yes
output: bookdown::pdf_document2
toc: false
linestretch: 1.5
fontsize: 12pt
header-includes:
       - \usepackage{subfig}
bibliography: refs_repro.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(bookdown)
library(knitr)
library(tidyverse)
```

```{r, echo = FALSE, results = FALSE}
library(png)
library(knitr)
img1_path <- "trajectory_proposal.png"
img1 <- readPNG(img1_path, native = TRUE, info = TRUE)
attr(img1, "info")
```

```{r trajectory, echo = FALSE, message = FALSE, out.width = "80%", fig.cap = "A preclinical research trajectory from the exploratory stage to a within-lab replication. The four panels along the arrow display four possible combinations of decision criteria and approaches to calculate the sample size for replication that can be employed throughout the trajectory.", fig.align = 'center'}
include_graphics(img1_path)
```

## For Supplement

Outcome variables are outlined in more detail in the following section. The standard trajectory constitutes our reference, as we consider it to be closest to current practice. We have stored data, results, and figures of all four trajectories in an online repository (insert URL here).

## Effect size precision

In both scenarios, the effect size estimates are more precise when employing T2, as can be seen by the narrower CI's in figure ??.

# Discussion

question that came up when writing up for Nat Meth
\ \ \ \ * Should we show that increased sample size at exploratory stage (power = 0.5; 0.8) does not increase sensitivity to levels of SESOI criterion? >> Ulf said not now

for intro or discussion

\ \ \ \ * Cite Howells 2014 section on power and sample sizes >> using clinically relevant SESOI to guide power calculations >> as example how to motivate high sample sizes to regulatory bodies

\ \ \ \ * On the subject of inflated effect sizes work in @ioannidis2008most and @tversky1971belief

\ \ \ \ * Include prior odds of a study and how that plays into evidence generation @ioannidis2005most

\ \ \ \ * Check sample sizes in Carneiro and Szucs studies to compare with our 10 EUs

## Percentage of experiments proceeding to replication stage

only large ES prevail in T1

\ \ \ \ When examining effect sizes that pass the filter (either related to a SESOI or conventional significance threshold) it becomes obvious that significance (with $\alpha$ = .05) it not an efficient filter for two reasons. First, it filters mostly large effects, and in both directions. Second, it misses out on many true effects which are then not further investigated. It is important to note that at this first time point in the trajectory when employing the conventional significance threshold, true positive associations are missed. This finding reinforces a systemic perspective. Only by evaluating experiments' outcomes throughout the trajectory, we are able to quantify associations we have missed. In reality, we do not know the pre-study odds of our hypothesized associations. Thus, it is all the more important that we apply decision criteria and designs that enable us to achieve high PPV even in situations when pre-study odds are low. Gievn that preclinical animal research stands at the beginning of the drug development process, we can assume that there are hundreds (if not thousands) of associations probed and only a few are true.

Pre-study odds: Where do animal studies stand? Within the drug development process, preclinical studies are in the range of a few true associations within a few hundred associations probed. So, pre-study odds are likely lower than in the "pessimistic" scenario presented here. This may vary with the specific field of research.

Consequently, in the first approach, the replication sample size was dependent on the outcome of the exploratory study, whereas using a SESOI always yielded the same sample size regardless of the exploratory effect size.

Current practice impedes efforts to replicate preclinical experiments >> because by using significance + standard we are actually prevent ourselves from detecting effects (PPV < pre-study odds)

Optimizing decision criteria and experimental designs by employing a SESOI increases chances to detect true effects >> easily applicable variations of current practice

Re-evaluation of how we generate evidence that informs decisions about bringing an intervention to the clinic
Re-evaluation: We do not want to waste animals, while it is an increase in animal numbers at first sight, in the long run, we might waste less animals because we have more robust results
Wasting animals for research that is not robust/conclusive is unethical

Regulatory authorities and animal welfare officers aim at reducing the number of animals used in experimentation. Focusing primarily on this outcome neglects aspects of quality and conclusiveness of scientific evidence generated using these animals.

Mimic phases of clinical drug development >> phase II trials >> 5 % percent significant level one-sided >> “proof-of-concept” >> effect signal >> lenient threshold at the beginning, more strict criteria as trials progress >> preclinical trials @mogil2017no

From an ethical point of view, the number animals needed in any experiment needs to be well justified. Intuitively, one would argue that replication studies contradict animal welfare considerations by nature, as it is often understood as repeating an experiment that has already been done. Regulatory authorities (in Europe) use this line of argumentation and do not or extremely rarely permit studies that set out to replicate results from a former experiment. If one adopts a systemic perspective (research trajectory), it becomes clear that depending on several decisions we make throughout the trajectory animal numbers change. In our study, animal numbers vary with the decision criterion employed at the first stage and the approach to calculate sample size for the replication.

\ \ \ \ * Daniel Lakens blogpost on feasibililty justification <http://daniellakens.blogspot.com/2020/08/feasibility-sample-size-justification.html>

## Limitations

in the limitations section you could cite the Bioarxiv paper (Bonapersona 2019) and compare the two distributions to support your choice and that it may not be overly pessimistic. Particular in fields where there was no progress like neurodegenerative diseases

Regarding the choice of 10 EUs: for the limitation section we need to argue that different initial sample sizes like 7 or 15 will not change results dramatically

Regarding Ioannidis distribution: We acknowledge that the effect sizes were mainly extracted from human studies. However, in large parts the distribution is in agreement with effect sizes reported to be typical of (some areas of) preclinical research (find good ref here \textcolor{red}{[This could be infectious diseases, usually antibiotics are all or nothing effects]}).

## Take home message

am Ende muss als take home message etwas stehen, dass Leute anwenden können. Also sowas wie: Unsere Simulationen zeigen 1., 2. 3. Deshalb schlagen wir folgendes vor, um präklinische Experimente valider (oder so in der Art) zu machen und Translation zu erleichtern: 1., 2. 3. Hier soll es ca. 3 konkrete Handlungsanweisungen geben, die leicht zu implementieren sind.

### 2021-02-05 Manuscript/Manuscript_intro_methods_results.Rmd
updated data files after R update

@@
 That is, effect magnitudes are larger than the true unknown effect [@colquhoun2014investigation; @gelman2014beyond].
-
-Along the trajectory, we have to make sure that at a given stage (which implies that we are in a specific operating mode), our strategy is aligned with the goal that we have at this stage.
-At early stages, we want to find possibly working interventions.
-In statistical terms this means that we aim for a low Type II error.
-We do not want to miss effects.
-That means that our tests should be sensitive.
-This comes at the cost of specificity.
-We allow for false positives at this stage.
-Dirnagl (2016) [@dirnagl2016thomas]
-
-*In exploratory mode, we allow for more false positives. This happens naturally, as we operate with low prior odds that our hypothesis is true.
-
-* focus in exploration should rather be on proper descriptive statistics, including measures of variance and confidence intervals.
-
-*
 
@@
 Importantly, we did not consider the *p*-value additionally.
-Applying this decision criterion resulted in `r outcomes_all_distributions_init_samp_size_10$rep_attempts[6] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[5] / 100` percent of experiments moving to replication in case of the "optimistic" distribution and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[2] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[1] / 100` percent for thes "pessimistic" distribution based on an SESOI of 0.5 and 1.0, respectively.
+Applying this decision criterion resulted in `r outcomes_all_distributions_init_samp_size_10$rep_attempts[6] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[5] / 100` percent of experiments moving to replication in case of the "optimistic" distribution and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[2] / 100` and `r outcomes_all_distributions_init_samp_size_10$rep_attempts[1] / 100` percent for the "pessimistic" distribution based on an SESOI of 0.5 and 1.0, respectively.
 Compared to the conventional significance threshold, the range of effect size estimates that proceeded to replication shifted and included less extreme values, and more values closer to zero.
@@
 Given our "optimistic" emoirical distribution, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[6], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[5], 2)` for SESOI of 0.5 and 1.0, respectively.
-In the "pessimistic" scenario, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[1], 2)`
+In the "pessimistic" scenario, the PPV is `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[2], 2)` and `r round(outcomes_all_distributions_init_samp_size_10$PPV_pop_prev[1], 2)`.
 

### 2021-02-05 Manuscript/Supplement.Rmd
updated data files after R update

---
title: 'Supplement Nature Methods'
csl: nature-methods.csl
indent: yes
output: bookdown::pdf_document2
toc: false
linestretch: 1.5
fontsize: 11pt
header-includes:
       - \usepackage{subfig}
bibliography: refs_repro.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(bookdown)
library(knitr)
library(tidyverse)
```

```{r data, include = FALSE, cache = FALSE}
load("all_outcomes_init_samp_size_10.RData")
load("exploratory_data_SESOI_Carneiro_10.RData")
load("exploratory_data_SESOI_Szucs_10.RData")
load("exploratory_data_significance_Carneiro_10.RData")
load("exploratory_data_significance_Szucs_10.RData")
```

```{r ES data, include = FALSE, cache = FALSE}
load("ES_data_Szucs.RData")
load("ES_data_Carneiro.RData")
```

```{r ES distribution, echo = FALSE, results = "hide"}
min(ES_data_Szucs$D)
max(ES_data_Szucs$D)

sum(ES_data_Szucs$D > 10)

median(ES_data_Szucs$D)
mean(ES_data_Szucs$D)

names(ES_data_Carneiro)[1] <- "D"

ES_data_Carneiro$D <- ifelse(ES_data_Carneiro$D < 0, -ES_data_Carneiro$D, -ES_data_Carneiro$D)

min(ES_data_Carneiro$D)
max(ES_data_Carneiro$D)
median(ES_data_Carneiro$D)
mean(ES_data_Carneiro$D)

ES_data_Carneiro <-
  ES_data_Carneiro %>% 
  mutate(distribution = "Pessimistic")

ES_Szucs <-
  ES_data_Szucs %>% 
  select(D) %>% 
  filter(D <= 10) %>% 
  mutate(distribution = "Optimistic")

ES_data <- bind_rows(ES_Szucs, ES_data_Carneiro)
```

```{r histograms, echo = FALSE, fig.cap = 'Empirical effect size distributions. Note that in (a) 45 values $>$ 10 were removed in order to display the distribution.', fig.subcap = c('Optimistic distribution', 'Pessimistic distribution'), out.width = '.45\\linewidth', fig.ncol = 2, fig.align = 'center'}

ES_histrogram_Szucs <-
  ggplot(data = ES_Szucs, aes(x = D)) +
  geom_histogram(binwidth = .2, color = "black", fill = "white", size = 0.3) +
  theme_bw() +
  labs(x = expression(paste("Empirical effect sizes (Cohen's ", italic("d"), ") ")), 
       y = "Frequency") + 
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black"))

ES_histrogram_Szucs

ES_histrogram_Carneiro <-
  ggplot(data = ES_data_Carneiro, aes(x = D)) +
  geom_histogram(binwidth = .2, color = "black", fill = "white", size = 0.3) +
  theme_bw() +
  labs(x = expression(paste("Empirical effect sizes (Cohen's ", italic("d"), ") ")), 
       y = "Frequency") + 
   scale_x_continuous(breaks = seq(-4, 6, 2)) +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black"))

ES_histrogram_Carneiro
```

```{r, hist, echo = FALSE, fig.cap = 'Exploratory effect sizes (left: based on optimistic empirical distribution, right: based on pessimistic empirical distribution). The panels show the effect sizes that were detected using one of the two decision criteria. Orange shaded bars (panels (a) and (b)) indicate those effect sizes that were identified for replication using the conventional significance threshold (p = .05). Blue shaded bars indicate effect sizes that were selected using a SESOI of 0.5 (panels (c) and (d)) or 1.0 (panels (e) and (f)), respectively. Note that in (a), (c), and (e) 24 values $>$ 10 were removed in order to display the distribution.', fig.align = 'center', fig.subcap = c('', ''), out.width = '.50\\linewidth', fig.ncol = 2}

exploratory_data_sig_Szucs_10 <-
  exploratory_data_sig_Szucs_10 %>%
  filter(pval_threshold == 0.05) %>%
  mutate(reverse_effect = ifelse(effect < 0, -effect, -effect)) %>% 
  filter(reverse_effect <= 10)

ES_histrogram_Szucs <-
  ggplot(data = exploratory_data_sig_Szucs_10,
         aes(x = reverse_effect,
             fill = factor(selection_sig))) +
  geom_histogram(bins = 100, color = "black", size = 0.3) +
  theme_bw() +
  labs(x = expression(paste("Exploratory effect sizes (Cohen's ", italic("d"), ") ")),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#E69F00")) +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Szucs

exploratory_data_sig_Carneiro_10 <-
  exploratory_data_sig_Carneiro_10 %>% 
  filter(pval_threshold == 0.05) %>% 
  mutate(reverse_effect = ifelse(effect < 0, -effect, -effect))

ES_histrogram_Carneiro_selected_sig <-
  ggplot(data = exploratory_data_sig_Carneiro_10, 
       aes(x = reverse_effect, fill = factor(selection_sig))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Exploratory effect sizes (Cohen's ", italic("d"), ") ")),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#E69F00")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Carneiro_selected_sig

exploratory_data_Szucs_10_0.5 <-
  exploratory_data_Szucs_10 %>% 
  filter(SESOI == 0.5) %>% 
  mutate(reverse_effect = ifelse(effect < 0, -effect, -effect)) %>% 
  filter(reverse_effect <= 10)

ES_histrogram_Szucs_selected_SESOI_0.5 <-
  ggplot(data = exploratory_data_Szucs_10_0.5, 
       aes(x = reverse_effect, fill = factor(selection_equiv))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Exploratory effect sizes (Cohen's ", italic("d"), ") ")),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Szucs_selected_SESOI_0.5

exploratory_data_Carneiro_10_0.5 <-
  exploratory_data_Carneiro_10 %>% 
  filter(SESOI == 0.5) %>% 
  mutate(reverse_effect = ifelse(effect < 0, -effect, -effect))

ES_histrogram_Carneiro_selected_SESOI_0.5 <-
  ggplot(data = exploratory_data_Carneiro_10_0.5, 
       aes(x = reverse_effect, fill = factor(selection_equiv))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Exploratory effect sizes (Cohen's ", italic("d"), ") ")),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Carneiro_selected_SESOI_0.5

exploratory_data_Szucs_10_1.0 <-
  exploratory_data_Szucs_10 %>% 
  filter(SESOI == "1.0") %>% 
  mutate(reverse_effect = ifelse(effect < 0, -effect, -effect)) %>% 
  filter(reverse_effect <= 10)

ES_histrogram_Szucs_selected_SESOI_1.0 <-
  ggplot(data = exploratory_data_Szucs_10_1.0, 
       aes(x = reverse_effect, fill = factor(selection_equiv))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Exploratory effect sizes (Cohen's ", italic("d"), ") ")),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Szucs_selected_SESOI_1.0

exploratory_data_Carneiro_10_1.0 <-
  exploratory_data_Carneiro_10 %>% 
  filter(SESOI == "1.0") %>% 
  mutate(reverse_effect = ifelse(effect < 0, -effect, -effect))

ES_histrogram_Carneiro_selected_SESOI_1.0 <-
  ggplot(data = exploratory_data_Carneiro_10_1.0, 
       aes(x = reverse_effect, fill = factor(selection_equiv))) +
  geom_histogram(bins = 90, color = "black",
                 size = 0.3, alpha = 0.8) +
  labs(#x = "",
       x = expression(paste("Exploratory effect sizes (Cohen's ", italic("d"), ") ")),
       y = "Frequency",
       fill = "Selected for \nreplication") +
  scale_fill_manual(breaks = c("0", "1"),
                    labels = c("no",
                               "yes"),
                    values = c("white", "#0072B2")) +
  theme_bw() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(axis.text.x = element_text(size = 15, colour = "black")) +
  theme(axis.text.y = element_text(size = 15, colour = "black")) +
  theme(legend.position = "none")

ES_histrogram_Carneiro_selected_SESOI_1.0

```

# Supplementary text 1: Limitations

### 2021-02-05 Manuscript/internal_revision/supplemetary_plots.Rmd
updated data files after R update

@@
 ```
+
+```{r}
+facet_names <- 
+  c("equivalence.SESOI" = "SESOI within CI \nSESOI ",
+    "significance.SESOI" = "Significance \nSESOI",
+    "equivalence.standard" = "SESOI within CI \nStandard",
+    "significance.standard" = "Significance \nStandard")
+
+plot_PPV <- 
+  ggplot(data = outcomes_10EU,
+         aes(x = SESOI, y = PPV_pop_prev)) + 
+  geom_point(size = 1.5) + 
+  facet_wrap(~ trajectory, nrow = 2, ncol = 2,  labeller = labeller(.rows = facet_names)) + 
+  labs(x = "SESOI", y = "Positive predictive value") +
+  # scale_x_discrete(labels = c("SESOI within CI \nSESOI",
+  #                             "Significance \nSESOI",
+  #                             "SESOI within CI \nStandard",
+  #                             "Significance \nStandard")) +
+  scale_x_continuous(breaks = c(0.3, 0.5, 0.7, 1.0)) +
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 11)) +
+  theme(axis.title.y = element_text(size = 11)) +
+  theme(axis.text.x = element_text(size = 7, colour = "black")) +
+  theme(axis.text.y = element_text(size = 10, colour = "black")) +
+  theme(strip.text.x = element_text(size = 11, colour = "black", face = "bold")) +
+  theme(strip.text.y = element_text(size = 11, colour = "black", face = "bold")) +
+  theme(strip.background = element_rect(fill = "white", color = "black"))
+
+# hlines <- data.frame(pre_study_odds = c(outcomes_10EU$Prevalence[1], outcomes_10EU$Prevalence[5],
+#                                         outcomes_10EU$Prevalence[9], outcomes_10EU$Prevalence[13]), 
+#                      # distribution   = c(rep(plot_data$distribution[1], 2), 
+#                      #                    rep(plot_data$distribution[5], 2)),
+#                      SESOI          = rep(c("1", "0.7", "0.5", "0.3"))) 
+# 
+# 
+# plot_PPV <- 
+#   plot_PPV + 
+#   geom_hline(data = hlines, 
+#              aes(yintercept = pre_study_odds),
+#              color = "red", lty = 2, size = .5)
+
+plot_PPV
+```
+
+```{r}
+facet_names <- 
+  c("equivalence.SESOI" = "SESOI within CI \nSESOI ",
+    "significance.SESOI" = "Significance \nSESOI",
+    "equivalence.standard" = "SESOI within CI \nStandard",
+    "significance.standard" = "Significance \nStandard")
+
+plot_FPR <- 
+  ggplot(data = outcomes_10EU,
+         aes(x = SESOI, y = FPR)) + 
+  geom_point(size = 1.5) + 
+  facet_wrap(~ trajectory, nrow = 2, ncol = 2,  labeller = labeller(.rows = facet_names)) + 
+  labs(x = "SESOI", y = "False positive rate") +
+  # scale_x_discrete(labels = c("SESOI within CI \nSESOI",
+  #                             "Significance \nSESOI",
+  #                             "SESOI within CI \nStandard",
+  #                             "Significance \nStandard")) +
+  scale_x_continuous(breaks = c(0.3, 0.5, 0.7, 1.0)) +
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 11)) +
+  theme(axis.title.y = element_text(size = 11)) +
+  theme(axis.text.x = element_text(size = 7, colour = "black")) +
+  theme(axis.text.y = element_text(size = 10, colour = "black")) +
+  theme(strip.text.x = element_text(size = 11, colour = "black", face = "bold")) +
+  theme(strip.text.y = element_text(size = 11, colour = "black", face = "bold")) +
+  theme(strip.background = element_rect(fill = "white", color = "black"))
+
+# hlines <- data.frame(pre_study_odds = c(outcomes_10EU$Prevalence[1], outcomes_10EU$Prevalence[5],
+#                                         outcomes_10EU$Prevalence[9], outcomes_10EU$Prevalence[13]), 
+#                      # distribution   = c(rep(plot_data$distribution[1], 2), 
+#                      #                    rep(plot_data$distribution[5], 2)),
+#                      SESOI          = rep(c("1", "0.7", "0.5", "0.3"))) 
+# 
+# 
+# plot_PPV <- 
+#   plot_PPV + 
+#   geom_hline(data = hlines, 
+#              aes(yintercept = pre_study_odds),
+#              color = "red", lty = 2, size = .5)
+
+plot_FPR
+```

### 2021-02-05 scripts/analysis/Carneiro_PPV_only_exploratory_stage.R

@@ -1,111 +1,0 @@
-
-
-outcomes_Carneiro_exploratory <- read.csv(file = "./data/Carneiro_distribution_only_exploratory_stage")
-
-outcomes_10 <-
-  outcomes_Carneiro_exploratory %>% 
-  filter(init_sample_size == 10)
-
-outcomes_10$H0 <- ifelse(outcomes_10$p_value <= .05, 2, 1)
-
-sum(outcomes_10$H0 == 2)
-
-outcomes_10$prev_pop      <- mat[1, 2]
-outcomes_10$all_positives <- mat[2, 2]
-outcomes_10$all_negatives <- mat[3, 2]
-
-mat
-
-
-dat <-
-  outcomes_10 %>% 
-  select(init_sample_size, ES_true, effect, H0,
-         prev_pop, all_positives, all_negatives)
-
-dat$effect <- ifelse(dat$effect < 0, -dat$effect, -dat$effect)
-
-data_false_positives <-
-  dat %>%
-  filter(H0 == 2 & ES_true < .5)
-
-data_false_negatives <-
-  dat %>%
-  filter(H0 == 1 & ES_true > .5)
-
-hist(data_false_positives$effect)
-
-hist(data_false_positives$ES_true)
-
-
-true_positives <-
-  dat %>%
-  # group_by(init_sample_size, ES_true, effect, H0,
-  #          prev_pop, all_positives, all_negatives) %>%
-  filter(H0 == 2 & ES_true > .5) %>%
-  summarize(true_pos = n())
-
-false_negatives <-
-  dat %>%
-  # group_by(init_sample_size, ES_true, effect, H0,
-  #          prev_pop, all_positives, all_negatives) %>%
-  filter(H0 == 1 & ES_true > .5) %>%
-  summarize(false_neg = n())
-
-true_negatives <-
-  dat %>%
-  # group_by(init_sample_size, ES_true, effect, H0,
-  #          prev_pop, all_positives, all_negatives) %>%
-  filter(H0 == 1 & ES_true < .5) %>%
-  summarize(true_neg = n())
-
-false_positives <-
-  dat %>%
-  # group_by(init_sample_size, ES_true, effect, H0,
-  #          prev_pop, all_positives, all_negatives) %>%
-  filter(H0 == 2 & ES_true < .5) %>%
-  summarize(false_pos = n())
-
-true_positives$false_neg <- false_negatives$false_neg
-true_positives$true_neg  <- true_negatives$true_neg
[…]

### 2021-02-05 scripts/analysis/Szucs_PPV_only_exploratory_stage.R

@@ -1,111 +1,0 @@
-
-
-outcomes_Szucs_exploratory <- read.csv(file = "./data/Szucs_distribution_only_exploratory_stage")
-
-outcomes_10 <-
-  outcomes_Szucs_exploratory %>% 
-  filter(init_sample_size == 10)
-
-sum(outcomes_10$H0 == 2)
-
-
-outcomes_10$H0 <- ifelse(outcomes_10$p_value <= .05, 2, 1)
-
-outcomes_10$prev_pop      <- mat[1, 2]
-outcomes_10$all_positives <- mat[2, 2]
-outcomes_10$all_negatives <- mat[3, 2]
-
-mat
-
-
-dat <-
-  outcomes_10 %>% 
-  select(init_sample_size, ES_true, effect, H0,
-         prev_pop, all_positives, all_negatives)
-
-dat$effect <- ifelse(dat$effect < 0, -dat$effect, -dat$effect)
-
-data_false_positives <-
-  dat %>%
-  filter(H0 == 2 & ES_true < .5)
-
-data_false_negatives <-
-  dat %>%
-  filter(H0 == 1 & ES_true > .5)
-
-hist(data_false_positives$effect)
-
-hist(data_false_positives$ES_true)
-
-
-true_positives <-
-  dat %>%
-  # group_by(init_sample_size, ES_true, effect, H0,
-  #          prev_pop, all_positives, all_negatives) %>%
-  filter(H0 == 2 & ES_true > .5) %>%
-  summarize(true_pos = n())
-
-false_negatives <-
-  dat %>%
-  # group_by(init_sample_size, ES_true, effect, H0,
-  #          prev_pop, all_positives, all_negatives) %>%
-  filter(H0 == 1 & ES_true > .5) %>%
-  summarize(false_neg = n())
-
-true_negatives <-
-  dat %>%
-  # group_by(init_sample_size, ES_true, effect, H0,
-  #          prev_pop, all_positives, all_negatives) %>%
-  filter(H0 == 1 & ES_true < .5) %>%
-  summarize(true_neg = n())
-
-false_positives <-
-  dat %>%
-  # group_by(init_sample_size, ES_true, effect, H0,
-  #          prev_pop, all_positives, all_negatives) %>%
-  filter(H0 == 2 & ES_true < .5) %>%
-  summarize(false_pos = n())
-
-true_positives$false_neg <- false_negatives$false_neg
-true_positives$true_neg  <- true_negatives$true_neg
[…]

### 2021-02-05 scripts/analysis/analysis_results_across_trajectory_Carneiro.R

@@ -1,0 +1,20 @@
+library(tidyverse)
+
+setwd("~/Documents/SimulateTranslation/")
+
+
+extractor <- function(k) {
+  
+  test_data <- read_csv(paste("./data/Carneiro_distribution/Frequentist_analysis/exploratory_stage_equiv_", k, sep = ""))
+}
+
+sets_equiv <- c("0.1", "0.3", "0.5", "0.7", "1.0")
+
+exploratory_data <- list()
+
+for (i in sets_equiv) {
+  
+  exploratory_data[[i]] <- extractor(i)
+  print(i)
+  
+}

### 2021-02-05 scripts/analysis/analysis_simulation_results_across_trajectory.R

@@ -1,6 +1,7 @@
 setwd("~/Documents/SimulateTranslation/")
 
 rm(list = ls())
+
 library(tidyverse)
 
 source("./scripts/analysis/prior_odds_for_analysis_Szucs.R")
@@ -10,9 +11,10 @@ source("./scripts/analysis/prior_odds_for_analysis_Szucs.R")
 ### read in data sets from the different trajectories 
 ### using fixed-N design in replication study
 
-final <- read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method2_0.1")
+# final <- 
+#   read.csv(file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_distribution_equiv_method1_0.1")
 
-# final <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1")
+final <- read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_equiv_method1_1.0")
 
 
 hist(final$effect, breaks = 100)
@@ -21,10 +23,10 @@ hist(final$ES_true, breaks = 100)
 
 
 ### add column that codes decision criterion from exploratory stage to confirmatory stage
-final$decision_crit <- "significance"
+final$decision_crit <- "equivalence"
 
 ### add column that codes approach to determine sample size for confirmatory study
-final$sampsize_approach <- "standard"
+final$sampsize_approach <- "SESOI"
 
 ### add column for outcome significant / not significant to match outcome column
 ### of sequential design data set
@@ -32,9 +34,9 @@ final$H0 <- ifelse(final$p_value <= .05, 2, 1)
 
 ### add column for prevalence 
 
-final$prev_pop      <- mat[1, 2]
-final$all_positives <- mat[2, 2]
-final$all_negatives <- mat[3, 2]
+final$prev_pop      <- mat[1, 1]
+final$all_positives <- mat[2, 1]
+final$all_negatives <- mat[3, 1]
 
 mat
 
@@ -44,7 +46,7 @@ dat <-
   final %>% 
   select(init_sample_size, rep_no, decision_crit, sampsize_approach, 
          rep_sample_size, ES_true, effect, H0,
-         prev_pop, all_positives, all_negatives, rep_attempts)
+         prev_pop, all_positives, all_negatives)
 
 # write.csv(dat, file = "./data/Carneiro_distribution/Frequentist_analysis/Carneiro_replication_sig_method1")
 

### 2021-02-05 scripts/analysis/analysis_simulation_results_second_stage.R

@@ -1,10 +1,14 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation/data")
+setwd("~/Documents/SimulateTranslation")
 
-#rm(list = ls())
+rm(list = ls())
+
 library(tidyverse)
 
+source("./scripts/analysis/prior_odds_for_analysis_Szucs.R")
 
-final <- read.csv(file = "Szucs_distribution_equiv_method2_1.0")
+
+final <- 
+  read.csv(file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1_p0.05")
 
 
 hist(final$effect, breaks = 100)
@@ -13,13 +17,11 @@ hist(final$ES_true, breaks = 100)
 
 
 ### add column that codes decision criterion from exploratory stage to confirmatory stage
-final$decision_crit <- "equivalence"
+final$decision_crit <- "significance"
 
 ### add column that codes approach to determine sample size for confirmatory study
-final$sampsize_approach <- "SESOI"
+final$sampsize_approach <- "standard"
 
-### add column that codes design applied in confirmatory study
-#finalfix$design <- rep("fixN")
 
 ### add column for outcome significant / not significant to match outcome column
 ### of sequential design data set
@@ -27,9 +29,9 @@ final$H0 <- ifelse(final$p_value <= .05, 2, 1)
 
 ### add column for prevalence 
 
-final$prev_pop      <- mat[1, 4]
-final$all_positives <- mat[2, 4]
-final$all_negatives <- mat[3, 4]
+final$prev_pop      <- mat[1, 3]
+final$all_positives <- mat[2, 3]
+final$all_negatives <- mat[3, 3]
 
 mat
 
@@ -42,25 +44,25 @@ dat <-
 true_positives <-
   dat %>%
   group_by(init_sample_size, decision_crit, sampsize_approach, H0, rep_attempts, prev_pop) %>%
-  filter(H0 == 2 & ES_true > 1) %>%
+  filter(H0 == 2 & ES_true > .5) %>%
   summarize(true_pos = n())
 
 false_positives <-
   dat %>%
   group_by(init_sample_size, decision_crit, sampsize_approach, H0, rep_attempts, prev_pop) %>%
-  filter(H0 == 2 & ES_true < 1) %>%
+  filter(H0 == 2 & ES_true < .5) %>%
   summarize(false_pos = n())
 
[…]

### 2021-02-05 scripts/analysis/prior_odds_for_analysis_Szucs.R

@@ -1,4 +1,4 @@
-setwd("~/Documents/QUEST/PhD/R/SimulateTranslation")
+setwd("~/Documents/SimulateTranslation")
 
 
 # source additional functions
@@ -14,11 +14,11 @@ hist(ES_true, breaks = 200)
 hist(current_ES, breaks = 200)
 
 #how many hypothesis over SESOI threshold
-SESOI       <- c(.5, 1)
+SESOI <- c(.1, .3, .5, .7, 1)
 
 mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
               dimnames = list(c("prev_pop", "all_positives", "all_negatives"), 
-                              c(.5, 1)))
+                              c(.1, .3, .5, .7, 1)))
 
 prev_pop      <- vector()
 all_positives <- vector()

### 2021-02-05 scripts/simulation/sim_fixN_Carneiro_decision_equivalence_with_SESOI.R

@@ -116,7 +116,7 @@ list_exploratory_data <-
 #the confidence interval generated here is used in the equivalence test
 exploratory_data_summary <- list()
 
-plan(multiprocess)
+plan(multisession)
 for (i in 1:length(samp_size_vector)) {
   
   exploratory_data_summary[[i]] <- 
@@ -134,7 +134,7 @@ selection_equiv <- list()
 for (i in 1:length(samp_size_vector)) {
   
   selection_equiv[[i]] <- future_map(exploratory_data_summary[[i]], get_decision_equiv,
-                                     SESOI = 0.7)
+                                     SESOI = 1.0)
   
 }
 

### 2021-02-05 scripts/simulation/sim_fixN_Carneiro_replication.R

@@ -1,6 +1,6 @@
 setwd("~/Documents/SimulateTranslation")
 
-rm(list = ls())
+# rm(list = ls())
 
 
 # source additional scripts
@@ -16,13 +16,13 @@ sum(dat$effect > 0)
 sum(dat$ES_true < 0)
 
 # data <-
-#   dat %>% 
-#   group_by(init_sample_size, study_id) %>% 
+#   dat %>%
+#   group_by(init_sample_size, study_id) %>%
 #   filter(selection_equiv == 1)
 
 data <-
-  dat %>% 
-  group_by(init_sample_size, study_id) %>% 
+  dat %>%
+  group_by(init_sample_size, study_id) %>%
   filter(selection_sig == 1)
 
 sum(data$effect < 0)
@@ -30,13 +30,13 @@ sum(data$effect > 0)
 sum(data$effect == 0)
 
 # selected <-
-#   data %>% 
-#   group_by(init_sample_size) %>% 
+#   data %>%
+#   group_by(init_sample_size) %>%
 #   summarize(selected = sum(selection_equiv == 1))
 
 selected <-
-  data %>% 
-  group_by(init_sample_size) %>% 
+  data %>%
+  group_by(init_sample_size) %>%
   summarize(selected = sum(selection_sig == 1))
 
 # now estimate sample size for replication study
@@ -47,13 +47,13 @@ rep_sample_size_std <- NULL
 
 for (i in 1:nrow(data)) {
 
-  rep_sample_size_std[i] <-
-    ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-                             method = 1))
-  
   # rep_sample_size_std[i] <-
   #   ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-  #                            method = 2, SESOI = 0.5, power = .5))
+  #                            method = 1))
+  
+  rep_sample_size_std[i] <-
+    ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
+                             method = 2, SESOI = 0.1, power = .5))
 }
 
 data$rep_samp_size_std <- rep_sample_size_std
@@ -115,43 +115,37 @@ for(i in select_experiments) {
     mutate(study_id = rep_exp_no)
 }
 
-# replication_data[[11138]]
 
-plan(multiprocess)
+plan(multisession)
 rep_data_summary <- 
[…]

### 2021-02-05 scripts/simulation/sim_fixN_Szucs_decision_significance.R

@@ -97,7 +97,7 @@ list_exploratory_data <-
 #the confidence interval generated here is used in the equivalence test
 exploratory_data_summary <- list()
 
-plan(multiprocess)
+plan(multisession)
 for (i in 1:length(samp_size_vector)) {
   
   exploratory_data_summary[[i]] <- 

### 2021-02-05 scripts/simulation/sim_fixN_Szucs_replication.R

@@ -29,7 +29,6 @@ sum(data$effect < 0)
 sum(data$effect > 0)
 sum(data$effect == 0)
 
-
 # selected <-
 #   data %>%
 #   group_by(init_sample_size) %>%
@@ -49,13 +48,13 @@ rep_sample_size_std <- NULL
 
 for (i in 1:nrow(data)) {
   
-  # rep_sample_size_std[i] <-
-  #   ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-  #                            method = 1))
-  
   rep_sample_size_std[i] <-
     ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
-                             method = 2, SESOI = 1.0, power = .5))
+                             method = 1))
+  
+  # rep_sample_size_std[i] <-
+  #   ceiling(calc_sample_size(data = data[i, ], sample_size = data[i, ]$init_sample_size,
+  #                            method = 2, SESOI = 0.1, power = .5))
 }
 
 data$rep_samp_size_std <- rep_sample_size_std
@@ -119,20 +118,14 @@ for(i in select_experiments) {
     mutate(study_id = rep_exp_no)
 }
 
-#replication_data[[111]]
-
-plan(multiprocess)
+plan(multisession)
 rep_data_summary <- 
   future_map(replication_data, get_summary_study_rep)
-
-# rep_data_summary[[1]]
-# data$ES_true[select_experiments]
 
 res_summary_rep <-
   data.frame(init_sample_size = data$init_sample_size[select_experiments],
              rep_no = c(1:rep_exp_no),
              rep_sample_size = rep_sample_size_std[select_experiments],
-             #rep_sample_size = rep_sample_size_std,
              t_value = unlist(map(rep_data_summary, "t_value")),
              p_value = unlist(map(rep_data_summary, "p_value")), #[seq(1, 2*rep_exp_no, 2)],
              effect = unlist(map(rep_data_summary, "effect")),
@@ -144,9 +137,11 @@ hist(res_summary_rep$effect, breaks = 200)
 res_summary_rep$effect <- ifelse(res_summary_rep$effect < 0,
                                  -res_summary_rep$effect, -res_summary_rep$effect)
 
-# write.csv(res_summary_rep, file = "./data/Szucs_distribution/Frequentist_analysis/Szucs_distribution_sig_method1_p0.05")
[…]

### 2021-02-05 additional_plots_FPR.R
updated data files after R update

@@ -1,0 +1,76 @@
+
+facet_names <- 
+  c("equivalence.SESOI" = "SESOI within CI \nSESOI ",
+    "significance.SESOI" = "Significance \nSESOI",
+    "equivalence.standard" = "SESOI within CI \nStandard",
+    "significance.standard" = "Significance \nStandard")
+ 
+plot_FPR <- 
+  ggplot(data = outcomes_10EU,
+         aes(x = SESOI, y = FPR,
+             shape = trajectory,
+             color = trajectory)) + 
+  geom_point(size = 3) + 
+  geom_line() +
+  # facet_wrap(~ trajectory, nrow = 2, ncol = 2,  labeller = labeller(.rows = facet_names)) + 
+  labs(x = "SESOI", y = "False positive rate") +
+  scale_x_continuous(breaks = c(0.3, 0.5, 0.7, 1.0)) +
+  scale_color_manual(values = c("#E69F00", "#CC79A7",  "#0072B2", "#009E73")) +
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 11)) +
+  theme(axis.title.y = element_text(size = 11)) +
+  theme(axis.text.x = element_text(size = 7, colour = "black")) +
+  theme(axis.text.y = element_text(size = 10, colour = "black")) +
+  theme(strip.text.x = element_text(size = 11, colour = "black", face = "bold")) +
+  theme(strip.text.y = element_text(size = 11, colour = "black", face = "bold")) +
+  theme(strip.background = element_rect(fill = "white", color = "black"))
+
+plot_FPR
+
+
+colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
+                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+pie(rep(1, 8), col = colorBlindBlack8)
+
+
+plot_PPV <- 
+  ggplot(data = outcomes_10EU,
+         aes(x = SESOI, y = PPV_pop_prev,
+             shape = trajectory,
+             color = trajectory)) + 
+  geom_point(size = 3) + 
+  # facet_wrap(~ trajectory, nrow = 2, ncol = 2,  labeller = labeller(.rows = facet_names)) + 
+  labs(x = "SESOI", y = "Positive predictive value") +
+  scale_x_continuous(breaks = c(0.3, 0.5, 0.7, 1.0)) +
+  scale_color_manual(values = c("#E69F00", "#CC79A7",  "#0072B2", "#009E73")) +
+  theme_bw() +
[…]

### 2021-02-08 data/,.R
changed instructions in simulation scripts for Anja

@@ -1,0 +1,90 @@
+setwd("~/Documents/SimulateTranslation/data")
+
+load("exploratory_data_SESOI_Carneiro_10.RData")
+
+exploratory_data_Carneiro_10_0.5 <-
+  exploratory_data_Carneiro_10 %>% 
+  filter(SESOI == 0.5)
+
+ES_histrogram_Carneiro_selected_SESOI_0.5 <-
+  ggplot(data = exploratory_data_Carneiro_10_0.5, 
+         aes(x = ES_true, fill = factor(selection_equiv))) +
+  geom_histogram(bins = 90, color = "black",
+                 size = 0.3, alpha = 0.8) +
+  labs(#x = "",
+    x = expression(paste("Cohen's ", italic("d"))),
+    y = "Frequency",
+    fill = "Selected for \nreplication") +
+  scale_fill_manual(breaks = c("0", "1"),
+                    labels = c("no",
+                               "yes"),
+                    values = c("white", "#0072B2")) +
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 16)) +
+  theme(axis.title.y = element_text(size = 16)) +
+  theme(axis.text.x = element_text(size = 15, colour = "black")) +
+  theme(axis.text.y = element_text(size = 15, colour = "black")) +
+  theme(legend.position = "none")
+
+ES_histrogram_Carneiro_selected_SESOI_0.5
+
+exploratory_data_Carneiro_10_1.0 <-
+  exploratory_data_Carneiro_10 %>% 
+  filter(SESOI == "1.0")
+
+ES_histrogram_Carneiro_selected_SESOI_1.0 <-
+  ggplot(data = exploratory_data_Carneiro_10_1.0, 
+         aes(x = ES_true, fill = factor(selection_equiv))) +
+  geom_histogram(bins = 90, color = "black",
+                 size = 0.3, alpha = 0.8) +
+  labs(#x = "",
+    x = expression(paste("Cohen's ", italic("d"))),
+    y = "Frequency",
+    fill = "Selected for \nreplication") +
+  scale_fill_manual(breaks = c("0", "1"),
+                    labels = c("no",
+                               "yes"),
+                    values = c("white", "#0072B2")) +
+  theme_bw() +
+  theme(axis.title.x = element_text(size = 16)) +
+  theme(axis.title.y = element_text(size = 16)) +
+  theme(axis.text.x = element_text(size = 15, colour = "black")) +
[…]

### 2021-02-08 scripts/simulation/functions_for_sim.R

@@ -1,4 +1,5 @@
 
+# generates study data for exploratory phase
 generate_study <- function(ES_true = 1, sample_size = samp_size, l_bias = 0, pop_sd = 1) {
   ES_mod <- ES_true + l_bias
   sample_data <- data.frame(values = c(rnorm(sample_size, 0, pop_sd),
@@ -11,7 +12,7 @@ generate_study <- function(ES_true = 1, sample_size = samp_size, l_bias = 0, pop
   return(sample_data)
 }
 
-
+# performs t-test and generates summary of study data for exploratory phase
 get_summary_study <- function(study_data) {
   
   t <- t.test(study_data$values ~ study_data$treatment,
@@ -46,7 +47,7 @@ get_summary_study <- function(study_data) {
   
 }
 
-
+# performs t-test and generates summary of study data for replication
 get_summary_study_rep <- function(study_data) {
 
   t <- t.test(study_data$values ~ study_data$treatment,
@@ -76,7 +77,7 @@ get_summary_study_rep <- function(study_data) {
 
 }
 
-
+# performs t-test and generates summary of study data for replication with Bayes factors
 get_summary_study_rep_BF <- function(study_data) {
 
   BF <- ttestBF(x = study_data$values[study_data$treatment == "treat"],
@@ -104,6 +105,7 @@ get_summary_study_rep_BF <- function(study_data) {
   
 }
 
+# decision criterion significance at exploratory stage
 get_decision_sig <- function(study_summary, pval_threshold) {
   
   bb <- study_summary$p_value
@@ -111,7 +113,7 @@ get_decision_sig <- function(study_summary, pval_threshold) {
   
 }
 
-
+# decision criterion SESOI at exploratory stage
 get_decision_equiv <- function(exploratory_data_summary,
                                SESOI) {
   
@@ -125,35 +127,10 @@ get_decision_equiv <- function(exploratory_data_summary,
   
 }
 
-
-# calc_sample_size <- function(study_summary, study_data, max_sample_size = 200,
-#                              alpha = .05, power = .8, SESOI, method = 2) {
-#   aa <- study_summary
-#   
-#   #if (method == 1) es_measured <- abs(aa$effect) #take initial study main effect
-#   if (method == 1) {
-#     es_measured <- abs(aa$effect) #take initial study main effect
-#     es_measured <- ifelse(es_measured > 16, 16, es_measured)
-#   }  
-#   
[…]

### 2021-02-08 scripts/simulation/sim_fixN_Szucs_decision_equivalence_with_SESOI.R

@@ -5,19 +5,13 @@ rm(list = ls())
 # Here we simulate a preclinical research trajectory
 # in the first stage an exploratory study is conducted with a limited number of animals
 # in the second stage this experiment is replicated based on some information from the first experiment
-# in the third stage a multi center study is conducted
-# outcome for each experiment is 
-# 1. the measured effect size after each run (with associated CI) and p-value
-# 2. the number of animals 
-# 3. the true effect size
 
-
-# source additional functions
+# source additional scripts
 source("./scripts/simulation/load_packages.R")
 source("./scripts/simulation/load_data_Szucs.R")
 source("./scripts/simulation/functions_for_sim_BF.R")
 
-
+# load additional libraries for parallel processing
 library(foreach)
 library(doMC)
 registerDoMC(cores = 4)
@@ -25,15 +19,20 @@ library(doParallel)
 registerDoParallel()
 getDoParWorkers()
 
-n_exp <- 10000
-ES_true <- ES_data_Szucs$D
+n_exp <- 10000 # number of experiments we run in 1st stage (exploration)
+ES_true <- ES_data_Szucs$D # empirical effect sizes (ES)
 
+# set seed to reproduce results
 set.seed(4321)
+
+# sample from ES distribution and show histograms of empirical and sampled ES
 current_ES <- sample(ES_true, n_exp)
 hist(ES_true, breaks = 200)
 hist(current_ES, breaks = 200)
 
-#how many hypothesis over SESOI threshold
+# how many hypothesis over SESOI threshold
+# make a matrix of prevalence, positives, and negatives for each SESOI
+# important for calculation of outcomes (PPV, FPR, FNR) later
 SESOI <- c(.1, .3, .5, .7, 1)
 
 mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
@@ -69,8 +68,12 @@ mat[3, ] <- all_negatives
 
 mat
 
+# here starts the actual simulation
+# we test three initial sample sizes (robustness check)
 samp_size_vector <- c(7, 10, 15)
 
+# we create a list of exploratory data
+# function generate_study() taken from script functions_for_sim.R
 list_exploratory_data <- 
   
   foreach(samp_size = samp_size_vector) %do% {
@@ -92,10 +95,12 @@ list_exploratory_data <-
   }
 
 
-#the confidence interval generated here is used in the equivalence test
[…]

### 2021-02-08 scripts/simulation/sim_fixN_Szucs_decision_significance.R

@@ -5,19 +5,13 @@ rm(list = ls())
 # Here we simulate a preclinical research trajectory
 # in the first stage an exploratory study is conducted with a limited number of animals
 # in the second stage this experiment is replicated based on some information from the first experiment
-# in the third stage a multi center study is conducted
-# outcome for each experiment is 
-# 1. the measured effect size after each run (with associated CI) and p-value
-# 2. the number of animals 
-# 3. the true effect size
 
-
-# source additional functions
+# source additional scripts
 source("./scripts/simulation/load_packages.R")
 source("./scripts/simulation/load_data_Szucs.R")
 source("./scripts/simulation/functions_for_sim_BF.R")
 
-
+# load additional libraries for parallel processing
 library(foreach)
 library(doMC)
 registerDoMC(cores = 4)
@@ -25,17 +19,21 @@ library(doParallel)
 registerDoParallel()
 getDoParWorkers()
 
-n_exp <- 10000
-ES_true <- ES_data_Szucs$D
+n_exp <- 10000 # number of experiments we run in 1st stage (exploration)
+ES_true <- ES_data_Szucs$D # empirical effect sizes (ES)
 
+# set seed to reproduce results
 set.seed(4321)
+
+# sample from ES distribution and show histograms of empirical and sampled ES
 current_ES <- sample(ES_true, n_exp)
 hist(ES_true, breaks = 200)
 hist(current_ES, breaks = 200)
 
-max(current_ES)
 
-#how many hypothesis over SESOI threshold
+# how many hypothesis over SESOI threshold
+# make a matrix of prevalence, positives, and negatives for each SESOI
+# important for calculation of outcomes (PPV, FPR, FNR) later
 SESOI <- c(.1, .3, .5, .7, 1)
 
 mat <- matrix(NA, nrow = 3, ncol = length(SESOI),
@@ -71,8 +69,12 @@ mat[3, ] <- all_negatives
 
 mat
 
+# here starts the actual simulation
+# we test three initial sample sizes (robustness check)
 samp_size_vector <- c(7, 10, 15)
 
+# we create a list of exploratory data
+# function generate_study() taken from script functions_for_sim.R
 list_exploratory_data <- 
   
   foreach(samp_size = samp_size_vector) %do% {
@@ -94,7 +96,9 @@ list_exploratory_data <-
   }
 
 
-#the confidence interval generated here is used in the equivalence test
[…]

### 2021-02-08 scripts/simulation/sim_fixN_Szucs_replication.R

@@ -1,13 +1,12 @@
 setwd("~/Documents/SimulateTranslation")
 
-rm(list = ls())
+# source script that runs the first part of the simulation and
+# provides the data frame with outcomes from the exploratory stage
+# either with significance as decision criterion
+# or SESOI as decision criterion
 
-
-# source additional scripts
-
-source("./scripts/simulation/sim_fixN_Szucs_decision_equivalence_with_SESOI.R")
-
-# source("./scripts/simulation/sim_fixN_Szucs_decision_significance.R")
+# source("./scripts/simulation/sim_fixN_Szucs_decision_equivalence_with_SESOI.R")
+source("./scripts/simulation/sim_fixN_Szucs_decision_significance.R")
 
 dat
 
@@ -15,34 +14,37 @@ sum(dat$effect < 0) # empirical effect sizes are negative because t.test functio
 sum(dat$effect > 0) 
 sum(dat$ES_true < 0)
 
-# data <-
-#   dat %>%
-#   group_by(init_sample_size, study_id) %>%
-#   filter(selection_sig == 1)
-
+# select only those experiments that were selected via the respective decision criterion
 data <-
   dat %>%
   group_by(init_sample_size, study_id) %>%
-  filter(selection_equiv == 1)
+  filter(selection_sig == 1)
+
+# data <-
+#   dat %>%
+#   group_by(init_sample_size, study_id) %>%
+#   filter(selection_equiv == 1)
 
 sum(data$effect < 0)
 sum(data$effect > 0)
 sum(data$effect == 0)
 
-# selected <-
-#   data %>%
-#   group_by(init_sample_size) %>%
-#   summarize(selected = sum(selection_sig == 1))
-
+# check how many were selected depending on initial sample size
 selected <-
   data %>%
   group_by(init_sample_size) %>%
-  summarize(selected = sum(selection_equiv == 1))
+  summarize(selected = sum(selection_sig == 1))
+
+# selected <-
+#   data %>%
+#   group_by(init_sample_size) %>%
+#   summarize(selected = sum(selection_equiv == 1))
 
 
 # now estimate sample size for replication study
-# A. Initial effect size (probably the worst solution but good as a floor benchmark, 1 in function)
-# B. Set a SESOI. With this all experiments will have the same number of EU (2 as function parameter)
[…]
